### 2025-01-05 paper/paper.tex
Add paper latex

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
% N.B.: do not change anything above this line. If you require additional packages, please load them directly after this line.
\usepackage{mwe}
% N.B.: you may delete the preceding line. It is used to display an example image in this template.

\begin{document}

\title{Should We Compute Feature Effects on Training or Validation Data?}

% \titlerunning{Should We Compute Feature Effects on Training or Validation Data?}
% If the full title of your paper is short enough to also fit in the running head, you can omit the abbreviated paper title here. You can check as follows: if you comment out the \titlerunning line, something will appear in the header of all odd-numbered pages of your PDF from page 3 onward. This something is either the full title (in which case all is well), or the error message "Title Suppressed Due to Excessive Length". If this error message appears, you're going to want to provide an abbreviated title within the \titlerunning command, because if you won't do it, Springer will do it for you.

%N.B.: Author information (both in the \author{} and \authorrunning{} command) should only be present in the Camera-Ready Version of your paper. The version that you initially submit for review, ought to be double-blind. So, when initially submitting your paper, use:
%\author{Author information scrubbed for double-blind reviewing}
\author{Timo Heiß\inst{1}}
% You may leave out the orcidID information, if you want to.
% Use \corr to indicate the corresponding author. Note the spacing around the \corr command. Only one author can be the corresponding author.

%N.B.: comment out the \authorrunning{} command for the double-blind version of your paper submitted for review. Later, if your paper is accepted, use the command for the Camera-Ready Version.
\authorrunning{Timo Heiß}
% First names are abbreviated in the running head.
% If there is one author, write 'A.L. Benjamin'.
% If there are two authors, write 'A.L. Benjamin and C.C. Broadus Jr.'
% If there are more than two authors, '[...] et al.' is used.

\institute{Department of Statistics, LMU Munich, Ludwigstr. 33, 80539 Munich, Germany \email{t.heiss@campus.lmu.de}}

\maketitle              % typeset the header of the contribution

\begin{abstract}
    The abstract should briefly summarize the contents of the paper in
    150--250 words.

    \keywords{Explainable AI  \and Feature effects \and Partial dependence plot \and Accumulated local effects}
\end{abstract}

\section{Introduction}
Most Machine Learning (ML) models can be considered black boxes --- opaque
systems that intrinsically do not allow insight into their internal reasoning,
making it often impossible to explain their decisions. However, this can be
problematic in many domains and applications, such as the healthcare, legal, or
finance sectors, where decisions must be transparent and
accountable~\cite{adadi_peeking_2018}.

Interpretability is crucial to enhance
trust~\cite{ribeiro_why_2016,teach_analysis_1981}, address potential
biases~\cite{guidotti_survey_2019}, fairness and ethical concerns
~\cite{lipton_mythos_2018}, and ensure compliance with regulations such as the
EU's General Data Protection Regulation (GDPR)~\cite{gdpr2016} and AI
Act~\cite{euaia2024}. To address these challenges, the field of Explainable AI
/ Interpretable ML has emerged~\cite{adadi_peeking_2018}. Although there are
many different methods\footnote{For an overview of Explainable AI methods, see
    e.g.~\cite{adadi_peeking_2018,kamath_introduction_2021,molnar_interpretable_2022}.},
we will focus on feature effect methods like \textit{Partial Dependence Plots
    (PDP)}~\cite{friedman_greedy_2001} and \textit{Accumulated Local Effects
    (ALE)}~\cite{apley_visualizing_2020}.

Due to the severity of many applications, it is crucial to utilize these
explainability methods correctly. In general, there are many pitfalls to be
aware of~\cite{molnar_general_2022}, including whether to compute explanations
in-sample, i.e.\ on training data, or out-of-sample, which we refer to as
validation data in the following. In loss-based methods such as
\textit{Permutation Feature Importance (PFI)
}\cite{breiman_random_2001,fisher_all_2019}, this choice is crucial and has
already been studied (e.g., in~\cite{molnar_general_2022}). Nevertheless, other
explainability methods, such as the \textit{Mean Decrease in Impurity (MDI)}
(or \textit{Gini Importance}) of Random Forests, or \textit{SHAP}
values~\cite{loecher_debiasing_2022,loecher_debiasing_2024}, have also been
found to exhibit biases when computed on training data.

However, to the best of our knowledge, there exist no similar studies for
feature effect methods like PDP and ALE. Existing works, including the original
papers of PDP and ALE, rely on training data without further justification
(e.g.,~\cite{apley_visualizing_2020,friedman_greedy_2001,molnar_interpretable_2022}).
In contrast, practitioners often advocate for using unseen test/validation
data\footnote{see, e.g.\ \url{https://github.com/SauceCat/PDPbox/issues/68}
    (10/27/2024)}, or base their choice on practical constraints like dataset
size\footnote{see, e.g.
    \url{https://forums.fast.ai/t/partial-dependence-plot/98465} (10/27/2024)}.
While the training set is usually larger and might thus lead to less variance
in the feature effect estimates, a too large dataset can increase computation
times substantially, particularly for the PDP~\cite{friedman_greedy_2001}. On
the other hand, although feature effects are not based on generalization error
like PFI, it is not clear how much they are affected by overfitting or
distribution shifts between training and validation data.

In this paper, we aim to answer this largely unaddressed, fundamental
methodological question of whether to use training or validation data to
compute feature effects. We perform an empirical simulation study, comparing
feature effect error and uncertainty for PDP and ALE across training data,
validation data, and cross-validation scenarios, considering various data
scenarios and model types. Our main contributions in this paper are as follows.
\begin{enumerate}
    \item We shed light on the question of whether to compute feature effects on training
          data, validation data, or in a cross-validated manner, grounded through our
          comprehensive simulation study, considering feature effect error, bias, and
          variance across different models and data scenarios.
    \item We define methods to quantify feature effect error and uncertainty in a
          theoretical framework, building upon previous work in this area.
    \item We provide an overview of commonly used test functions for simulation studies
          in Interpretable ML, including applications and purposes.
\end{enumerate}

These contributions have several important implications: Our empirically
grounded recommendations enable practitioners and researchers to make informed
decisions about which dataset to select for feature effect computation, helping
to understand potential implications of their choices. Additionally, our
framework for evaluating feature effects and our systematic collection of test
functions provide a foundation for future research in Interpretable ML, with
the latter specifically facilitating test function choice for simulation
studies.

The remainder of this paper is structured as follows. In
\textsc{Section~\ref{sec:related-works}}, we introduce the considered feature
effect methods PDP and ALE, as well as related works on feature effect error
and uncertainty, and give an overview of common test functions for simulation
studies. In \textsc{Section~\ref{sec:quantify-fe-u}}, we extend existing works
and give our definitions of feature effect errors and uncertainty. We then
describe the methodology and set-up of our simulation studies in
\textsc{Section~\ref{sec:methodology-set-up}}, present the results in
\textsc{Section~\ref{sec:results}}, and discuss their implications and
limitations in \textsc{Section~\ref{sec:discussion}}. In
\textsc{Section~\ref{sec:conclusion}}, we briefly conclude our work.

\section{Background \& Related Work}\label{sec:related-works}

\subsection{Feature Effects}

The \textit{Partial Dependence Plot (PDP)} by
Friedman~\cite{friedman_greedy_2001} describes the marginal effect of one or
two features on the prediction of a model $\hat f$. For a feature set $X_S$
(with $S \subseteq \{1,\ldots,p\}$, $|S| = 1$ or $|S| = 2$), the PDP is defined
as

\begin{equation}
    PDP_{\hat f, S} = \mathbb{E}_{X_C}[\hat{f}(x_S, X_C)] = \int f(x_S, x_C)d\mathbb{P}(x_C),
\end{equation}

\noindent where $X_C$ is the complement feature subset. $PDP_{\hat f, S}$ is a
function of $x_S$ and can be estimated by Monte Carlo integration:

\begin{equation}\label{eq:pdp-estimate}
    \widehat{PDP}_{\hat f, S}(x_S) = \frac{1}{n} \sum_{i=1}^{n} \hat{f}(x_S, x_C^{(i)}).
\end{equation}

\noindent Here, $x_C^{(i)}$ are the actual complement feature values from the dataset of $n$ instances.
To plot this function, a grid of $G$ grid points
$\{(x_S^{(g)}, \widehat{PD}_{\hat f, S}(x_S^{(g)}))\}_{g=1}^G$ can be used~\cite{molnar_relating_2023}.
Molnar et al.~\cite{molnar_general_2022} recommend using quantile-based over equidistant grids.

The PDP assumes that the features in $S$ are independent of the features in
$C$. If this is violated, the perturbations may produce unrealistic data points
outside the underlying joint distribution of the data. This extrapolation issue
can cause misleading
interpretations~\cite{molnar_interpretable_2022,molnar_general_2022}.\\

\noindent The \textit{Accumulated Local Effects (ALE)} plot is an alternative to the PDP that
solves the extrapolation issue~\cite{apley_visualizing_2020}. Using the notation above,
for $|S|=1$, the ALE plot is defined as

\begin{eqnarray}
    ALE_{\hat f,S}(x_S) &=& \int_{x_{\text{min},s}}^{x_S} E_{X_C|X_S}
    \left[\hat{f}^S(X_S, X_C)|X_S = z_S\right] dz_S - \text{constant} \\
    &=& \int_{x_{\text{min},s}}^{x_S} \int_{x_C}
    \hat{f}^S(z_S, x_C)\mathbb{P}(x_C|z_S)dx_{C}dz_{S} - \text{constant},
\end{eqnarray}

\noindent where $\hat{f}^S(x_S, x_C) = \frac{\partial \hat{f}(x_S, x_C)}{\partial x_S}$.
The constant is chosen so that $\widehat{ALE}_{\hat f,S}(X_S)$ is centered with a mean of $0$
w.r.t.\ the marginal distribution of $X_S$. The uncentered ALE can be estimated by

\begin{equation}\label{eq:ale-estimate-uncentered}
    \widehat{\widetilde{ALE}}_{\hat f, S}(x) =
    \sum_{k=1}^{k_S(x)} \frac{1}{n_S(k)} \sum_{i:x_S^{(i)} \in N_S(k)}
    \left[\hat f(z_{k,S}, x_{C}^{(i)}) - \hat f(z_{k-1,S}, x_{C}^{(i)})\right].
\end{equation}

\noindent Here, $\{N_S(k) = (z_{k-1,S}, z_{k,S}]\}_{k=1}^{K}$ partitions the samples
$\{x^{(i)}_S\}_{i=1}^n$ into $K$ intervals or neighborhoods $N_S(k)$. $n_S(k)$ denotes
the number of observations in the $k$th interval $N_S(k)$, $k_S(x)$ represents the index
of the interval to which a particular value $x$ of feature $x_S$ belongs.
The uncentered ALE is centered by

\begin{equation}\label{eq:ale-estimate-centered}
    \widehat{ALE}_{\hat f, S}(x) =
    \widehat{\widetilde{ALE}}_{\hat f, S}(x)
    - \frac{1}{n} \sum_{i=1}^n \widehat{\widetilde{ALE}}_{\hat f, S}(x_S^{(i)})
\end{equation}

\noindent to have a mean effect of $0$. For the grid that defines the intervals,
the quantiles of the empirical distribution of $\{x^{(i)}_S\}_{i=1}^n$ can be
used~\cite{apley_visualizing_2020}.\\

\noindent Other feature effect methods include the \textit{M-Plot (Marginal Plot)},
which, however, suffers from the omitted variable
bias~\cite{apley_visualizing_2020,friedman_greedy_2001},
or \textit{functional ANOVA (fANOVA)}, which decomposes feature effects
into main and interaction effects~\cite{hooker_discovering_2004}.

Furthermore, methods exist that extend previous ones, such as \textit{Robust
    and Heterogeneity-aware ALE (RHALE)}~\cite{gkolemis_rhale_2023}, or
\textit{Accumulated Total Derivative Effect (ATDEV)} plots, which can be
decomposed into ALE and \textit{Accumulated Cross Effects
    (ACE)}~\cite{liu_model_2018}.

In addition to these global feature effect methods, there are regional effect
plots such as \textit{REPID}~\cite{herbinger_repid_2022}, as well as local
methods, including \textit{ICE (Individual Conditional Expectation)}
curves~\cite{goldstein_peeking_2015} or SHAP dependence plots
(e.g.,~\cite{molnar_interpretable_2022}) based on SHAP
values~\cite{lundberg_unified_2017}.

For the remainder of this paper, we will focus on PDP and ALE as the most
popular global feature effect methods and refer to them when speaking of
feature effects.

\subsection{Feature Effect Error Decomposition}

To quantify the error of a computed feature effect, a ``ground truth'' needs to
be defined first. We follow the approach of Molnar et
al.~\cite{molnar_relating_2023} and define ground truth versions of PDP and ALE
directly on the data generating process (DGP) by applying PDP and ALE to the
underlying ground truth function $f$ (instead of model $\hat f$).

For PDP, we can directly use the definition of Molnar et
al.~\cite{molnar_relating_2023}:

\begin{definition}[Definition 1 from~\cite{molnar_relating_2023}]
    The PDP ground truth is the PDP applied to function
    $f : \mathcal{X} \xrightarrow{} \mathcal{Y}$
    of the data generating process.

    \begin{equation}
        PDP_{f,S}(x_S) = \mathbb{E}_{X_C}[f(x_S,X_C)]
    \end{equation}
\end{definition}

As stated by Molnar et al.~\cite{molnar_relating_2023}, their results also
apply to conditional variants of the PDP such as ALE. We now make this
definition explicit:

\begin{definition}
    The ALE ground truth is the ALE applied to function $f : \mathcal{X} \xrightarrow{} \mathcal{Y}$ of the data generating process.

    \begin{equation}
        ALE_{f,S}(x_S) = \int_{x_{\text{min},s}}^{x_S} E_{X_C|X_S} \left[f^S(X_S, X_C)|X_S = z_S\right] dz_S - \text{constant}
    \end{equation}

    \noindent where $f^S(x_S, x_C) = \frac{\partial f(x_S, x_C)}{\partial x_S}$ and \text{constant} chosen such that the effect has a mean of $0$ w.r.t. the marginal distribution of $X_S$.
\end{definition}

%With these definitions, $\widehat{PDP}_{\hat f,S}$ and $\widehat{ALE}_{\hat f,S}$
%can now be treated as statistical estimators of the ground truth feature effects
%(cf.~\cite{molnar_relating_2023}). 
Note that different ground truth effects may also be defined, and our choices
come with certain implications and limitations, such as omitting the
\textit{aggregation bias}\footnote{for details,
    see~\cite{herbinger_repid_2022}}~\cite{mehrabi_survey_2021}.

With more complex ground truth functions $f$, it may become increasingly
difficult to derive the ground truth feature effects analytically, especially
for ALE. In these cases, we therefore propose to also estimate those effects by
Monte Carlo integration, yielding $\widehat{PDP}_{f,S}(x_S)$ and
$\widehat{ALE}_{f,S}(x_S)$ (obtained by plugging in $f$ instead of $\hat f$
into the estimators in equations (\ref{eq:pdp-estimate}) and
(\ref{eq:ale-estimate-uncentered}) / (\ref{eq:ale-estimate-centered})).\\

\noindent Summarizing, we have now defined four quantities per feature effect:
$PDP_{f,S}$ $\widehat{PDP}_{f,S}$, $PDP_{\hat f,S}$, and $\widehat{PDP}_{\hat f,S}$
(analogue for ALE). We can now define different errors between each of these
quantities. In this paper, we focus on the MSE, as it can be decomposed into
bias and variance (see e.g.~\cite{geman_neural_1992}).
Taking, for example, $PDP_{f,S}$ as ground truth, we can define the MSE of
$PDP_{\hat f,S}$ at a point $x_S$ as follows~\cite{molnar_relating_2023}:

\begin{align}
    \text{MSE}(x_S; PDP_{f,S}, PDP_{\hat f,S})
    &= \mathbb{E}_F[{(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f,S}(x_S))}^2]\\
    &= \underbrace{{(PDP_f(x) - \mathbb{E}_F[PDP_{\hat{f}}(x)])}^2}_{Bias^2} + \underbrace{\text{Var}_F[PDP_{\hat{f}}(x)]}_{Variance}
\end{align}

\noindent Here $F$ denotes the distribution of trained models.

% estimation of PDP / ALE + additional MC error

$\mathbb{E}_F$ and $\text{Var}_F$ could be estimated by averaging over
multiple models of the same inducer fitted to $M$ different training data
sets sampled independently from the DGP. In the simulation scenario, this yields:

\begin{align}
    \widehat{\text{MSE}}(x_S; PDP_{f,S}, PDP_{\hat f,S}) &= \frac{1}{M} \sum_{m=1}^{M} {(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f^{(m)},S}(x_S))}^2
\end{align}

% also for bias and variance (including m-1)

% estimation of MC error

% aggregation over x

\section{Methodology \& Experimental Set-Up}\label{sec:methodology-set-up}

\section{Results}\label{sec:results}

\section{Discussion}\label{sec:discussion}

\section{Conclusion}\label{sec:conclusion}

% (see Fig.\@ \ref{fig1})

% \begin{figure}[t]
% \includegraphics[width=\textwidth]{example-image-duck}
% \caption{A figure caption is always placed below the illustration.
% Please note that short captions are centered, while long ones are
% justified by the macro package automatically.} \label{fig1}
% \end{figure}

% \begin{table}[t]
% \caption{Table captions should be placed above the
% tables.}\label{tab1}
% \begin{tabular}{lll}
% \toprule
% Heading level &  Example & Font size and style\\
% \midrule
% Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
% 1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
% 2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
% 3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
% 4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
% \bottomrule
% \end{tabular}
% \end{table}

\begin{credits}
    \subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is used for
    general acknowledgments, for example: This study was funded by X (grant number
    Y).

    \subsubsection{\discintname}
    The authors have no competing interests to declare that are relevant to the
    content of this article.
\end{credits}

%
% ---- Bibliography ----
\bibliographystyle{splncs04}
\bibliography{paper/mybibliography}
\end{document}

### 2025-01-07 paper/paper.tex
paper: add feature effect error chain and error estimators

@@
 \usepackage{booktabs}
+\usepackage{svg}
 \usepackage[misc]{ifsym}
@@
 
-\begin{align}
+\begin{equation}
     \text{MSE}(x_S; PDP_{f,S}, PDP_{\hat f,S})
-    &= \mathbb{E}_F[{(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f,S}(x_S))}^2]\\
-    &= \underbrace{{(PDP_f(x) - \mathbb{E}_F[PDP_{\hat{f}}(x)])}^2}_{Bias^2} + \underbrace{\text{Var}_F[PDP_{\hat{f}}(x)]}_{Variance}
-\end{align}
+    = \mathbb{E}_F[{(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f,S}(x_S))}^2]
+\end{equation}
+\begin{equation}
+    = \underbrace{{(PDP_{f,S}(x_S) - \mathbb{E}_F[PDP_{\hat{f},S}(x_S)])}^2}_{Bias^2} + \underbrace{\text{Var}_F[PDP_{\hat{f},S}(x_S)]}_{Variance}
+\end{equation}
 
-\noindent Here $F$ denotes the distribution of trained models.
+\noindent Here, $F$ denotes the distribution of trained models. The bias
+is linked to the bias of the model, the variance comes from the variance
+in the model fits (randomness in training data, randomness in model training procedure).
 
-% estimation of PDP / ALE + additional MC error
+Since we usually cannot determine $PDP_{\hat f,S}$, we need to estimate it by
+Monte Carlo integration, yielding $\widehat{PDP}_{\hat f,S}(x_S)$. This,
+however, introduces an additional variance term (we use the random variable
+$X_{mc}$ for the Monte Carlo samples (e.g., training or validation data)):
 
-$\mathbb{E}_F$ and $\text{Var}_F$ could be estimated by averaging over
+\begin{equation}
+    \begin{split}
+        \text{MSE}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S})
+        = \underbrace{{(PDP_{f,S}(x_S) - \mathbb{E}_F[PDP_{\hat{f}, S}(x_S)])}^2}_{Bias^2} \\
+        + \underbrace{\text{Var}_F[PDP_{\hat{f},S}(x_S)]}_{Variance} + \underbrace{\mathbb{E}_F\text{Var}_{X_{mc}}[\widehat{PDP}_{\hat{f},S}(x_S)]}_{MC-Variance}
+    \end{split}
+\end{equation}
+
+\begin{proof}
+    For better readability, we will omit the subscript $S$ as well as the point $x_S$ and use $X=X_{mc}$ in this proof:
+
+    \begin{align*}
+        \text{MSE}(PDP_{f}, \widehat{PDP}_{\hat f})
+         & = \mathbb{E}_F\mathbb{E}_X[(PDP_f - \widehat{PDP}_{\hat f})^2]                                                                                         \\
+         & = \mathbb{E}_F \mathbb{E}_X[PDP_f^2 - 2PDP_f\widehat{PDP}_{\hat f} + \widehat{PDP}_{\hat f}^2]                                                         \\
+         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\mathbb{E}_X[\widehat{PDP}_{\hat f}^2]                                                      \\
+         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}] + \mathbb{E}_F[\mathbb{E}_X[\widehat{PDP}_{\hat f}]^2] \\
+         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}] + \text{Var}_F(\mathbb{E}_X[\widehat{PDP}_{\hat f}])   \\
+         & \quad + \mathbb{E}_F(\mathbb{E}_X[\widehat{PDP}_{\hat f}])^2                                                                                           \\
+         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] +  Var_F[PDP_{\hat f}] + \mathbb{E}_F[PDP_{\hat f}]^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}]  \\
+         & = (PDP_f - \mathbb{E}_F[PDP_{\hat f}])^2 + \text{Var}_F[PDP_{\hat f}] + \mathbb{E}_F[\text{Var}_X(\widehat{PDP}_{\hat f})]
+    \end{align*}
+
+    \noindent At multiple points, we use the fact that $\mathbb{E}_X[\widehat{PDP}_{\hat f}]
+        = PDP_{\hat f}$ (cf.~\cite{molnar_relating_2023}).
+\end{proof}
+
+\noindent We see that the variance due to MC integration also depends on the model
+distribution $F$. Similarly, one could use the estimate $\widehat{PDP}_{f,S}$
+as groundtruth, introducing an additional variance term $Var_{X_{mc2}}$when
+estimated on different MC sample (proof in
+\textsc{Appendix~\ref{app:proof-mc-variance}}).
+An overview of all error terms in this chian can be found in
+\textbf{Fig.\@~\ref{fig:error-graph}}.\\
+
+\begin{figure}[t]
+    \includegraphics[width=\textwidth]{img/error_graph.pdf}
+    \caption{Error chain in feature effect estimation
+        (modified, original version can be found in~\cite{molnar_relating_2023})}\label{fig:error-graph}
+\end{figure}
+
+\noindent $\mathbb{E}_F$ and $\text{Var}_F$ could be estimated by averaging over
 multiple models of the same inducer fitted to $M$ different training data
-sets sampled independently from the DGP. In the simulation scenario, this yields:
+sets sampled independently from the DGP. We propose the following estimators:
+\begin{equation}
+    \widehat{\text{MSE}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = \frac{1}{M} \sum_{m=1}^{M} {(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f^{(m)},S}(x_S))}^2
+\end{equation}
+\begin{equation}
+    \widehat{\text{Bias}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = (PDP_{f,S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat{f}^{(m)},S}(x_S))
+\end{equation}
+\begin{equation}
+    \widehat{\text{Variance}}(x_S; \widehat{PDP}_{\hat f,S}) =
+    \frac{1}{M-1}\sum_{m=1}^M\left(\widehat{PDP}_{\hat f^{(m)},S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat f^{(m)},S}(x_S)\right)^2
+\end{equation}
 
-\begin{align}
-    \widehat{\text{MSE}}(x_S; PDP_{f,S}, PDP_{\hat f,S}) &= \frac{1}{M} \sum_{m=1}^{M} {(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f^{(m)},S}(x_S))}^2
-\end{align}
+\noindent Note that these are similar to the approach in~\cite{molnar_relating_2023},
+but we do not specify which data points to use for MC integration. The
+variance captures both the variance in the model fits and the variance due to
+MC integration. To estimate the MC variance, we propose the following estimators:
+\begin{equation}
+    \widehat{\text{Variance}}_{MC}(x_S; PDP_f, \widehat{PDP}_f) = \frac{1}{K}\sum_{k=1}^K(PDP_f(x_S) - \widehat{PDP}_f^{(k)}(x_S))^2
+\end{equation}
 
-% also for bias and variance (including m-1)
+and
+\begin{equation}
+    \begin{split}
+        \widehat{\text{Variance}}_{MC}(x_S; \widehat{PDP}_{\hat{f}}) = \\ \frac{1}{M(K-1)}\sum_{m=1}^M \sum_{k=1}^K
+        \left(\widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S) - \frac{1}{K}\sum_{k=1}^K \widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S)\right)^2
+    \end{split}
+\end{equation}
 
-% estimation of MC error
+\noindent For more convenient analysis of the errors, one could also
+aggregate them over the marginal distribution of $X_S$ (e.g.,
+estimated by averaging over the grid points if chosen appropriately)
+to obtain a single error measure per feature effect.
 
-% aggregation over x
+\section{Overview of Test Functions}\label{sec:test-functions}
 
@@
 \bibliography{paper/mybibliography}
+
+% appendix
+\newpage
+\appendix
+\section{Proof: Monte Carlo Variance for Groundtruth}\label{app:proof-mc-variance}
+\begin{proof}
+    For better readability, we will omit the subscript $S$ as well as the point $x_S$ and use $X=X_{mc}$ in this proof:
+
+    \begin{align*}
+        \mathbb{E}_F\mathbb{E}_X[(PDP_f - \widehat{PDP}_f)^2] & = \mathbb{E}_F\mathbb{E}_X[PDP_f^2 - 2PDP_f \widehat{PDP}_f + \widehat{PDP}_f^2]                                 \\
+                                                              & = PDP_f^2 - 2PDP_f \mathbb{E}_F\mathbb{E}_X[\widehat{PDP}_f] + \mathbb{E}_F\mathbb{E}_X[\widehat{PDP}_f^2]       \\
+                                                              & = PDP_f^2 - 2PDP_f^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_f] + \mathbb{E}_F[\mathbb{E}_X[\widehat{PDP}_f]^2] \\
+                                                              & = PDP_f^2 - 2PDP_f^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_f] + PDP_f^2                                       \\
+                                                              & = \mathbb{E}_F\text{Var}_X[\widehat{PDP}_f]                                                                      \\
+                                                              & = \text{Var}_X[\widehat{PDP}_f]
+    \end{align*}
+
+    \noindent Again we use $\mathbb{E}_X[\widehat{PDP}_{f}] = PDP_{f}$ (cf.~\cite{molnar_relating_2023}) as well as the fact
+    that all quantities based on $f$ do not depend on the model distribution $F$.
+\end{proof}
+
 \end{document}

### 2025-01-07 current_research_feature_effects/main.py
Add catching of errors during model training

@@ -1,5 +1,6 @@
 from configparser import ConfigParser
 from pathlib import Path
+import warnings
 import os
 from typing_extensions import List, Dict, Tuple
 from multiprocessing import Pool, cpu_count
@@ -147,17 +148,25 @@ def simulate(
 
                     model.set_params(**model_params)
 
-                    model.fit(X_train, y_train)
+                    try:
+                        # train model
+                        model.fit(X_train, y_train)
 
-                    # save model
-                    os.makedirs(Path(str(groundtruth)) / config.get("storage", "models"), exist_ok=True)
-                    dump(
-                        model,
-                        Path(os.getcwd()) / str(groundtruth) / config.get("storage", "models") / f"{model_name}.joblib",
-                    )
+                        # save model
+                        os.makedirs(Path(str(groundtruth)) / config.get("storage", "models"), exist_ok=True)
+                        dump(
+                            model,
+                            Path(os.getcwd()) / str(groundtruth) / config.get("storage", "models") / f"{model_name}.joblib",
+                        )
 
-                    # evaluate model
-                    model_results = eval_model(model, X_train, y_train, X_test, y_test)
+                        # evaluate model
+                        model_results = eval_model(model, X_train, y_train, X_test, y_test)
+                    except Exception as e:
+                        model_results = (np.nan,)*6
+                        warnings.warn(
+                            f"Training of model {model_name} failed with error:\n{e}"
+                        )
+
                     df_model_result = pd.DataFrame(
                         {
                             "model_id": [model_name],

### 2025-01-07 current_research_feature_effects/data_generating/data_generation.py
Add theoretical feature effects to simple groundtruth

@@ -139,6 +139,22 @@ class Groundtruth(ABC, BaseEstimator):
             The theoretical partial dependence function for the feature.
         """
 
+    @abstractmethod
+    def get_theoretical_accumulated_local_effects(self, feature: str) -> Callable:
+        """Get the theoretical accumulated local effects function for a feature
+           (up to constant terms).
+
+        Parameters
+        ----------
+        feature : str
+            The feature for which to compute the accumulated local effects function.
+
+        Returns
+        -------
+        Callable
+            The theoretical accumulated local effects function for the feature.
+        """
+
     def get_theoretical_quantiles(self, feature: str, quantiles: np.ndarray) -> np.ndarray:
         """Get theoretical quantiles for a feature's marginal distribution.
 

### 2025-01-07 current_research_feature_effects/data_generating/friedman1.py

@@ -1,4 +1,4 @@
-from typing import Literal, Callable
+from typing import Callable
 import pandas as pd
 import numpy as np
 from scipy.integrate import quad
@@ -36,7 +36,7 @@ class Friedman1Groundtruth(Groundtruth):
 
         return 10 * np.sin(np.pi * X[:, 0] * X[:, 1]) + 20 * (X[:, 2] - 0.5) ** 2 + 10 * X[:, 3] + 5 * X[:, 4]
 
-    def get_theoretical_partial_dependence(self, feature: Literal["x_1", "x_2", "x_3", "x_4", "x_5"]) -> Callable:
+    def get_theoretical_partial_dependence(self, feature: str) -> Callable:
         """Get the theoretical partial dependence function for a feature.
         Only uniform feature distribution is supported.
 
@@ -89,3 +89,19 @@ class Friedman1Groundtruth(Groundtruth):
                 return complete_integral_12 + integral_3 + 5 + 5 * x5
 
         return partial_dependence
+
+    def get_theoretical_accumulated_local_effects(self, feature: str) -> Callable:
+        """Get the theoretical accumulated local effects function for a feature.
+        Only uniform feature distribution is supported.
+
+        Parameters
+        ----------
+        feature : str
+            The feature for which to compute the accumulated local effects function.
+
+        Returns
+        -------
+        Callable
+            The theoretical accumulated local effects function for the feature.
+        """
+        raise NotImplementedError("Theoretical accumulated local effects are not implemented for Friedman1.")

### 2025-01-07 current_research_feature_effects/data_generating/physics_data.py

@@ -1,4 +1,4 @@
-from typing import Literal, Callable
+from typing import Callable
 import numpy as np
 import pandas as pd
 from scipy import constants
@@ -37,10 +37,13 @@ class NewtonUniversalGravitationGroundtruth(Groundtruth):
             / ((X[:, 2] - X[:, 3]) ** 2 + (X[:, 4] - X[:, 5]) ** 2 + (X[:, 6] - X[:, 7]) ** 2)
         )
 
-    def get_theoretical_partial_dependence(
-        self, feature: Literal["m_1", "m_2", "x_1", "x_2", "y_1", "y_2", "z_1", "z_2"]
-    ) -> Callable:
+    def get_theoretical_partial_dependence(self, feature: str) -> Callable:
         raise NotImplementedError("Theoretical partial dependence not implemented for NewtonUniversalGravitation.")
+
+    def get_theoretical_accumulated_local_effects(self, feature: str) -> Callable:
+        raise NotImplementedError(
+            "Theoretical accumulated local effects not implemented for NewtonUniversalGravitation."
+        )
 
 
 class WaveInterferenceGroundtruth(Groundtruth):
@@ -73,5 +76,8 @@ class WaveInterferenceGroundtruth(Groundtruth):
 
         return np.sqrt(x1**2 + x2**2 + 2 * x1 * x2 * np.cos(theta1 - theta2))
 
-    def get_theoretical_partial_dependence(self, feature: Literal["x_1", "x_2", "theta_1", "theta_2"]) -> Callable:
+    def get_theoretical_partial_dependence(self, feature: str) -> Callable:
         raise NotImplementedError("Theoretical partial dependence not implemented for WaveInterference.")
+
+    def get_theoretical_accumulated_local_effects(self, feature: str) -> Callable:
+        raise NotImplementedError("Theoretical accumulated local effects not implemented for WaveInterference.")

### 2025-01-07 current_research_feature_effects/data_generating/simple.py

@@ -1,4 +1,4 @@
-from typing import Callable, Literal
+from typing import Callable
 import pandas as pd
 import numpy as np
 
@@ -29,8 +29,13 @@ class SimpleAdditiveGroundtruth(Groundtruth):
             X = X.values
         return X[:, 0] + 0.5 * X[:, 1] ** 2
 
-    def get_theoretical_partial_dependence(self, feature: Literal["x_1", "x_2"]) -> Callable:
+    def get_theoretical_partial_dependence(self, feature: str) -> Callable:
         raise NotImplementedError("Theoretical partial dependence not implemented for SimpleAdditiveGroundtruth.")
+
+    def get_theoretical_accumulated_local_effects(self, feature: str) -> Callable:
+        raise NotImplementedError(
+            "Theoretical accumulated local effects not implemented for SimpleAdditiveGroundtruth."
+        )
 
 
 class SimpleInteractionGroundtruth(Groundtruth):
@@ -57,8 +62,13 @@ class SimpleInteractionGroundtruth(Groundtruth):
             X = X.values
         return X[:, 0] * X[:, 1]
 
-    def get_theoretical_partial_dependence(self, feature: Literal["x_1", "x_2"]) -> Callable:
+    def get_theoretical_partial_dependence(self, feature: str) -> Callable:
         raise NotImplementedError("Theoretical partial dependence not implemented for SimpleInteractionGroundtruth.")
+
+    def get_theoretical_accumulated_local_effects(self, feature: str) -> Callable:
+        raise NotImplementedError(
+            "Theoretical accumulated local effects not implemented for SimpleInteractionGroundtruth."
+        )
 
 
 class SimpleCombinedGroundtruth(Groundtruth):
@@ -85,5 +95,58 @@ class SimpleCombinedGroundtruth(Groundtruth):
             X = X.values
         return X[:, 0] + 0.5 * X[:, 1] ** 2 + X[:, 0] * X[:, 1]
 
-    def get_theoretical_partial_dependence(self, feature: Literal["x_1", "x_2"]) -> Callable:
-        raise NotImplementedError("Theoretical partial dependence not implemented for SimpleCombinedGroundtruth.")
+    def get_theoretical_partial_dependence(self, feature: str) -> Callable:
+        marginal_distrs = [feature[0] for feature in self.marginal_distributions]
+        if not all(distr == "normal" for distr in marginal_distrs):
+            raise NotImplementedError("Only normal marginals are supported.")
+
[…]

### 2025-01-07 current_research_feature_effects/feature_effects.py

@@ -1,9 +1,11 @@
-from typing_extensions import List, Dict, Callable, Tuple
+from typing_extensions import List, Dict, Callable, Tuple, Literal
 from copy import deepcopy
 import pandas as pd
 import numpy as np
 from sklearn.base import BaseEstimator
 from sklearn.model_selection import KFold
+
+from current_research_feature_effects.data_generating.data_generation import Groundtruth
 
 
 def _partial_dependence(estimator: BaseEstimator, X: np.ndarray, feature: int, grid: np.ndarray) -> Dict:
@@ -259,6 +261,68 @@ def compute_ales(
         )
 
     return ales
+
+
+def compute_theoretical_effects(
+    groundtruth: Groundtruth,
+    effect: Literal["pdp", "ale"],
+    feature_names: List[str],
+    grid_values: List[np.ndarray],
+    center_curves: bool = False,
+    remove_first_last: bool = False,
+) -> List[Dict]:
+    """
+    Compute theoretical partial dependence plots for a given groundtruth
+    and apply it to the grid_values.
+
+    Parameters
+    ----------
+    groundtruth : Groundtruth
+        Groundtruth object to compute theoretical partial dependence for.
+    effect : Literal["pdp", "ale"]
+        Type of effect to compute: partial dependence or accumulated local effects.
+    feature_names : List[str]
+        Names of features to compute theoretical partial dependence for.
+    grid_values : List[np.ndarray]
+        Grid of values to compute theoretical partial dependence for.
+    center_curves : bool, optional
+        Whether to center curves by substracting the mean (based on grid values) or not, by default False.
+    remove_first_last : bool, optional
+        Whether to remove the first and last grid values from the effects before centering and
+        returning or not, by default False.
+
+    Returns
+    -------
+    List[Dict]
+        List of dictionaries containing the feature name, grid values and computed partial dependence values.
+    """
+    effects = []
+    for f_name, grid in zip(feature_names, grid_values):
+        if effect == "pdp":
+            feature_effect = groundtruth.get_theoretical_partial_dependence(feature=f_name)(grid)
+        elif effect == "ale":
+            feature_effect = groundtruth.get_theoretical_accumulated_local_effects(feature=f_name)(grid)
+        else:
+            raise ValueError("Effect type not supported")
+
+        grid_ = grid
+
+        if remove_first_last:
+            feature_effect = feature_effect[1:-1]
+            grid_ = grid_[1:-1]
+
+        if center_curves:
[…]

### 2025-01-16 scripts/ablation_study_mc.py
code: add mc ablation study results

@@ -90,13 +90,13 @@ def perform_mc_ablation_study(groundtruth: Groundtruth, study_params: Dict):
         results["ale"][n_samples] = {k: v / study_params["k"] for k, v in ale_variances.items()}
 
     # save results, groundtruth and grid values
-    os.makedirs(EXPERIMENT_PATH / str(groundtruth), exist_ok=True)
-    dump(results, EXPERIMENT_PATH / str(groundtruth) / "ablation_results.joblib")
-    dump(groundtruth, EXPERIMENT_PATH / str(groundtruth) / "groundtruth.joblib")
+    os.makedirs(Path(str(groundtruth)), exist_ok=True)
+    dump(results, Path(str(groundtruth)) / "ablation_results.joblib")
+    dump(groundtruth, Path(str(groundtruth)) / "groundtruth.joblib")
     if study_params["remove_first_last"]:
-        dump([grid[1:-1] for grid in grid_values], EXPERIMENT_PATH / str(groundtruth) / "grid_values.joblib")
+        dump([grid[1:-1] for grid in grid_values], Path(str(groundtruth)) / "grid_values.joblib")
     else:
-        dump(grid_values, EXPERIMENT_PATH / str(groundtruth) / "grid_values.joblib")
+        dump(grid_values, Path(str(groundtruth)) / "grid_values.joblib")
 
 
 if __name__ == "__main__":
@@ -144,6 +144,7 @@ if __name__ == "__main__":
         "remove_first_last": config.getboolean("study_params", "remove_first_last"),
     }
 
+    os.chdir(EXPERIMENT_PATH)
     num_processes = min(len(groundtruths), cpu_count())
 
     # perform ablation studies in parallel

### 2025-01-18 scripts/main.py
code: restructure main study script and config with arguments

@@ -1,3 +1,4 @@
+import argparse
 from configparser import ConfigParser
 from pathlib import Path
 import warnings
@@ -25,11 +26,7 @@ from current_research_feature_effects.feature_effects import (
     compute_ales,
     compute_cv_feature_effect,
     compare_effects,
-    get_modified_grids,
 )
-
-sim_config = ConfigParser()
-sim_config.read("config.ini")
 
 
 def simulate(
@@ -53,7 +50,7 @@ def simulate(
     feature_names = groundtruth.feature_names
     grid_points = config.getint("feature_effects", "grid_points")
     quantiles = np.linspace(0, 1, grid_points, endpoint=True)
-    base_grids = [groundtruth.get_theoretical_quantiles(feature, quantiles) for feature in feature_names]
+    grid_values = [groundtruth.get_theoretical_quantiles(feature, quantiles) for feature in feature_names]
     center_curves = config["feature_effects"].getboolean("centered")
     remove_first_last = config["feature_effects"].getboolean("remove_first_last")
 
@@ -65,6 +62,8 @@ def simulate(
         snr=0,
         seed=config.getint("simulation_metadata", "mc_data_seed"),
     )
+
+    k_cv = config.getint("simulation_metadata", "k_cv")
 
     for sim_no in range(n_sim):
         for n_train, n_val in n_train_vals:
@@ -79,32 +78,26 @@ def simulate(
                     seed=sim_no,
                 )
 
-                # cv = KFold(n_splits=5, shuffle=True, random_state=42)
-                # X_all, y_all = np.concatenate([X_train, X_val], axis=0), np.concatenate([y_train, y_val], axis=0)
-                # cv_splits = cv.split(X=X_all, y=y_all)
-                # val_folds = [X_all[split[1]] for split in cv_splits]
+                cv = KFold(n_splits=5, shuffle=True, random_state=42)
+                X_all, y_all = np.concatenate([X_train, X_val], axis=0), np.concatenate([y_train, y_val], axis=0)
 
-                # train_grid, val_grid, mc_grid, *cv_grids = get_modified_grids(
-                #     base_grids=base_grids, Xs=[X_train, X_val, X_mc] + val_folds, feature_names=feature_names
-                # )
+                pdp_groundtruth = compute_pdps(
+                    groundtruth,
+                    X_mc,
+                    feature_names,
+                    grid_values=grid_values,
+                    center_curves=center_curves,
+                    remove_first_last=remove_first_last,
+                )
 
-                # pdp_groundtruth = compute_pdps(
[…]

### 2025-01-19 paper/paper.tex
paper: add chapter on test functions for simulation studies

@@
 \begin{eqnarray}
-    ALE_{\hat f,S}(x_S) &=& \int_{x_{\text{min},s}}^{x_S} E_{X_C|X_S}
+    ALE_{\hat f,S}(x_S) &=& \int_{x_{\text{min},s}}^{x_S} \mathbb{E}_{X_C|X_S}
     \left[\hat{f}^S(X_S, X_C)|X_S = z_S\right] dz_S - \text{constant} \\
@@
     \begin{equation}
-        ALE_{f,S}(x_S) = \int_{x_{\text{min},s}}^{x_S} E_{X_C|X_S} \left[f^S(X_S, X_C)|X_S = z_S\right] dz_S - \text{constant}
+        ALE_{f,S}(x_S) = \int_{x_{\text{min},s}}^{x_S} \mathbb{E}_{X_C|X_S} \left[f^S(X_S, X_C)|X_S = z_S\right] dz_S - \text{constant}
     \end{equation}
@@
 
-\section{Overview of Test Functions}\label{sec:test-functions}
+While our definitions are based on the PDP, they can be directly applied to the
+ALE as well.
+
+\section{Test Functions for Simulation Studies}\label{sec:test-functions}
+
+Test functions play a crucial role in research, e.g.\ for evaluating and
+comparing different methodological approaches, or when conducting simulation
+studies. This section synthesizes commonly used test functions across various
+domains, providing structured guidance for researchers --- particularly those
+in Interpretable ML --- in selecting appropriate test functions for their
+simulation studies. By examining test functions from different fields and their
+purposes of application, we aim to facilitate experimental design decisions.
+While this overview is not exhaustive, it highlights key test functions and
+their characteristics to help researchers make informed choices for their
+specific research needs.
+
+\paragraph{Test Functions in Optimization.}
+The field of optimization, where test functions are essential to enable the
+assessment and comparison of optimization algorithms, has established a rich
+foundation of test functions. A fundamental approach involves using simple
+mathematical expressions like the sphere function~\cite{more_testing_1981}.
+These basic functions are often combined with more complex ones like Branin or
+Rosenbrock to create comprehensive test suites that incorporate important
+properties such as nonlinearity, non-separability, and
+scalability~\cite{whitley_building_1998}. A notable framework in this domain is
+the Comparing Continuous Optimizer (COCO) platform with its Black Box
+Optimization Benchmark (BBOB), offering a structured approach to testing
+continuous optimization algorithms through artificial test
+functions~\cite{hansen_coco_2016}. These classical test function suites are
+well-established in optimization and may also serve as a basis for
+Interpretable ML researchers. However, it is important to note that the ability
+of these artifical test functions to represent complex real-world behavior is
+limited~\cite{zaefferer_simulation-based_2017}.
+
+\paragraph{Physics-Inspired Test Functions.}
+
+Physics-derived functions offer a compelling source of real-world test cases,
+with the Feynman Symbolic Regression Database (FSReD) being a prominent
+example. FSReD comprises 100 physics equations from the seminal
+\textit{Feynman's Lectures on Physics (34--36)}, supplemented by 20 more
+challenging equations from other seminal physics texts~\cite{udrescu_ai_2020}.
+These equations span diverse physics domains including classical mechanics,
+electromagnetism, and quantum mechanics. They involve between one and nine
+independent variables and incorporate various elementary functions such as
+arithmetic operations, trigonometric functions, and exponentials. From that,
+tabular datasets were generated through random sampling from defined value
+ranges.
+
+Building upon this foundation,
+Matsubara~et~al.~\cite{matsubara_rethinking_2024} addressed several limitations
+of the original FSReD. Their enhanced framework introduces a three-tiered
+categorization of problems (easy, medium, hard) based on their complexity,
+incorporates dummy variables to simulate irrelevant features, and implements
+more realistic sampling ranges and strategies. Detailed specifications for all
+formulas, including their sampling parameters, are available in their work.
+
+While initially developed for symbolic regression tasks, many Feynman equations
+may serve as suitable test functions for simulation studies in Interpretable
+ML. Their basis in physical principles provides real-world relevance, though
+researchers should carefully select equations that align with their specific
+analytical objectives and complexity requirements.
+
+\paragraph{Test Functions for Interpretable Machine Learning.}
+The Interpretable ML field itself has developed several specialized test
+functions designed to evaluate specific aspects of interpretability methods.
+
+Goldstein et al.~\cite{goldstein_peeking_2015} used several simple test
+functions to demonstrate the behavior of Individual Conditional Expectation
+(ICE) curves. These include a simple additive function to demonstrate the
+absence of interactions, simple interactions to reveal heterogeneity that might
+be obscured by averaging procedures such as PDPs, and a specially designed
+function with an empty quadrant for assessing extrapolation behavior.
+
+Similarly, Liu et al.~\cite{liu_model_2018} focus on simple functions before
+progressing to more complex ones. They begin with basic two-variable scenarios
+--- using additive functions, interaction functions, and combinations thereof
+--- and examine these under both independent and correlated feature conditions
+to compare various feature effect methods. The advantage of these simple test
+functions is that solutions (e.g., feature effects) can also be computed
+analytically, and that they allow for deeper and more fine-grained analysis of
+individual aspects.
+
+A more complex test function suite was proposed by
+Tsang~et~al.~\cite{tsang_detecting_2017}, specifically designed to evaluate the
+detection of variable interactions. Their functions incorporate various types
+of interactions with different orders, strengths, non-linearities, and
+overlaps. While this makes them particularly valuable for interaction
+detection, they are also useful for evaluating other interpretability methods
+in scenarios with complex interactions.
+
+The Friedman functions~\cite{breiman_bagging_1996,friedman_multivariate_1991}
+serve as classical benchmarks applicable across various interpretability tasks.
+These three functions combine linear and non-linear effects with interactions,
+incorporating dummy variables and random noise terms to reflect realistic
+complexity. For detailed specifications, see~\cite{breiman_bagging_1996}.
+
+When choosing test functions for simulation studies in Interpretable ML,
+researchers should consider several criteria, including the specific aspects of
+interpretability being evaluated, the desired complexity level and number of
+variables, the presence of specific challenges such as correlation between
+features or interactions, the need for analytical solutions for validation, as
+well as the relevance to real-world applications in the domain of interest.
 
@@
 \section{Conclusion}\label{sec:conclusion}
-
-% (see Fig.\@ \ref{fig1})
-
-% \begin{figure}[t]
-% \includegraphics[width=\textwidth]{example-image-duck}
-% \caption{A figure caption is always placed below the illustration.
-% Please note that short captions are centered, while long ones are
-% justified by the macro package automatically.} \label{fig1}
-% \end{figure}
 
@@
 \bibliographystyle{splncs04}
-\bibliography{paper/mybibliography}
+\bibliography{mybibliography}
 

### 2025-01-19 current_research_feature_effects/feature_effects.py
code: add feature effect class

@@ -1,11 +1,131 @@
-from typing_extensions import List, Dict, Callable, Tuple, Literal
+from typing_extensions import List, Dict, Callable, Tuple, Literal, Union
 from copy import deepcopy
+from dataclasses import dataclass
 import pandas as pd
 import numpy as np
 from sklearn.base import BaseEstimator
 from sklearn.model_selection import KFold
 
 from current_research_feature_effects.data_generating.data_generation import Groundtruth
+
+
+EffectType = Literal["pdp", "ale"]
+
+
+@dataclass
+class FeatureEffect:
+    """
+    Class to store and manipulate feature effects (PDP or ALE).
+    """
+
+    effect_type: EffectType
+    features: Dict[str, Dict[str, np.ndarray]]  # {feature_name: {'grid': array, 'effect': array}}
+
+    def __init__(
+        self,
+        effect_type: EffectType,
+        feature_effects: List[Dict],  # List of dicts with 'feature', 'grid_values', 'effect' keys
+    ):
+        """
+        Initialize from list of feature effect dictionaries.
+
+        Parameters
+        ----------
+        effect_type : EffectType
+            Type of effect (PDP or ALE).
+        feature_effects : List[Dict]
+            List of dictionaries containing feature effect information.
+        """
+        self.effect_type = effect_type
+        self.features = {
+            effect["feature"]: {"grid": np.array(effect["grid_values"]), "effect": np.array(effect["effect"])}
+            for effect in feature_effects
+        }
+
+    def __repr__(self) -> str:
+        features_str = ", ".join(self.features.keys())
+        return f"FeatureEffect(type={self.effect_type}, features=[{features_str}])"
+
+    def _validate_operation(self, other: "FeatureEffect") -> None:
+        """Validate that two FeatureEffects can be combined."""
+        if not isinstance(other, FeatureEffect):
+            raise TypeError(f"Unsupported operand type: {type(other)}")
+
+        if self.effect_type != other.effect_type:
+            raise ValueError(f"Cannot combine different effect types: {self.effect_type} and {other.effect_type}")
+
+        if set(self.features.keys()) != set(other.features.keys()):
+            raise ValueError("Feature sets must match for arithmetic operations")
+
+        for feature in self.features:
+            if not np.array_equal(self.features[feature]["grid"], other.features[feature]["grid"]):
+                raise ValueError(f"Grid values must match for feature {feature}")
+
[…]

### 2025-01-19 current_research_feature_effects/model_training.py
code: outsource model initialization

@@ -1,11 +1,14 @@
-from typing_extensions import Literal
+from typing_extensions import Literal, Dict
 from pathlib import Path
+from configparser import ConfigParser
+import os
 import numpy as np
 import optuna
 from sklearn.base import BaseEstimator
 from sklearn.metrics import mean_squared_error, mean_absolute_error
 
 from current_research_feature_effects.mappings import suggested_hps_for_model
+from current_research_feature_effects.data_generating.data_generation import generate_data, Groundtruth
 
 
 def _objective(
@@ -63,3 +66,74 @@ def optimize(
     study.optimize(objective, n_trials=n_trials, catch=(np.linalg.LinAlgError, RuntimeError))
 
     return study
+
+
+def initialize_model(
+    model_dict: Dict, model_str: str, groundtruth: Groundtruth, n_train: int, snr: float, config: ConfigParser
+) -> BaseEstimator:
+    """Initialize model with hyperparameters (tuned or default).
+
+    This function initializes a model with hyperparameters. If the hyperparameters are set to "to_tune" in the model
+    dictionary, the function will either load the best hyperparameters from a tuning study or run a new tuning study
+    if the study does not exist. The function will then set the hyperparameters of the model and return it.
+
+    Parameters
+    ----------
+    model_dict : Dict
+        Dictionary containing the model, and the hyperparameters to tune per dataset and dataset size.
+    model_str : str
+        Name of the model.
+    groundtruth : Groundtruth
+        Groundtruth object of the dataset.
+    n_train : int
+        Number of training samples.
+    snr : float
+        Signal-to-noise ratio.
+    config : ConfigParser
+        Config containing simulation metadata.
+
+    Returns
+    -------
+    BaseEstimator
+        Initialized model.
+    """
+    model: BaseEstimator = model_dict["model"]
+    if model_dict["model_params"][str(groundtruth)][n_train] == "to_tune":
+        tuning_studies_dir = config.get("storage", "tuning_studies_folder")
+        os.makedirs(Path(os.getcwd()) / str(groundtruth) / tuning_studies_dir, exist_ok=True)
+        study_name = f"{model_str}_{n_train}_{int(snr)}"
+        storage_name = f"sqlite:///{str(groundtruth)}/{tuning_studies_dir}/{model_str}.db"
+        try:
+            model_params = optuna.load_study(
+                study_name=study_name,
+                storage=storage_name,
+            ).best_params
+        except KeyError as e:
+            if str(e) == "'Record does not exist.'":
[…]

### 2025-01-19 current_research_feature_effects/feature_effects.py
code: add feature effect metric computation

@@ -509,3 +509,36 @@ def compute_cv_feature_effect(
     averaged_effects /= len(effects)
 
     return averaged_effects
+
+
+def compute_feature_effect_metrics(estimates: List[FeatureEffect], groundtruth: FeatureEffect):
+    """
+    Compute pointwise metrics to compare feature effects estimates with groundtruth.
+
+    Parameters
+    ----------
+    estimates : List[FeatureEffect]
+        List of feature effect estimates.
+    groundtruth : FeatureEffect
+        Groundtruth feature effect.
+
+    Returns
+    -------
+    Dict
+        Dictionary containing the computed metrics.
+    """
+    metrics = {}
+
+    mean_estimate = sum(estimates) / len(estimates)
+
+    for estimate in estimates:
+        se = (estimate - groundtruth) ** 2
+        sd = (estimate - mean_estimate) ** 2
+        metrics["MSE"] = se if "MSE" not in metrics else metrics["MSE"] + se
+        metrics["Variance"] = sd if "Variance" not in metrics else metrics["Variance"] + sd
+
+    metrics["MSE"] /= len(estimates)
+    metrics["Variance"] /= len(estimates)
+    metrics["Bias^2"] = (mean_estimate - groundtruth) ** 2
+
+    return metrics

### 2025-01-19 current_research_feature_effects/utils.py

@@ -34,7 +34,7 @@ def parse_sim_params(sim_config: ConfigParser) -> Dict:
             sim_config.get("simulation_params", "n_val").split(","),
         )
     ]
-    param_dict["snr"] = [float(x) for x in sim_config.get("simulation_params", "snr").split(",")]
+    param_dict["snr"] = sim_config.getfloat("simulation_params", "snr")
 
     with open(models_config_path, "r") as file:
         models_config: Dict = yaml.safe_load(file)

### 2025-01-19 scripts/main.py

@@ -9,13 +9,10 @@ from joblib import dump
 import numpy as np
 import pandas as pd
 from sqlalchemy import create_engine
-from sklearn.base import BaseEstimator
-from sklearn.metrics import mean_squared_error
 from sklearn.model_selection import KFold
-import optuna
 
 from current_research_feature_effects.data_generating.data_generation import generate_data, Groundtruth
-from current_research_feature_effects.model_training import optimize
+from current_research_feature_effects.model_training import initialize_model
 from current_research_feature_effects.model_eval import eval_model
 from current_research_feature_effects.utils import (
     parse_sim_params,
@@ -25,7 +22,7 @@ from current_research_feature_effects.feature_effects import (
     compute_pdps,
     compute_ales,
     compute_cv_feature_effect,
-    compare_effects,
+    compute_feature_effect_metrics,
 )
 
 
@@ -34,7 +31,7 @@ def simulate(
     groundtruth: Groundtruth,
     n_sim: int,
     n_train_vals: List[Tuple[int, int]],
-    snrs: List[float],
+    snr: float,
     config: ConfigParser,
 ):
     np.random.seed(42)
@@ -55,7 +52,7 @@ def simulate(
     remove_first_last = config["feature_effects"].getboolean("remove_first_last")
 
     # sample data for MC approximation of groundtruth feature effects
-    X_mc, y_mc, _, _ = generate_data(
+    X_mc, _, _, _ = generate_data(
         groundtruth=groundtruth,
         n_train=config.getint("simulation_metadata", "n_mc"),
         n_test=1,
@@ -65,217 +62,184 @@ def simulate(
 
     k_cv = config.getint("simulation_metadata", "k_cv")
 
-    for sim_no in range(n_sim):
-        for n_train, n_val in n_train_vals:
-            for snr in snrs:
-                # generate data
-                X_train, y_train, X_val, y_val, X_test, y_test = generate_data(
-                    groundtruth=groundtruth,
-                    n_train=n_train,
-                    n_val=n_val,
-                    n_test=config.getint("simulation_metadata", "n_test"),
-                    snr=snr,
-                    seed=sim_no,
-                )
+    pdps_train = {}
+    pdps_val = {}
+    pdps_cv = {}
 
-                cv = KFold(n_splits=5, shuffle=True, random_state=42)
-                X_all, y_all = np.concatenate([X_train, X_val], axis=0), np.concatenate([y_train, y_val], axis=0)
+    ales_train = {}
+    ales_val = {}
+    ales_cv = {}
 
[…]

### 2025-01-30 current_research_feature_effects/model_training.py
Fix optuna DB conflict

@@ -102,7 +102,7 @@ def initialize_model(
         tuning_studies_dir = config.get("storage", "tuning_studies_folder")
         os.makedirs(Path(os.getcwd()) / str(groundtruth) / tuning_studies_dir, exist_ok=True)
         study_name = f"{model_str}_{n_train}_{int(snr)}"
-        storage_name = f"sqlite:///{str(groundtruth)}/{tuning_studies_dir}/{model_str}.db"
+        storage_name = f"sqlite:///{str(groundtruth)}/{tuning_studies_dir}/{model_str}_{n_train}_{int(snr)}.db"
         try:
             model_params = optuna.load_study(
                 study_name=study_name,

### 2025-01-30 current_research_feature_effects/plotting/table_formatting.py
Add plotting and table formatting utils functions for feature effect error analysis

@@ -1,0 +1,69 @@
+import pandas as pd
+import numpy as np
+
+
+def format_dataframe(df: pd.DataFrame) -> pd.DataFrame:
+    """
+    Format the dataframe to a more readable format for the table.
+
+    Parameters
+    ----------
+    df : pd.DataFrame
+        Dataframe containing the results of the model training.
+
+    Returns
+    -------
+    pd.DataFrame
+        Formatted dataframe.
+    """
+    pivoted = df.pivot_table(
+        index=["model", "n_train", "split"], columns=["feature", "metric"], values="value"
+    ).reset_index()
+
+    pivoted.columns.name = None
+    df_out = pivoted.groupby(by=["model", "n_train", "split"]).mean()
+    sorted_cols = [
+        (feature, metric)
+        for feature in sorted(set(col[0] for col in df_out.columns))
+        for metric in ["MSE", "Bias^2", "Variance"]
+    ]
+    sorted_idx = [
+        (model, n, split)
+        for model in ["LinReg", "GAM_OF", "GAM_OT", "SVM_OF", "SVM_OT", "XGBoost_OF", "XGBoost_OT"]
+        for n in sorted(df_out.index.get_level_values("n_train").unique())
+        for split in ["train", "val", "cv"]
+    ]
+    df_out = df_out[sorted_cols].reindex(pd.MultiIndex.from_tuples(sorted_idx))
+
+    return df_out
+
+
+def highlight_min_feature_metric(data: pd.DataFrame) -> pd.DataFrame:
+    """
+    Highlight the minimum value in each feature-metric pair.
+
+    Parameters
+    ----------
+    data : pd.DataFrame
+        Dataframe containing the results of the feature effect analysis.
+
+    Returns
+    -------
+    pd.DataFrame
+        Dataframe with the minimum values highlighted.
+    """
+    mask = pd.DataFrame(False, index=data.index, columns=data.columns)
+    groups = data.index.droplevel(2).unique()
+
+    for model_n_train in groups:
+        group_mask = (data.index.get_level_values(0) == model_n_train[0]) & (
+            data.index.get_level_values(1) == model_n_train[1]
+        )
+
+        for feature in data.columns.get_level_values(0).unique():
+            for metric in data.columns.get_level_values(1).unique():
+                col = (feature, metric)
+                is_min = data[col][group_mask] == data[col][group_mask].min()
+                mask.loc[group_mask, col] = is_min
+
[…]

### 2025-02-01 current_research_feature_effects/utils.py
Fix parallelization bug for model saving

@@ -12,6 +12,7 @@ import yaml
 import multiprocessing
 from logging.handlers import QueueHandler, QueueListener
 import sys
+from time import sleep
 import numpy as np
 import pandas as pd
 from sqlalchemy import Engine
@@ -138,7 +139,14 @@ def create_and_set_sim_dir(sim_config: ConfigParser, config_path: Path) -> None:
         raise ValueError(f"Simulation {base_dir} already exists.")
 
 
-def save_model_results(model_metrics: Dict[str, float], conn: Engine, params: SimulationParameter, sim_no: int):
+def save_model_results(
+    model_metrics: Dict[str, float],
+    conn: Engine,
+    params: SimulationParameter,
+    sim_no: int,
+    max_retries: int = 10,
+    retry_delay: float = 0.1,
+) -> bool:
     """
     Save model results to database.
 
@@ -152,6 +160,15 @@ def save_model_results(model_metrics: Dict[str, float], conn: Engine, params: Si
         Simulation parameters.
     sim_no : int
         Simulation number.
+    max_retries : int, optional
+        Maximum number of retries, by default 10
+    retry_delay : float, optional
+        Delay between retries, by default 0.1
+
+    Returns
+    -------
+    bool
+        True if results were saved successfully.
     """
     df_model_result = pd.DataFrame(
         {
@@ -163,11 +180,18 @@ def save_model_results(model_metrics: Dict[str, float], conn: Engine, params: Si
         | model_metrics
     )
 
-    df_model_result.to_sql(
-        "model_results",
-        con=conn,
-        if_exists="append",
-    )
+    for attempt in range(max_retries):
+        try:
+            df_model_result.to_sql(
+                "model_results",
+                con=conn,
+                if_exists="append",
+            )
+            return True
+        except Exception as e:
+            if attempt == max_retries - 1:
+                raise Exception(f"Failed to save results after {max_retries} attempts: {e}")
+            sleep(retry_delay * (2**attempt))  # Exponential backoff
 
 
 def save_fe_aggregated_results(

### 2025-02-01 paper/paper.tex
paper: add model hyperparameters and performances

@@
 \usepackage{amssymb}
+\usepackage{enumitem}
+\usepackage{multirow}
+\usepackage{tabularx}
+\usepackage{ragged2e}
 \usepackage{booktabs} 
@@
 
-\title{Should We Compute Feature Effects on Training or Validation Data?}
+\title{Understanding Implications of Dataset Choice for Feature Effect Estimation:
+    A Simulation-Based Investigation through Error Decomposition}
 
-% \titlerunning{Should We Compute Feature Effects on Training or Validation Data?}
+\titlerunning{Understanding Implications of Dataset Choice for Feature Effect Estimation}
 % If the full title of your paper is short enough to also fit in the running head, you can omit the abbreviated paper title here.
@@
 \textsc{Appendix~\ref{app:proof-mc-variance}}).
-An overview of all error terms in this chian can be found in
+An overview of all error terms in this chain can be found in
 \textbf{Fig.\@~\ref{fig:error-graph}}.\\
@@
 
-and
+\noindent and
 \begin{equation}
@@
 
+\paragraph{Datasets.}
+Building upon \textsc{Section~\ref{sec:test-functions}}, we employ three
+distinct datasets of varying complexity for our simulation study:
+
+\begin{itemize}[label=--]
+    \item \textbf{SimpleNormalCorrelated} consists of four
+          standard-normally distributed features, where the first two
+          features exhibit strong correlation ($\rho = 0.9$) while the others are
+          independent dummy variables. The target variable is given by this simple
+          formula:
+          \begin{equation}
+              f_1(\boldsymbol{x}) = x_1 + \frac{x_2^2}{2} + x_1 x_2
+          \end{equation}
+          This test function is inspired by~\cite{liu_model_2018} and aims
+          to focus on the impact of correlation and interactions.
+    \item \textbf{Friedman1} implements the classical Friedman1 benchmark
+          function~\cite{breiman_bagging_1996,friedman_multivariate_1991}
+          with seven uniformly distributed features between 0 and 1, all mutually
+          independent:
+          \begin{equation}
+              f_2(\boldsymbol{x}) = 10 \sin(\pi x_1 x_2) + 20(x_3 - \frac{1}{2})^2 + 10 x_4 + 5 x_5
+          \end{equation}
+          This dataset includes a mix of linear and different non-linear effects with
+          interactions.
+    \item \textbf{Feynman I.29.16} is based on the Feynman equation I.29.16
+          describing wave interference, comprising six independent features. Using
+          the refined sampling strategies from Matsubara et al. \cite{matsubara_rethinking_2024},
+          two log-uniformly distributed variables on $[0.1, 10]$, two angles uniformly
+          distributed over $[0, 2\pi]$, and two uniformly distributed dummy variables on [0,1]:
+          \begin{equation}
+              f_3(\boldsymbol{x}) = \sqrt{x_1^2 + x_2^2 + 2 x_1 x_2 \cos(\theta_1 - \theta_2)}
+          \end{equation}
+          This dataset is designed to reflect a physics-based relationship for more real-world relevance.
+
+\end{itemize}
+
+\noindent To generate the datasets, a standard normally distributed noise term $\epsilon$ is added to each function,
+scaled by a signal-to-noise ratio of five\footnote{To determine the factor by which the noise is multiplied,
+    the standard deviation of the signal is computed over 100'000 randomly drawn samples of $y$ and divided by
+    the signal-to-noise ratio of five.}. For each test function, we consider two dataset sizes:
+$(n_{train}=1000, n_{val}=250)$ and $(n_{train}=8000, n_{val}=2000)$.
+
+\paragraph{Model Inducers.}
+We consider the following model inducers for our simulation study:
+\begin{itemize}[label=--]
+    \item \textbf{LinReg}: a simple linear regression model as baseline.
+    \item \textbf{GAM\_OT}: a Generalized Additive Model (GAM) with spline terms
+          and tensor splines for first-order interactions, number of splines
+          and penalization optimally tuned.
+    \item \textbf{GAM\_OT}: as above, but with hyperparameters chosen to overfit.
+    \item \textbf{SVM\_OT}: a Support Vector Machine (SVM) with optimally tuned hyperparameters.
+    \item \textbf{SVM\_OF}: a Support Vector Machine (SVM) with hyperparameters chosen to overfit.
+    \item \textbf{XGBoost\_OT}: an XGBoost model with optimally tuned hyperparameters.
+    \item \textbf{XGBoost\_OF}: an XGBoost model with hyperparameters chosen to overfit.
+\end{itemize}
+
+\noindent Hyperparameters are pre-selected, i.e.\ carefully hand-picked for overfitting scenarios
+and tuned on a different sample for optimal tuning. Details on the hyperparameters and the
+performance of the models can be found in \textsc{Appendix~\ref{app:model-hps-perf}}.
+
 \section{Results}\label{sec:results}
 
-\section{Discussion}\label{sec:discussion}
+\section{Conclusions}\label{sec:conclusion}
 
-\section{Conclusion}\label{sec:conclusion}
+\subsection{Summary \& Discussion}
 
-% \begin{table}[t]
-% \caption{Table captions should be placed above the
-% tables.}\label{tab1}
-% \begin{tabular}{lll}
-% \toprule
-% Heading level &  Example & Font size and style\\
-% \midrule
-% Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
-% 1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
-% 2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
-% 3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
-% 4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
-% \bottomrule
-% \end{tabular}
-% \end{table}
+\subsection{Limitations \& Future Work}
 
@@
 
+\section{Model Hyperparameters and Performance}\label{app:model-hps-perf}
+In this appendix, we give details on the models used in the simulation study.
+Specifically, we provide details on the hyperparameter selection and the
+finally used hyperparameters, as well as the performance of the models.
+
+\subsection{Hyperparameters}
+
+An overview of all hyperparameters used for the different models can be found
+in \textbf{Table~\ref{tab:model-hyperparameters}}. Note that the linear
+regression is excluded as we do not have any hyperparameters to tune.
+
+Hyperparameters for the overfitting models (OF) were carefully hand-picked to
+achieve strong performance on the training data while performing relatively
+poorly on the validation data.
+
+The optimal hyperparameters (OT) were chosen by tuning the models on a separate
+data sample. Training and validation data were sampled independently from the
+correspondings DGPs, with a training size of $n_{train}$ and a validation size
+of 10000 to get a reliable performance estimate. While this scenario is
+unrealistic in practice, it allows as to consider hyperparameters as
+``pre-selected'' and avoid costly nested resampling strategies for the
+simulation study. Each model was tuned for 200 trials using a
+\textit{Tree-structured Parzen Estimator (TPE)}~\cite{bergstra_algorithms_2011}
+using validation MSE as objective to minimize (equivalent to maximizing
+R2-score).
+
+\begin{table}
+    \scriptsize
+    \begin{tabularx}{\textwidth}{>{\RaggedRight\arraybackslash}m{1.6cm}p{1.225cm}p{2.05cm}>{\RaggedRight\arraybackslash}X}
+        \toprule
+        \textbf{Dataset}                        & \textbf{n\_train}                 & \textbf{Model}       & \textbf{Hyperparameters}                                                                                                                                                                          \\
+        \midrule
+        \multirow[t]{14}{=}{\textbf{Simple                                                                                                                                                                                                                                                                     \\Normal\\Correlated}} & \multirow[t]{7}{*}{\textbf{1000}} & \textbf{GAM\_OF} & n\_bases: 50; lam: 0.0005;  \\
+        \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 20; lam: 15.6807;                                                                                                                                                                       \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 800; gamma: 10;                                                                                                                                                                                \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 917.9061; gamma: 0.0030;                                                                                                                                                                       \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1200; max\_depth: 16; learning\_rate: 0.35; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                           \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OT} & n\_estimators: 1640; max\_depth: 5; learning\_rate: 0.0062; subsample: 0.5601; min\_child\_weight: 1.6999; colsample\_bytree: 0.7632; colsample\_bylevel: 0.6944; lambda: 0.0156; alpha: 0.0660;  \\
+        \cline{2-4}
+        \textbf{}                               & \multirow[t]{7}{*}{\textbf{8000}} & \textbf{GAM\_OF}     & n\_bases: 64; lam: 1e-05;                                                                                                                                                                         \\
+        \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 5; lam: 0.0010;                                                                                                                                                                         \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 1000; gamma: 10;                                                                                                                                                                               \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 864.4724; gamma: 0.0085;                                                                                                                                                                       \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1500; max\_depth: 18; learning\_rate: 0.4000; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                         \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OT} & n\_estimators: 2586; max\_depth: 5; learning\_rate: 0.0044; subsample: 0.9484; min\_child\_weight: 1.4257; colsample\_bytree: 0.8471; colsample\_bylevel: 0.8672; lambda: 5.1002; alpha: 0.0026;  \\
+        \cline{1-4}
+        \multirow[t]{14}{=}{\textbf{Friedman1}} & \multirow[t]{7}{*}{\textbf{1000}} & \textbf{GAM\_OF}     & n\_bases: 50; lam: 0.0001;                                                                                                                                                                        \\
+        \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 21; lam: 0.0402;                                                                                                                                                                        \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 1000; gamma: 15;                                                                                                                                                                               \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 917.1949; gamma: 0.2102;                                                                                                                                                                       \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1000; max\_depth: 14; learning\_rate: 0.3; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                            \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OT} & n\_estimators: 2621; max\_depth: 8; learning\_rate: 0.0335; subsample: 0.6192; min\_child\_weight: 5.4066; colsample\_bytree: 0.7651; colsample\_bylevel: 0.5224; lambda: 11.6021; alpha: 4.5342; \\
+        \cline{2-4}
+        \textbf{}                               & \multirow[t]{7}{*}{\textbf{8000}} & \textbf{GAM\_OF}     & n\_bases: 80; lam: 1e-08;                                                                                                                                                                         \\
+        \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 22; lam: 0.0657;                                                                                                                                                                        \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 1000; gamma: 18;                                                                                                                                                                               \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 901.1903; gamma: 0.2602;                                                                                                                                                                       \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1200; max\_depth: 14; learning\_rate: 0.3; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                            \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OT} & n\_estimators: 3691; max\_depth: 5; learning\_rate: 0.0070; subsample: 0.6643; min\_child\_weight: 1.4075; colsample\_bytree: 0.8403; colsample\_bylevel: 0.8186; lambda: 0.0399; alpha: 5.0734;  \\
+        \cline{1-4}
+        \multirow[t]{14}{=}{\textbf{Feynman                                                                                                                                                                                                                                                                    \\I.29.16}} & \multirow[t]{7}{*}{\textbf{1000}} & \textbf{GAM\_OF} & n\_bases: 50; lam: 0.0001;  \\
+        \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 31; lam: 0.3260;                                                                                                                                                                        \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 200; gamma: 8;                                                                                                                                                                                 \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 11.4317; gamma: 0.1394;                                                                                                                                                                        \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1000; max\_depth: 14; learning\_rate: 0.3; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                            \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OT} & n\_estimators: 4246; max\_depth: 9; learning\_rate: 0.0327; subsample: 0.8004; min\_child\_weight: 6.3792; colsample\_bytree: 0.8420; colsample\_bylevel: 0.8357; lambda: 14.1131; alpha: 6.0556; \\
+        \cline{2-4}
+        \textbf{}                               & \multirow[t]{7}{*}{\textbf{8000}} & \textbf{GAM\_OF}     & n\_bases: 64; lam: 5e-07;                                                                                                                                                                         \\
+        \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 32; lam: 0.4500;                                                                                                                                                                        \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 400; gamma: 10;                                                                                                                                                                                \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 22.5167; gamma: 0.1114;                                                                                                                                                                        \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1000; max\_depth: 14; learning\_rate: 0.3; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                            \\
+        \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OT} & n\_estimators: 3962; max\_depth: 7; learning\_rate: 0.0372; subsample: 0.9351; min\_child\_weight: 4.0962; colsample\_bytree: 0.8640; colsample\_bylevel: 0.7728; lambda: 29.2036; alpha: 5.1631; \\
+        \bottomrule
+    \end{tabularx}
+    \caption{Hyperparameters for the models used in the simulation study}
+    \label{tab:model-hyperparameters}
+\end{table}
+
+\subsection{Model Performance Evaluation}
+
+To ensure that the models perform as expected, we evaluate their performance
+both on the training data and a holdout test data with the latter consisting
+of 10000 samples to get a reliable performance estimate. The performances
+evaluated over the 30 repetitions are aggregated in \textbf{Fig.~\ref{fig:model-performance}}
+showing the R2-scores. As intended, the overfitting models perform better on
+the training data than the optimally tuned models, while being outperformed on
+the holdout test data and also exhibiting higher variance in their generalization
+performance. Note that the linear regression model serves only as a baseline. The
+overfitted SVM (SVM\_OF) show substantially worse performances on test data compared
+to the other models. Therefore, we excluded it from further analysis.
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{img/model_performances.png}
+    \caption{Performance of the models on training (left) and holdout test data (right), each boxplot aggregates 30 repetitions}
+    \label{fig:model-performance}
+\end{figure}
+
 \end{document}

### 2025-02-01 current_research_feature_effects/feature_effects.py
Add ablation study script for variance decomposition

@@ -1,4 +1,4 @@
-from typing_extensions import List, Dict, Callable, Tuple, Literal, Union
+from typing_extensions import List, Dict, Callable, Literal, Union
 from copy import deepcopy
 from dataclasses import dataclass
 import pandas as pd
@@ -230,43 +230,43 @@ def _accumulated_local_effects(
     return res_df
 
 
-def get_modified_grids(
-    base_grids: List[np.ndarray], Xs: List[np.ndarray], feature_names: List[str]
-) -> Tuple[np.ndarray]:
-    """
-    Get modified grids based on the minimum and maximum values of the datasets.
+# def get_modified_grids(
+#     base_grids: List[np.ndarray], Xs: List[np.ndarray], feature_names: List[str]
+# ) -> Tuple[np.ndarray]:
+#     """
+#     Get modified grids based on the minimum and maximum values of the datasets.
 
-    Parameters
-    ----------
-    base_grids : List[np.ndarray]
-        List of base grids to be modified.
-    Xs : List[np.ndarray]
-        List of datasets to compute the minimum and maximum values from.
-    feature_names : List[str]
-        Names of features to compute the minimum and maximum values from.
+#     Parameters
+#     ----------
+#     base_grids : List[np.ndarray]
+#         List of base grids to be modified.
+#     Xs : List[np.ndarray]
+#         List of datasets to compute the minimum and maximum values from.
+#     feature_names : List[str]
+#         Names of features to compute the minimum and maximum values from.
 
-    Returns
-    -------
-    Tuple[np.ndarray]
-        Tuple of modified grids based on the minimum and maximum values of the datasets,
-        length of the tuple is equal to the number of datasets.
-    """
-    mins = np.array([[np.min(X[:, i]) for i in range(X.shape[1])] for X in Xs]).T
-    maxs = np.array([[np.max(X[:, i]) for i in range(X.shape[1])] for X in Xs]).T
+#     Returns
+#     -------
+#     Tuple[np.ndarray]
+#         Tuple of modified grids based on the minimum and maximum values of the datasets,
+#         length of the tuple is equal to the number of datasets.
+#     """
+#     mins = np.array([[np.min(X[:, i]) for i in range(X.shape[1])] for X in Xs]).T
+#     maxs = np.array([[np.max(X[:, i]) for i in range(X.shape[1])] for X in Xs]).T
 
-    common_mins = mins.max(axis=1)
[…]

### 2025-02-01 scripts/ablation_study_variance_decomposition.py

@@ -1,0 +1,208 @@
+import argparse
+from configparser import ConfigParser
+from pathlib import Path
+import warnings
+import logging
+import os
+from copy import deepcopy
+from multiprocessing import Pool, cpu_count
+import numpy as np
+from sqlalchemy import create_engine
+from sklearn.model_selection import KFold
+
+from current_research_feature_effects.data_generating.data_generation import generate_data
+from current_research_feature_effects.model_training import initialize_model
+from current_research_feature_effects.utils import (
+    create_parameter_space,
+    create_and_set_sim_dir,
+    setup_logger,
+    configure_worker_logger,
+    save_fe_aggregated_results,
+    save_fe_results,
+    SimulationParameter,
+)
+from current_research_feature_effects.feature_effects import (
+    compute_pdps,
+    compute_ales,
+    compute_cv_feature_effect,
+    compute_estimator_mc_variances,
+)
+
+
+def simulate(
+    params: SimulationParameter,
+):
+    np.random.seed(42)
+    logging.info(f"Starting simulation with parameters: {params.groundtruth}, {params.model_name}, {params.n_train}")
+
+    # create directories
+    os.makedirs(str(params.groundtruth), exist_ok=True)
+    os.makedirs(Path(str(params.groundtruth)) / "results", exist_ok=True)
+
+    # create databases for results
+    engine_effects_results = create_engine(
+        f"sqlite:///{str(params.groundtruth)}{params.config.get('storage', 'effects_results')}"
+    )
+
+    feature_names = params.groundtruth.feature_names
+    grid_points = params.config.getint("feature_effects", "grid_points")
+    quantiles = np.linspace(0.0001, 0.9999, grid_points, endpoint=True)
+    grid_values = [params.groundtruth.get_theoretical_quantiles(feature, quantiles) for feature in feature_names]
+    center_curves = params.config["feature_effects"].getboolean("centered")
+    remove_first_last = params.config["feature_effects"].getboolean("remove_first_last")
+    k_cv = params.config.getint("simulation_metadata", "k_cv")
+
+    pdps = {}
+    ales = {}
+
+    K = params.config.getint("simulation_metadata", "n_datasets")
+    M = params.config.getint("simulation_params", "n_sim")
+    for sim_no in range(M):
+        logging.info(
+            f"Starting simulation {sim_no+1}/{M} " + f"for {params.groundtruth} {params.model_name} {params.n_train}."
+        )
[…]

### 2025-02-02 current_research_feature_effects/feature_effects.py
Fix error in ablation study on variance decomposition

@@ -556,23 +556,19 @@ def compute_feature_effect_metrics(estimates: List[FeatureEffect], groundtruth: 
     return metrics
 
 
-def compute_estimator_mc_variances(estimates: List[FeatureEffect], M: int, K: int) -> FeatureEffect:
+def compute_variance(estimates: List[FeatureEffect]) -> FeatureEffect:
     """
-    Compute Monte Carlo variances of feature effect estimates.
+    Compute variance of feature effect estimates.
 
     Parameters
     ----------
     estimates : List[FeatureEffect]
         List of feature effect estimates.
-    M : int
-        Number of model refits.
-    K : int
-        Number of Monte Carlo simulations.
 
     Returns
     -------
     FeatureEffect
-        Object containing the computed Monte Carlo variances.
+        Feature effect object containing the computed variance.
     """
     mean_estimate = sum(estimates) / len(estimates)
-    return sum((estimate - mean_estimate) ** 2 for estimate in estimates) / (M * (K - 1))
+    return sum((estimate - mean_estimate) ** 2 for estimate in estimates) / (len(estimates) - 1)

### 2025-02-02 scripts/ablation_study_variance_decomposition.py

@@ -25,7 +25,7 @@ from current_research_feature_effects.feature_effects import (
     compute_pdps,
     compute_ales,
     compute_cv_feature_effect,
-    compute_estimator_mc_variances,
+    compute_variance,
 )
 
 
@@ -52,8 +52,8 @@ def simulate(
     remove_first_last = params.config["feature_effects"].getboolean("remove_first_last")
     k_cv = params.config.getint("simulation_metadata", "k_cv")
 
-    pdps = {}
-    ales = {}
+    pdp_variances = {}
+    ale_variances = {}
 
     K = params.config.getint("simulation_metadata", "n_datasets")
     M = params.config.getint("simulation_params", "n_sim")
@@ -100,6 +100,8 @@ def simulate(
             models.append(model_fold)
 
         # compute feature effects on K different MC datasets
+        pdps = {}
+        ales = {}
         for k in range(K):
             # sample new dataset
             X_train, y_train, X_val, y_val = generate_data(
@@ -154,14 +156,26 @@ def simulate(
                 pdps[split] = [pdp] if split not in pdps else pdps[split] + [pdp]
                 ales[split] = [ale] if split not in ales else ales[split] + [ale]
 
-    # compute MC variances
+        # compute variance per dataset
+        for split in ["train", "val", "cv"]:
+            pdp_variances[split] = (
+                [compute_variance(pdps[split])]
+                if split not in pdp_variances
+                else pdp_variances[split] + [compute_variance(pdps[split])]
+            )
+            ale_variances[split] = (
+                [compute_variance(ales[split])]
+                if split not in ale_variances
+                else ale_variances[split] + [compute_variance(ales[split])]
+            )
+
+    # average over simulations
     pdp_metrics = {
-        split: {"MC Variance": compute_estimator_mc_variances(pdps_split, M=M, K=K)}
-        for split, pdps_split in pdps.items()
+        split: {"MC Variance": sum(pdp_variances[split]) / len(pdp_variances[split])} for split in pdp_variances.keys()
     }
+
     ale_metrics = {
-        split: {"MC Variance": compute_estimator_mc_variances(ales_split, M=M, K=K)}
-        for split, ales_split in ales.items()
+        split: {"MC Variance": sum(ale_variances[split]) / len(ale_variances[split])} for split in ale_variances.keys()
     }
 
     # save metrics

### 2025-02-03 paper/paper.tex

@@
     \widehat{\text{MSE}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = \frac{1}{M} \sum_{m=1}^{M} {(PDP_{f,S}(x_S) - \widehat{PDP}_{\hat f^{(m)},S}(x_S))}^2
+    \label{eq:mse-estimator}
 \end{equation}
@@
     \widehat{\text{Bias}}(x_S; PDP_{f,S}, \widehat{PDP}_{\hat f,S}) = (PDP_{f,S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat{f}^{(m)},S}(x_S))
+    \label{eq:bias-estimator}
 \end{equation}
@@
     \frac{1}{M-1}\sum_{m=1}^M\left(\widehat{PDP}_{\hat f^{(m)},S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat f^{(m)},S}(x_S)\right)^2
+    \label{eq:variance-estimator}
 \end{equation}
@@
 
+To address the research question of which data to choose to estimate feature
+effects, we conduct a comprehensive simulation study. For different models,
+datasets, and dataset sizes, we estimate the feature effects PDP and ALE on
+training data, validation data, and in a cross-validated manner. We then
+compute the feature effect error as MSE with respect to the ground truth
+feature effects and decompose it into bias and variance components. By
+comparing these error terms across the different estimation strategies
+(training, validation, cross-validation), we aim to provide nuanced empirical
+insights into the implications of the data choice on feature effect estimation.
+For deeper analysis, we conduct two ablation studies: one decomposing the
+feature effect variance into model variance and Monte Carlo variance, and
+another examining the impact of the dataset size on the Monte Carlo variance.
+In the following, we describe our experimental set-up for the main simulation
+study (\textsc{Section~\ref{sec:experiment-main}}) and the ablation studies
+(\textsc{Section~\ref{sec:experiment-ablation}}).
+
+\subsection{Main Experiment}\label{sec:experiment-main}
+
 \paragraph{Datasets.}
@@
 
-\paragraph{Model Inducers.}
-We consider the following model inducers for our simulation study:
+\paragraph{Models.}
+We consider a set of seven complementary learners, spanning different modeling
+paradigms:
+
 \begin{itemize}[label=--]
-    \item \textbf{LinReg}: a simple linear regression model as baseline.
+    \item \textbf{LinReg}: a simple linear regression model as a baseline.
     \item \textbf{GAM\_OT}: a Generalized Additive Model (GAM) with spline terms
@@
           and penalization optimally tuned.
-    \item \textbf{GAM\_OT}: as above, but with hyperparameters chosen to overfit.
+    \item \textbf{GAM\_OF}: as above, but with hyperparameters chosen to overfit.
     \item \textbf{SVM\_OT}: a Support Vector Machine (SVM) with optimally tuned hyperparameters.
@@
 \noindent Hyperparameters are pre-selected, i.e.\ carefully hand-picked for overfitting scenarios
-and tuned on a different sample for optimal tuning. Details on the hyperparameters and the
-performance of the models can be found in \textsc{Appendix~\ref{app:model-hps-perf}}.
+and tuned on a different dataset for optimal tuning. Details on the hyperparameters and performances
+of the models can be found in \textsc{Appendix~\ref{app:model-hps-perf}}.
+
+\paragraph{Feature Effect Estimation.}
+For the trained models, we estimate the feature effects 
+$\widehat{PDP}_{\hat f,S}$ and $\widehat{ALE}_{\hat f,S}$ per feature using
+equations~(\ref{eq:pdp-estimate}) and (\ref{eq:ale-estimate-uncentered}). For
+Monte Carlo integration, we use for each the full training and validation set
+as well as a cross-validation strategy. In the latter case, we use a 5-fold
+cross-validation on the combined samples of training and validation set, in
+each fold fit the model on four folds, and estimate the feature effects on the
+remaining fold. We then average the feature effects over all folds. We also
+estimate the ground truth feature effects $\widehat{PDP}_{f,S}$ and
+$\widehat{ALE}_{f,S}$ since the complexity of the Feynman equation makes
+analytical computation infeasible. For the Monte Carlo integration, we use
+10000 samples additionally sampled from the DGP, which introduces only a
+negligible additional variance term as we show in (\dots) We use 100 grid
+points, defined by the quantiles of the theoretical distribution of $X_S$ to
+enable a comparison across feature effects. Additionally, we center the curves
+after estimation to have a mean effect of $0$ w.r.t.\ the grid points, and we
+omit the first and the last grid point after estimation to avoid boundary
+effects especially occuring in ALE plots.
+
+\paragraph{Feature Effect Errors.} To quantify the error of the estimated feature effects, we compute the MSE of
+the estimated model feature effects with respect to the estimated ground truth
+features effects, as well as their bias and variance. For that, we use the
+estimators defined in equations~(\ref{eq:mse-estimator}) to
+(\ref{eq:variance-estimator}), using $\widehat{PDP}_{f,S}$ instead of
+$PDP_{f,S}$ as ground truth (analogously for ALE). We repeat each
+dataset-size-model combination $M=30$ times to estimate the error terms, where
+each repetition involves drawing a new training and validation set from the
+DGP, refitting the models, and estimating the feature effects. To aggregate the
+errors into a single measure per feature effect, we average the errors over the
+marginal distribution of $X_S$ by averaging over the grid points.
+
+\subsection{Ablation Experiments}\label{sec:experiment-ablation}
+\dots
 
@@
 To ensure that the models perform as expected, we evaluate their performance
-both on the training data and a holdout test data with the latter consisting
-of 10000 samples to get a reliable performance estimate. The performances
-evaluated over the 30 repetitions are aggregated in \textbf{Fig.~\ref{fig:model-performance}}
-showing the R2-scores. As intended, the overfitting models perform better on
-the training data than the optimally tuned models, while being outperformed on
-the holdout test data and also exhibiting higher variance in their generalization
-performance. Note that the linear regression model serves only as a baseline. The
-overfitted SVM (SVM\_OF) show substantially worse performances on test data compared
-to the other models. Therefore, we excluded it from further analysis.
+both on the training data and a holdout test data with the latter consisting of
+10000 samples to get a reliable performance estimate. The performances
+evaluated over the 30 repetitions are aggregated in
+\textbf{Fig.~\ref{fig:model-performance}} showing the R2-scores. As intended,
+the overfitting models perform better on the training data than the optimally
+tuned models, while being outperformed on the holdout test data and also
+exhibiting higher variance in their generalization performance. Note that the
+linear regression model serves only as a baseline. The overfitted SVM (SVM\_OF)
+show substantially worse performances on test data compared to the other
+models. Therefore, we excluded it from further analysis.
 
@@
     \includegraphics[width=\textwidth]{img/model_performances.png}
-    \caption{Performance of the models on training (left) and holdout test data (right), each boxplot aggregates 30 repetitions}
+    \caption{R2-score of the models on training (left) and holdout test data (right), each boxplot aggregates 30 repetitions}
     \label{fig:model-performance}

### 2025-02-07 paper/paper.tex
paper: add related works and rewrite intro

@@
 making it often impossible to explain their decisions. However, this can be
-problematic in many domains and applications, such as the healthcare, legal, or
-finance sectors, where decisions must be transparent and
+problematic in many critical domains and applications, such as the healthcare,
+legal, or finance sectors, where decisions must be transparent and
 accountable~\cite{adadi_peeking_2018}.
 
-Interpretability is crucial to enhance
+Interpretability in ML is crucial to enhance
 trust~\cite{ribeiro_why_2016,teach_analysis_1981}, address potential
@@
 Act~\cite{euaia2024}. To address these challenges, the field of Explainable AI
-/ Interpretable ML has emerged~\cite{adadi_peeking_2018}. Although there are
-many different methods\footnote{For an overview of Explainable AI methods, see
-    e.g.~\cite{adadi_peeking_2018,kamath_introduction_2021,molnar_interpretable_2022}.},
-we will focus on feature effect methods like \textit{Partial Dependence Plots
-    (PDP)}~\cite{friedman_greedy_2001} and \textit{Accumulated Local Effects
-    (ALE)}~\cite{apley_visualizing_2020}.
+/ Interpretable ML has emerged~\cite{adadi_peeking_2018}. In the last years, a
+wide variety of methods has been developed, including model-specific and
+model-agnostic approaches ranging from local feature attributions to global
+feature importances and effects\footnote{For an overview of Explainable AI
+    methods, see
+    e.g.~\cite{adadi_peeking_2018,kamath_introduction_2021,molnar_interpretable_2022}.}.
 
 Due to the severity of many applications, it is crucial to utilize these
-explainability methods correctly. In general, there are many pitfalls to be
-aware of~\cite{molnar_general_2022}, including whether to compute explanations
-in-sample, i.e.\ on training data, or out-of-sample, which we refer to as
-validation data in the following. In loss-based methods such as
-\textit{Permutation Feature Importance (PFI)
-}\cite{breiman_random_2001,fisher_all_2019}, this choice is crucial and has
-already been studied (e.g., in~\cite{molnar_general_2022}). Nevertheless, other
-explainability methods, such as the \textit{Mean Decrease in Impurity (MDI)}
-(or \textit{Gini Importance}) of Random Forests, or \textit{SHAP}
-values~\cite{loecher_debiasing_2022,loecher_debiasing_2024}, have also been
-found to exhibit biases when computed on training data.
+explainability methods correctly to avoid misleading or incorrect conclusions.
+In general, there are many pitfalls to be aware of~\cite{molnar_general_2022},
+including whether to compute explanations in-sample, i.e.\ on training data, or
+out-of-sample, which we refer to as validation data in the following. Existing
+works have studied the implications of this choice for many methods, including
+studies on \textit{Permutation Feature Importance (PFI) } (e.g.,
+in~\cite{molnar_general_2022}), \textit{Mean Decrease in Impurity (MDI)} of
+Random Forests~\cite{loecher_debiasing_2022}, or \textit{SHAP}
+values~\cite{loecher_debiasing_2024}.
 
-However, to the best of our knowledge, there exist no similar studies for
-feature effect methods like PDP and ALE. Existing works, including the original
-papers of PDP and ALE, rely on training data without further justification
-(e.g.,~\cite{apley_visualizing_2020,friedman_greedy_2001,molnar_interpretable_2022}).
-In contrast, practitioners often advocate for using unseen test/validation
-data\footnote{see, e.g.\ \url{https://github.com/SauceCat/PDPbox/issues/68}
-    (10/27/2024)}, or base their choice on practical constraints like dataset
-size\footnote{see, e.g.
-    \url{https://forums.fast.ai/t/partial-dependence-plot/98465} (10/27/2024)}.
-While the training set is usually larger and might thus lead to less variance
-in the feature effect estimates, a too large dataset can increase computation
-times substantially, particularly for the PDP~\cite{friedman_greedy_2001}. On
-the other hand, although feature effects are not based on generalization error
-like PFI, it is not clear how much they are affected by overfitting or
-distribution shifts between training and validation data.
+However, to the best of our knowledge, there exist no such systematic studies
+for feature effect methods like \textit{Partial Dependence Plots
+    (PDP)}~\cite{friedman_greedy_2001} and \textit{Accumulated Local Effects
+    (ALE)}~\cite{apley_visualizing_2020}. Current literature
+(e.g.,~\cite{apley_visualizing_2020,friedman_greedy_2001,molnar_interpretable_2022})
+predominantly uses training data without explicit justification, while
+practitioners also advocate for using unseen validation data (for details, see
+\textsc{Section~\ref{sec:related-works}}). Factors that may influence this
+choice include potential biases arising from overfitting, the dataset size, and
+computational constraints. Although PDP and ALE are not based on generalization
+error like PFI, it is not studied if and how they are affected by overfitting.
+In addition, larger datasets may improve the feature effect estimates, but also
+increase computation times.\\
 
-In this paper, we aim to answer this largely unaddressed, fundamental
+\noindent In this paper, we aim to answer the largely unaddressed, fundamental
 methodological question of whether to use training or validation data to
-compute feature effects. We perform an empirical simulation study, comparing
-feature effect error and uncertainty for PDP and ALE across training data,
-validation data, and cross-validation scenarios, considering various data
-scenarios and model types. Our main contributions in this paper are as follows.
+compute feature effects. We perform an empirical simulation study, estimating
+PDP and ALE on training data, validation data, and in a cross-validated manner
+for different models and dataset. We decompose the error of PDP and ALE and
+compare the error components to understand the implications of the data choice.
+Our main contributions can be summarized as follows:
+
 \begin{enumerate}
@@
           comprehensive simulation study, considering feature effect error, bias, and
-          variance across different models and data scenarios.
-    \item We define methods to quantify feature effect error and uncertainty in a
-          theoretical framework, building upon previous work in this area.
-    \item We provide an overview of commonly used test functions for simulation studies
-          in Interpretable ML, including applications and purposes.
+          variance across different models and datasets.
+    \item We extend previous work to provide a theoretical framework for decomposing
+          feature effect errors and estimating the corresponding components.
+    \item We provide an overview of commonly used test functions for simulation studies,
+          including applications and purposes, as guidance for researchers in
+          Interpretable ML.
 \end{enumerate}
 
-These contributions have several important implications: Our empirically
+\noindent These contributions have several important implications: Firstly, our empirically
 grounded recommendations enable practitioners and researchers to make informed
-decisions about which dataset to select for feature effect computation, helping
-to understand potential implications of their choices. Additionally, our
-framework for evaluating feature effects and our systematic collection of test
-functions provide a foundation for future research in Interpretable ML, with
-the latter specifically facilitating test function choice for simulation
-studies.
+decisions about which data to choose for feature effect computation, helping to
+understand potential implications of their choices. Additionally, our framework
+for evaluating feature effects and systematic collection of test functions
+provide a foundation for future research in Interpretable ML, with the latter
+specifically facilitating test function choice for simulation studies.\\
 
-The remainder of this paper is structured as follows. In
-\textsc{Section~\ref{sec:related-works}}, we introduce the considered feature
-effect methods PDP and ALE, as well as related works on feature effect error
-and uncertainty, and give an overview of common test functions for simulation
-studies. In \textsc{Section~\ref{sec:quantify-fe-u}}, we extend existing works
-and give our definitions of feature effect errors and uncertainty. We then
-describe the methodology and set-up of our simulation studies in
+\noindent The remainder of this paper is structured as follows. In
+\textsc{Section~\ref{sec:related-works}}, we provide an overview of related
+works on feature effects and test functions for simulation studies. In
+\textsc{Section~\ref{sec:background}}, introduce the considered feature
+effect methods PDP and ALE, and demonstrate how to decompose the error of these
+feature effects, providing corresponding definitions and estimators.
+We then describe the methodology and set-up of our simulation studies in
 \textsc{Section~\ref{sec:methodology-set-up}}, present the results in
-\textsc{Section~\ref{sec:results}}, and discuss their implications and
-limitations in \textsc{Section~\ref{sec:discussion}}. In
-\textsc{Section~\ref{sec:conclusion}}, we briefly conclude our work.
+\textsc{Section~\ref{sec:results}}, and conclude our work with a discussion 
+of their implications and limitations in \textsc{Section~\ref{sec:conclusion}}.
 
-\section{Background \& Related Work}\label{sec:related-works}
+\section{Related Works}\label{sec:related-works}
+
+\subsection{Feature Effects}
+
+\paragraph{Feature Effect Methods.}
+Current literature offers various methods for analyzing feature effects in ML.
+One of the most popular methods is the \textit{Partial Dependence Plot
+    (PDP)}~\cite{friedman_greedy_2001}, which describes the marginal effect of one
+or two features on the model prediction. The PDP assumes that the features of
+interest are independent of the remaining features. When this assumption is
+violated, the method may produce unrealistic data points outside the underlying
+joint distribution of the data. This extrapolation issue can cause misleading
+interpretations~\cite{molnar_interpretable_2022,molnar_general_2022}.
+
+\textit{M-Plots (Marginal Plots)} try to address this issue by considering the
+conditional distribution of the remaining features given the feature of interest
+but instead suffer from the omitted variable
+bias~\cite{apley_visualizing_2020,friedman_greedy_2001}.
+
+The \textit{Accumulated Local Effects (ALE)} plot~\cite{apley_visualizing_2020}
+addresses both issues by accumulating local differences in model predictions
+along the feature of interest, computing conditional expectations within small
+intervals rather than across the entire feature range.
+
+Alternative approaches include \textit{functional ANOVA (fANOVA)}, which
+decomposes feature effects into main and interaction
+effects~\cite{hooker_discovering_2004}. Recent extensions have further
+addressed limitations of these classical methods. For example, \textit{Robust
+    and Heterogeneity-aware ALE (RHALE)}~\cite{gkolemis_rhale_2023} adds
+heterogeneity quantification to ALE and improving robustness, or
+\textit{Accumulated Total Derivative Effect (ATDEV)} plots can be decomposed
+into ALE and \textit{Accumulated Cross Effects (ACE)}~\cite{liu_model_2018}.
+
+Beyond these global feature effect methods, there are regional effect plots
+such as \textit{REPID}~\cite{herbinger_repid_2022}, and local methods like
+\textit{ICE (Individual Conditional Expectation)}
+curves~\cite{goldstein_peeking_2015} or SHAP dependence
+plots~\cite{lundberg_local_2020}.
+
+In this paper, we focus on PDP and ALE and refer to them when speaking of
+feature effects.
+
+\paragraph{Feature Effect Error and Uncertainty.}
+Multiple works have approached measuring uncertainty in feature effects. For
+probabilistic ML models, Moosbauer et al.\cite{moosbauer_explaining_2021}
+derived model-specific confidence bands for PDPs, while applied studies have
+proposed bootstrap-based
+approaches\cite{esselman_landscape_2015,grange_using_2019}. While the former
+are not model-agnostic, the latter often capture only the variance introduced
+by Monte Carlo approximation. Nonetheless, the model variance is another source
+of uncertainty in PDPs and ALEs, and accounting for it requires multiple model
+fits~\cite{apley_visualizing_2020,molnar_general_2022}.
+
+Molnar et al.~\cite{molnar_relating_2023} give formalizations of PDP (and PFI)
+as statistical estimators of ground truth estimands, demonstrate the
+decomposition of its mean squared error (MSE) into model bias, model variance,
+and Monte Carlo variance, and provide estimators for both Monte Carlo and
+overall variance. Our work builds upon and extends this idea.
+
+\paragraph{Data Choice for Feature Effect Estimation.}
+For many interpretability methods, the choice whether to compute explanations
+in-sample (on training data) or out-of-sample (on validation data) can
+significantly impact interpretations: For loss-based methods such as
+\textit{Permutation Feature Importance (PFI)
+}\cite{breiman_random_2001,fisher_all_2019}, this choice is crucial and has
+been extensively studied (e.g., in~\cite{molnar_general_2022}). Similar
+concerns have been identified for other explainability methods: both the
+\textit{Mean Decrease in Impurity (MDI)} of Random Forests and \textit{SHAP}
+values have been shown to potentially exhibit biases when computed on training
+data~\cite{loecher_debiasing_2022,loecher_debiasing_2024}.
+
+For feature effect methods, however, the implications of this choice remain
+largely unexplored. The original works introducing
+PDP~\cite{friedman_greedy_2001} and ALE~\cite{apley_visualizing_2020}, as well
+as general introductory literature on Interpretable
+ML~\cite{molnar_interpretable_2022}, predominantly use training data without
+explicit justification. In contrast, practitioners often advocate for using
+holdout data or base their choice on practical constraints such as dataset
+size\footnote{for examples, see
+    \url{https://github.com/SauceCat/PDPbox/issues/68} and
+    \url{https://forums.fast.ai/t/partial-dependence-plot/98465} (both accessed
+    10/27/2024)}. A too large dataset can increase computation times substantially,
+particularly for PDPs~\cite{friedman_greedy_2001}. Moreover, Molnar et
+al.~\cite{molnar_relating_2023} estimate the PDP on holdout data when aiming to
+quantify the variance of feature effect estimates. Nonetheless, a systematic
+study on the implications of the data choice for feature effect estimation is
+missing so far.
+
+\subsection{Test Functions for Simulation Studies}\label{sec:test-functions}
+
+Test functions play a crucial role in research, e.g.\ for evaluating different
+methodological approaches, or when conducting simulation studies. In the
+following, we synthesize commonly used test functions across various domains,
+providing structured guidance for researchers --- particularly those in
+Interpretable ML --- in selecting appropriate test functions for their
+simulation studies. By examining test functions from different fields and their
+purposes of application, we aim to facilitate experimental design decisions
+rather than giving an exhaustive overview.
+
+\paragraph{Test Functions in Optimization.}
+The field of optimization, where test functions are essential to enable the
+assessment and comparison of optimization algorithms, has established a rich
+foundation of test functions. A fundamental approach involves using simple
+mathematical expressions like the sphere function~\cite{more_testing_1981}.
+These basic functions are often combined with more complex ones like Branin or
+Rosenbrock to create comprehensive test suites that incorporate important
+properties such as nonlinearity, non-separability, and
+scalability~\cite{whitley_building_1998}. A notable framework in this domain is
+the Comparing Continuous Optimizer (COCO) platform with its Black Box
+Optimization Benchmark (BBOB), offering a structured approach to testing
+continuous optimization algorithms through artificial test
+functions~\cite{hansen_coco_2016}. These classical test function suites are
+well-established in optimization and may also serve as a basis for
+Interpretable ML researchers. However, the ability of these artifical test
+functions to represent complex real-world behavior is often
+limited~\cite{zaefferer_simulation-based_2017}.
+
+\paragraph{Physics-Inspired Test Functions.}
+
+Physics-derived functions offer a compelling source of real-world test cases,
+with the \textit{Feynman Symbolic Regression Database
+    (FSReD)}~\cite{udrescu_ai_2020} being a prominent example. FSReD comprises 100
+physics equations from the seminal \textit{Feynman's Lectures on Physics
+    (34--36)}, supplemented by 20 more challenging equations from other seminal
+physics texts. These equations span diverse physics domains and involve a
+varying number of variables and various elementary functions such as arithmetic
+operations, trigonometric functions, and exponentials. Tabular datasets are
+generated through random sampling from defined value ranges.
+
+Matsubara~et~al.~\cite{matsubara_rethinking_2024} addressed several limitations
+of the original FSReD. They introduce a three-tiered categorization of problems
+(easy, medium, hard) based on their complexity, incorporate dummy variables to
+simulate irrelevant features, and implement more realistic sampling ranges and
+strategies. Detailed specifications for all formulas, including their sampling
+parameters, are available in their work.
+
+While initially developed for symbolic regression tasks, many Feynman equations
+may serve as suitable test functions for simulation studies in Interpretable
+ML. Their basis in physical principles provides real-world relevance, though
+researchers should carefully select equations that align with their specific
+analytical objectives and complexity requirements.
+
+\paragraph{Test Functions for Interpretable Machine Learning.}
+The Interpretable ML field itself has developed several specialized test
+functions designed to evaluate specific aspects of interpretability methods.
+
+Goldstein et al.~\cite{goldstein_peeking_2015} used several simple test
+functions to demonstrate the behavior of Individual Conditional Expectation
+(ICE) curves. These include a simple additive function to demonstrate the
+absence of interactions, simple interactions to reveal heterogeneity that might
+be obscured by averaging procedures such as PDPs, and a specially designed
+function with an empty quadrant for assessing extrapolation behavior.
+
+Similarly, Liu et al.~\cite{liu_model_2018} focus on simple functions before
+progressing to more complex ones. They begin with basic two-variable scenarios
+--- using additive functions, interaction functions, and combinations thereof
+--- and examine these under both independent and correlated feature conditions
+to compare various feature effect methods. The advantage of these simple test
+functions is that solutions (e.g., feature effects) can also be computed
+analytically, and that they allow for deeper and more fine-grained analysis of
+individual aspects.
+
+A more complex test function suite was proposed by
+Tsang~et~al.~\cite{tsang_detecting_2017}, specifically designed to evaluate the
+detection of variable interactions. Their functions incorporate various types
+of interactions with different orders, strengths, non-linearities, and
+overlaps. While this makes them particularly valuable for interaction
+detection, they are also useful for evaluating other interpretability methods
+in scenarios with complex interactions.
+
+The Friedman functions~\cite{breiman_bagging_1996,friedman_multivariate_1991}
+serve as classical benchmarks applicable across various interpretability tasks.
+These three functions combine linear and non-linear effects with interactions,
+incorporating dummy variables and random noise terms to reflect realistic
+complexity. For detailed specifications, see~\cite{breiman_bagging_1996}.
+
+When choosing test functions for simulation studies in Interpretable ML,
+researchers should consider several criteria, including the specific aspects of
+interpretability being evaluated, the desired complexity level and number of
+variables, the presence of specific challenges such as correlation between
+features or interactions, the need for analytical solutions for validation, as
+well as the relevance to real-world applications in the domain of interest.
+
+\section{Background}\label{sec:background}
 
@@
 $\{(x_S^{(g)}, \widehat{PD}_{\hat f, S}(x_S^{(g)}))\}_{g=1}^G$ can be used~\cite{molnar_relating_2023}.
-Molnar et al.~\cite{molnar_general_2022} recommend using quantile-based over equidistant grids.
-
-The PDP assumes that the features in $S$ are independent of the features in
-$C$. If this is violated, the perturbations may produce unrealistic data points
-outside the underlying joint distribution of the data. This extrapolation issue
-can cause misleading
-interpretations~\cite{molnar_interpretable_2022,molnar_general_2022}.\\
+Molnar et al.~\cite{molnar_general_2022} recommend using quantile-based over equidistant grids.\\
 
@@
 used~\cite{apley_visualizing_2020}.\\
-
-\noindent Other feature effect methods include the \textit{M-Plot (Marginal Plot)},
-which, however, suffers from the omitted variable
-bias~\cite{apley_visualizing_2020,friedman_greedy_2001},
-or \textit{functional ANOVA (fANOVA)}, which decomposes feature effects
-into main and interaction effects~\cite{hooker_discovering_2004}.
-
-Furthermore, methods exist that extend previous ones, such as \textit{Robust
-    and Heterogeneity-aware ALE (RHALE)}~\cite{gkolemis_rhale_2023}, or
-\textit{Accumulated Total Derivative Effect (ATDEV)} plots, which can be
-decomposed into ALE and \textit{Accumulated Cross Effects
-    (ACE)}~\cite{liu_model_2018}.
-
-In addition to these global feature effect methods, there are regional effect
-plots such as \textit{REPID}~\cite{herbinger_repid_2022}, as well as local
-methods, including \textit{ICE (Individual Conditional Expectation)}
-curves~\cite{goldstein_peeking_2015} or SHAP dependence plots
-(e.g.,~\cite{molnar_interpretable_2022}) based on SHAP
-values~\cite{lundberg_unified_2017}.
-
-For the remainder of this paper, we will focus on PDP and ALE as the most
-popular global feature effect methods and refer to them when speaking of
-feature effects.
 
@@
     \widehat{\text{Variance}}_{MC}(x_S; PDP_f, \widehat{PDP}_f) = \frac{1}{K}\sum_{k=1}^K(PDP_f(x_S) - \widehat{PDP}_f^{(k)}(x_S))^2
+    \label{eq:mc-variance-estimator-groundtruth}
 \end{equation}
@@
     \end{split}
+    \label{eq:mc-variance-estimator-model}
 \end{equation}
@@
 ALE as well.
-
-\section{Test Functions for Simulation Studies}\label{sec:test-functions}
-
-Test functions play a crucial role in research, e.g.\ for evaluating and
-comparing different methodological approaches, or when conducting simulation
-studies. This section synthesizes commonly used test functions across various
-domains, providing structured guidance for researchers --- particularly those
-in Interpretable ML --- in selecting appropriate test functions for their
-simulation studies. By examining test functions from different fields and their
-purposes of application, we aim to facilitate experimental design decisions.
-While this overview is not exhaustive, it highlights key test functions and
-their characteristics to help researchers make informed choices for their
-specific research needs.
-
-\paragraph{Test Functions in Optimization.}
-The field of optimization, where test functions are essential to enable the
-assessment and comparison of optimization algorithms, has established a rich
-foundation of test functions. A fundamental approach involves using simple
-mathematical expressions like the sphere function~\cite{more_testing_1981}.
-These basic functions are often combined with more complex ones like Branin or
-Rosenbrock to create comprehensive test suites that incorporate important
-properties such as nonlinearity, non-separability, and
-scalability~\cite{whitley_building_1998}. A notable framework in this domain is
-the Comparing Continuous Optimizer (COCO) platform with its Black Box
-Optimization Benchmark (BBOB), offering a structured approach to testing
-continuous optimization algorithms through artificial test
-functions~\cite{hansen_coco_2016}. These classical test function suites are
-well-established in optimization and may also serve as a basis for
-Interpretable ML researchers. However, it is important to note that the ability
-of these artifical test functions to represent complex real-world behavior is
-limited~\cite{zaefferer_simulation-based_2017}.
-
-\paragraph{Physics-Inspired Test Functions.}
-
-Physics-derived functions offer a compelling source of real-world test cases,
-with the Feynman Symbolic Regression Database (FSReD) being a prominent
-example. FSReD comprises 100 physics equations from the seminal
-\textit{Feynman's Lectures on Physics (34--36)}, supplemented by 20 more
-challenging equations from other seminal physics texts~\cite{udrescu_ai_2020}.
-These equations span diverse physics domains including classical mechanics,
-electromagnetism, and quantum mechanics. They involve between one and nine
-independent variables and incorporate various elementary functions such as
-arithmetic operations, trigonometric functions, and exponentials. From that,
-tabular datasets were generated through random sampling from defined value
-ranges.
-
-Building upon this foundation,
-Matsubara~et~al.~\cite{matsubara_rethinking_2024} addressed several limitations
-of the original FSReD. Their enhanced framework introduces a three-tiered
-categorization of problems (easy, medium, hard) based on their complexity,
-incorporates dummy variables to simulate irrelevant features, and implements
-more realistic sampling ranges and strategies. Detailed specifications for all
-formulas, including their sampling parameters, are available in their work.
-
-While initially developed for symbolic regression tasks, many Feynman equations
-may serve as suitable test functions for simulation studies in Interpretable
-ML. Their basis in physical principles provides real-world relevance, though
-researchers should carefully select equations that align with their specific
-analytical objectives and complexity requirements.
-
-\paragraph{Test Functions for Interpretable Machine Learning.}
-The Interpretable ML field itself has developed several specialized test
-functions designed to evaluate specific aspects of interpretability methods.
-
-Goldstein et al.~\cite{goldstein_peeking_2015} used several simple test
-functions to demonstrate the behavior of Individual Conditional Expectation
-(ICE) curves. These include a simple additive function to demonstrate the
-absence of interactions, simple interactions to reveal heterogeneity that might
-be obscured by averaging procedures such as PDPs, and a specially designed
-function with an empty quadrant for assessing extrapolation behavior.
-
-Similarly, Liu et al.~\cite{liu_model_2018} focus on simple functions before
-progressing to more complex ones. They begin with basic two-variable scenarios
---- using additive functions, interaction functions, and combinations thereof
---- and examine these under both independent and correlated feature conditions
-to compare various feature effect methods. The advantage of these simple test
-functions is that solutions (e.g., feature effects) can also be computed
-analytically, and that they allow for deeper and more fine-grained analysis of
-individual aspects.
-
-A more complex test function suite was proposed by
-Tsang~et~al.~\cite{tsang_detecting_2017}, specifically designed to evaluate the
-detection of variable interactions. Their functions incorporate various types
-of interactions with different orders, strengths, non-linearities, and
-overlaps. While this makes them particularly valuable for interaction
-detection, they are also useful for evaluating other interpretability methods
-in scenarios with complex interactions.
-
-The Friedman functions~\cite{breiman_bagging_1996,friedman_multivariate_1991}
-serve as classical benchmarks applicable across various interpretability tasks.
-These three functions combine linear and non-linear effects with interactions,
-incorporating dummy variables and random noise terms to reflect realistic
-complexity. For detailed specifications, see~\cite{breiman_bagging_1996}.
-
-When choosing test functions for simulation studies in Interpretable ML,
-researchers should consider several criteria, including the specific aspects of
-interpretability being evaluated, the desired complexity level and number of
-variables, the presence of specific challenges such as correlation between
-features or interactions, the need for analytical solutions for validation, as
-well as the relevance to real-world applications in the domain of interest.
 
@@
 
-\paragraph{Feature Effect Estimation.}
-For the trained models, we estimate the feature effects 
-$\widehat{PDP}_{\hat f,S}$ and $\widehat{ALE}_{\hat f,S}$ per feature using
+\paragraph{Feature effect estimation.}
+For the trained models, we estimate the feature effects $\widehat{PDP}_{\hat
+        f,S}$ and $\widehat{ALE}_{\hat f,S}$ per feature using
 equations~(\ref{eq:pdp-estimate}) and (\ref{eq:ale-estimate-uncentered}). For
@@
 
-\paragraph{Feature Effect Errors.} To quantify the error of the estimated feature effects, we compute the MSE of
+\paragraph{Feature effect errors.} To quantify the error of the estimated feature effects, we compute the MSE of
 the estimated model feature effects with respect to the estimated ground truth
@@
 \subsection{Ablation Experiments}\label{sec:experiment-ablation}
-\dots
+
+As described earlier, we conduct two ablation studies for more detailed
+insights into the Monte Carlo variance.
+
+\paragraph{Variance decomposition study.} The variance of the feature effects estimated with the procedure described in
+the main experiment \textsc{Section~\ref{sec:experiment-main}} consists of two
+components: the model variance and the variance due to the Monte Carlo
+integration. To disentangle these two sources of variance, we aim to
+additionally estimate the Monte Carlo variance using the estimator in
+equation~(\ref{eq:mc-variance-estimator-model}). This allows us to substract it
+from the total variance estimated in the main experiment and obtain the model
+variance, giving more nuanced insights into the implications of the data choice
+on feature effect estimation. To obtain this estimate, we maintain the set-up
+of the main experiment but for each trained model $\hat f^{(m)}$ (plus five
+models from CV $\hat f^{(m,1)} - \hat f^{(m,5)}$), we draw $K=30$ new training
+and validation sets from the DGP and estimate the feature effects on these sets
+plus in a cross-validated manner with the already fitted models. This allows us
+to estimate the Monte Carlo variance as in
+equation~(\ref{eq:mc-variance-estimator-model}). Since this study is
+computationally more expensive, we do this only for XGBoost (OT and OF) with
+the same datasets and sizes as in the main experiment.
+
+\paragraph{Impact of dataset size on Monte Carlo variance.} To investigate the impact of the dataset size on the Monte Carlo variance, we
+estimate the Monte Carlo variance between the analytical ground truth feature
+effects and the estimated ground truth feature effects on different dataset
+sizes. As estimator for the Monte Carlo variance, we use
+equation~(\ref{eq:mc-variance-estimator-groundtruth}). With this approach, we
+entirely omit the model and thus eliminate other error sources such as model
+bias and variance (as with a perfect model fit). To estimate the ground truth
+feature effects, we use 50 different dataset sizes ranging from $10^1$ to
+$10^6$ on a logarithmic scale. Note that we only consider the
+SimpleNormalCorrelated and Friedman1 datasets for this study as an analytical
+computation of the FeynmanI.29.16 feature effects is infeasible. For other
+details, we maintain the set-up of the main experiment.
 
@@
 simulation study. Each model was tuned for 200 trials using a
-\textit{Tree-structured Parzen Estimator (TPE)}~\cite{bergstra_algorithms_2011}
+\textit{Tree-structured Parzen Estimator (TPE)}~\cite{bergstra2011algorithms}
 using validation MSE as objective to minimize (equivalent to maximizing

### 2025-02-08 paper/paper.tex
paper: add main experiment result plots and refine bibliography

@@
 \usepackage{booktabs} 
-\usepackage{svg}
+\usepackage{subcaption}
 \usepackage[misc]{ifsym}
@@
 
+\subsection{Main Experiment}
+
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}[b]{0.49\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_GAM_OF.png}
+        \caption{GAM\_OF}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_GAM_OT.png}
+        \caption{GAM\_OT}
+    \end{subfigure}
+    \\[10pt]
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_XGBoost_OF.png}
+        \caption{XGBoost\_OF}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_XGBoost_OT.png}
+        \caption{XGBoost\_OT}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_SVM_OT.png}
+        \caption{SVM\_OT}
+    \end{subfigure}
+    \caption{Main experiment results for PDP on SimpleNormalCorrelated dataset. \dots}
+    \label{fig:pdp-results-snc}
+ \end{figure}
+
+ \begin{figure}[htbp]
+    \centering
+    \begin{subfigure}[b]{0.49\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_GAM_OF.png}
+        \caption{GAM\_OF}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_GAM_OT.png}
+        \caption{GAM\_OT}
+    \end{subfigure}
+    \\[10pt]
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_XGBoost_OF.png}
+        \caption{XGBoost\_OF}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_XGBoost_OT.png}
+        \caption{XGBoost\_OT}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_SVM_OT.png}
+        \caption{SVM\_OT}
+    \end{subfigure}
+    \caption{Main experiment results for ALE on SimpleNormalCorrelated dataset. \dots}
+    \label{fig:ale-results-snc}
+ \end{figure}
+
 \section{Conclusions}\label{sec:conclusion}

### 2025-02-08 current_research_feature_effects/plotting/plots.py
Refine plot function

@@ -1,6 +1,7 @@
 # from configparser import ConfigParser
 # import math
 from typing_extensions import Dict, Literal, List, Tuple, Optional
+from pathlib import Path
 import pandas as pd
 import numpy as np
 import matplotlib.pyplot as plt
@@ -24,20 +25,37 @@ from current_research_feature_effects.plotting.utils import (
 # from current_research_feature_effects.feature_effects import compute_pdps, compute_ales
 
 
-def plot_feature_effect(effect: FeatureEffect) -> plt.Figure:
+def plot_feature_effect(effect: FeatureEffect | List[FeatureEffect], labels: List[str] = None) -> plt.Figure:
     """
     Plot feature effects for a given feature effect object.
 
     Parameters
     ----------
-    effect : FeatureEffect
-        A FeatureEffect object containing the feature effects to plot.
+    effect : FeatureEffect | List[FeatureEffect]
+        A FeatureEffect object or a list of FeatureEffect objects to be plotted.
+    labels : List[str], optional
+        A list of labels for the different FeatureEffect objects (default is None).
 
     Returns
     -------
     plt.Figure
         A Figure object containing the generated plots.
     """
+    if isinstance(effect, list):
+        n_features = len(effect[0].features)
+        fig, axs = plt.subplots(1, n_features, figsize=(5 * n_features, 5))
+        for i, feature in enumerate(effect[0].features):
+            for e in effect:
+                axs[i].plot(
+                    e.features[feature]["grid"], e.features[feature]["effect"], **get_feature_effect_plot_style()
+                )
+            axs[i].set_title(f"{effect[0].effect_type.upper()} of ${feature}$")
+
+        if labels is not None:
+            axs[-1].legend(labels)
+
+        return fig
+
     n_features = len(effect.features)
     fig, axs = plt.subplots(1, n_features, figsize=(5 * n_features, 5))
     for i, feature in enumerate(effect.features):
@@ -121,51 +139,6 @@ def boxplot_model_results(
     return fig
 
 
-# def boxplot_feature_effect_results(
-#     features: List[Literal["x_1", "x_2", "x_3", "x_4", "x_5"]],
-#     df: pd.DataFrame,
-#     effect_type: Literal["PDP", "ALE"],
-# ) -> plt.Figure:
-#     set_style()
-#     fig = plt.figure(
-#         figsize=(min(6 * len(features), 18), math.ceil(len(features) / 3) * 5),
-#         dpi=300,
-#     )
-#     fig.suptitle(
[…]

### 2025-02-08 paper/paper.tex
paper: add main analysis plots

@@
 
+\clearpage
+\section{Additional Main Experiment Results}\label{app:additional-results}
+
+\subsection{SimpleNormalCorrelated Dataset}
+
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_pdp_LinReg.png}
+        \caption{LinReg PDP results on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_pdp_GAM_OF.png}
+        \caption{GAM\_OF PDP results on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_pdp_GAM_OT.png}
+        \caption{GAM\_OT PDP results on SimpleNormalCorrelated}
+    \end{subfigure}
+\end{figure*}
+
+\begin{figure*}[h!]
+    \ContinuedFloat
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_pdp_XGBoost_OF.png}
+        \caption{XGBoost\_OF PDP results on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_pdp_XGBoost_OT.png}
+        \caption{XGBoost\_O PDP results on SimpleNormalCorrelatedT}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_pdp_SVM_OF.png}
+        \caption{SVM\_OF PDP results on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_pdp_SVM_OT.png}
+        \caption{SVM\_OT PDP results on SimpleNormalCorrelated}
+    \end{subfigure}
+    \caption{Main experiment results for PDP on SimpleNormalCorrelated dataset. $x_3$ and $x_4$ are dummy variables.}
+    \label{fig:pdp-results-snc-all}
+\end{figure*}
+
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_ale_LinReg.png}
+        \caption{LinReg ALE results on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_ale_GAM_OF.png}
+        \caption{GAM\_OF ALE results on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_ale_GAM_OT.png}
+        \caption{GAM\_OT ALE results on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_ale_XGBoost_OF.png}
+        \caption{XGBoost\_OF ALE results on SimpleNormalCorrelated}
+    \end{subfigure}
+\end{figure*}
+
+\begin{figure*}[h!]
+    \ContinuedFloat
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_ale_XGBoost_OT.png}
+        \caption{XGBoost\_OT ALE results on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_ale_SVM_OF.png}
+        \caption{SVM\_OF ALE results on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.67\textwidth]{img/SNC-all/feature_effect_errors_ale_SVM_OT.png}
+        \caption{SVM\_OT ALE results on SimpleNormalCorrelated}
+    \end{subfigure}
+    \caption{Main experiment results for ALE on SimpleNormalCorrelated dataset. $x_3$ and $x_4$ are dummy variables.}
+    \label{fig:ale-results-snc-all}
+\end{figure*}
+
+\clearpage
+\subsection{Friedman1 Dataset}
+
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_pdp_LinReg.png}
+        \caption{LinReg PDP results on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_pdp_GAM_OF.png}
+        \caption{GAM\_OF PDP results on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_pdp_GAM_OT.png}
+        \caption{GAM\_OT PDP results on Friedman1}
+    \end{subfigure}
+\end{figure*}
+
+\begin{figure*}[ht]
+    \centering
+    \ContinuedFloat
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_pdp_XGBoost_OF.png}
+        \caption{XGBoost\_OF PDP results on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_pdp_XGBoost_OT.png}
+        \caption{XGBoost\_OT PDP results on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_pdp_SVM_OF.png}
+        \caption{SVM\_OF PDP results on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_pdp_SVM_OT.png}
+        \caption{SVM\_OT PDP results on Friedman1}
+    \end{subfigure}
+    \caption{Main experiment results for PDP on Friedman1 dataset. $x_6$ and $x_7$ are dummy variables.}
+    \label{fig:pdp-results-friedman1}
+\end{figure*}
+
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_ale_LinReg.png}
+        \caption{LinReg ALE results on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_ale_GAM_OF.png}
+        \caption{GAM\_OF ALE results on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_ale_GAM_OT.png}
+        \caption{GAM\_OT ALE results on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_ale_XGBoost_OF.png}
+        \caption{XGBoost\_OF ALE results on Friedman1}
+    \end{subfigure}
+\end{figure*}
+
+\begin{figure*}[ht]
+    \centering
+    \ContinuedFloat
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_ale_XGBoost_OT.png}
+        \caption{XGBoost\_OT ALE results on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_ale_SVM_OF.png}
+        \caption{SVM\_OF ALE results on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/feature_effect_errors_ale_SVM_OT.png}
+        \caption{SVM\_OT ALE results on Friedman1}
+    \end{subfigure}
+    \caption{Main experiment results for ALE on Friedman1 dataset. $x_6$ and $x_7$ are dummy variables.}
+\label{fig:ale-results-friedman1}
+\end{figure*}
+
+\clearpage
+\subsection{Feynman I.29.16 Dataset}
+
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_pdp_LinReg.png}
+        \caption{LinReg PDP results on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_pdp_GAM_OF.png}
+        \caption{GAM\_OF PDP results on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_pdp_GAM_OT.png}
+        \caption{GAM\_OT PDP results on FeynmanI.29.16}
+    \end{subfigure}
+\end{figure*}
+
+\begin{figure*}[ht]
+    \centering
+    \ContinuedFloat
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_pdp_XGBoost_OF.png}
+        \caption{XGBoost\_OF PDP results on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_pdp_XGBoost_OT.png}
+        \caption{XGBoost\_OT PDP results on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_pdp_SVM_OF.png}
+        \caption{SVM\_OF PDP results on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_pdp_SVM_OT.png}
+        \caption{SVM\_OT PDP results on FeynmanI.29.16}
+    \end{subfigure}
+    \caption{Main experiment results for PDP on FeynmanI.29.16 dataset. $d_1$ and $d_2$ are dummy variables.}
+    \label{fig:pdp-results-feynman}
+\end{figure*}
+
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_ale_LinReg.png}
+        \caption{LinReg ALE results on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_ale_GAM_OF.png}
+        \caption{GAM\_OF ALE results on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_ale_GAM_OT.png}
+        \caption{GAM\_OT ALE results on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_ale_XGBoost_OF.png}
+        \caption{XGBoost\_OF ALE results on FeynmanI.29.16}
+    \end{subfigure}
+\end{figure*}
+
+\begin{figure*}[ht]
+    \centering
+    \ContinuedFloat
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_ale_XGBoost_OT.png}
+        \caption{XGBoost\_OT ALE results on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_ale_SVM_OF.png}
+        \caption{SVM\_OF ALE results on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.86\textwidth]{img/FeynmanI.29.16-all/feature_effect_errors_ale_SVM_OT.png}
+        \caption{SVM\_OT ALE results on FeynmanI.29.16}
+    \end{subfigure}
+    \caption{Main experiment results for ALE on FeynmanI.29.16 dataset. $d_1$ and $d_2$ are dummy variables.}
+    \label{fig:ale-results-feynman}
+\end{figure*}
+
 \end{document}

### 2025-02-09 current_research_feature_effects/plotting/plots.py
update model performance analysis plots

@@ -107,9 +107,16 @@ def plot_groundtruth(groundtruth: Groundtruth, grid_points: Dict[str, np.ndarray
 
 
 def boxplot_model_results(
-    metric: Literal["mse", "mae", "r2"], df: pd.DataFrame, ylim: Optional[Tuple[float, float]] = None
+    metric: Literal["mse", "mae", "r2"],
+    df: pd.DataFrame,
+    ylim: Optional[Tuple[float, float]] = None,
+    large_font: bool = False,
 ) -> plt.Figure:
     set_style()
+    if large_font:
+        _set_fontsize("large")
+    else:
+        _set_fontsize("standard")
     fig, ax = plt.subplots(1, 2, figsize=(12, 6), dpi=300, sharey=True)
     fig.suptitle("Model evaluation", fontsize=16, fontweight="bold")
     ax[0].set_title(f"{metric} on train set")
@@ -122,6 +129,7 @@ def boxplot_model_results(
         **get_boxplot_style(),
     )
     ax[0].legend().set_visible(False)
+    ax[0].set_ylabel(metric)
     if ylim is not None:
         ax[0].set_ylim(ylim)
     sns.boxplot(
@@ -198,6 +206,53 @@ def boxplot_model_results(
 #     return fig
 
 
+def _set_fontsize(size: Literal["standard", "large", "xlarge"]) -> None:
+    """
+    Set the font size of the plots.
+
+    Parameters
+    ----------
+    size : {'standard', 'large', 'xlarge'}
+        The font size to set.
+    """
+    if size == "xlarge":
+        plt.rcParams.update(
+            {
+                "font.size": 16,
+                "axes.labelsize": 18,
+                "axes.titlesize": 20,
+                "figure.titlesize": 22,
+                "xtick.labelsize": 16,
+                "ytick.labelsize": 16,
+                "legend.fontsize": 16,
+            }
+        )
+    elif size == "large":
+        plt.rcParams.update(
+            {
+                "font.size": 14,
+                "axes.labelsize": 16,
+                "axes.titlesize": 18,
+                "figure.titlesize": 20,
+                "xtick.labelsize": 14,
+                "ytick.labelsize": 14,
+                "legend.fontsize": 14,
+            }
+        )
+    else:
+        plt.rcParams.update(
+            {
+                "font.size": 11,
+                "axes.labelsize": 12,
+                "axes.titlesize": 12,
+                "figure.titlesize": 16,
[…]

### 2025-02-10 current_research_feature_effects/plotting/plots.py
add plot for further bias variance investigation

@@ -307,6 +307,78 @@ def plot_feature_effect_error_table(
     plt.show()
 
 
+def plot_fe_bias_variance(error_dict, sharey=True, large_font=False) -> plt.Figure:
+    """
+    Plot the bias and variance of the feature effects over the grid points.
+
+    Parameters
+    ----------
+    error_dict : Dict
+        A dictionary containing the feature effect errors.
+    sharey : bool, optional
+        If True, share the y-axis across all subplots (default is True).
+    large_font : bool, optional
+        If True, use larger font sizes (default is False).
+
+    Returns
+    -------
+    plt.Figure
+        A Figure object containing the generated plots.
+    """
+    sns.set_theme(style="ticks")
+    palette = sns.color_palette("Set2")
+
+    if large_font:
+        _set_fontsize("xlarge")
+    else:
+        _set_fontsize("standard")
+
+    n_features = len(list(list(error_dict.values())[0].values())[0].features)
+    fig, axes = plt.subplots(len(error_dict), n_features, figsize=(5 * n_features, 5 * len(error_dict)), sharey=sharey)
+
+    colors = {"Bias^2": palette[1], "Variance": palette[2]}  # Orange  # Purple
+
+    legend_lines = []
+    legend_labels = []
+
+    for i, (split, metrics) in enumerate(error_dict.items()):
+        for metric in ["Bias^2", "Variance"]:
+            for j, feature in enumerate(metrics[metric].features):
+                line = axes[i, j].plot(
+                    metrics[metric].features[feature]["grid"],
+                    metrics[metric].features[feature]["effect"],
+                    linewidth=2,
+                    color=colors[metric],
+                    marker="+",
+                )[0]
+
+                if i == 0 and j == 0:
+                    legend_lines.append(line)
+                    legend_labels.append(metric)
+
+                axes[i, j].set_title(f"{split}: {metrics[metric].effect_type.upper()} of ${feature}$", pad=10)
+                axes[i, j].set_xlabel(f"${feature}$")
+
+                sns.despine(ax=axes[i, j])
+
+    fig.legend(
+        legend_lines,
+        legend_labels,
+        loc="upper center",
+        bbox_to_anchor=(0.5, 1.0),
+        ncol=2,
+        frameon=False,
+        fontsize=18 if large_font else 12,
+    )
+
+    plt.tight_layout()
+    plt.subplots_adjust(top=0.9)
+    fig.set_dpi(200)
[…]

### 2025-02-10 paper/paper.tex
paper: add chapter on main study results and appendix on further main study investigations

@@
           including applications and purposes, as guidance for researchers in
-          Interpretable ML.
+          Interpretable ML.  % chktex 13
 \end{enumerate}
@@
 \paragraph{Feature Effect Methods.}
-Current literature offers various methods for analyzing feature effects in ML.
+Current literature offers various methods for analyzing feature effects in ML. % chktex 13
 One of the most popular methods is the \textit{Partial Dependence Plot
@@
 may serve as suitable test functions for simulation studies in Interpretable
-ML. Their basis in physical principles provides real-world relevance, though
+ML. Their basis in physical principles provides real-world relevance, though  % chktex 13
 researchers should carefully select equations that align with their specific
@@
 To plot this function, a grid of $G$ grid points
-$\{(x_S^{(g)}, \widehat{PD}_{\hat f, S}(x_S^{(g)}))\}_{g=1}^G$ can be used~\cite{molnar_relating_2023}.
+$\{(x_S^{(g)}, \widehat{PD}_{\hat f, S}(x_S^{(g)}))\}_{g=1}^G$ can be used~\cite{molnar_relating_2023}.  % chktex 3
 Molnar et al.~\cite{molnar_general_2022} recommend using quantile-based over equidistant grids.\\
@@
 
-\noindent Here, $\{N_S(k) = (z_{k-1,S}, z_{k,S}]\}_{k=1}^{K}$ partitions the samples
-$\{x^{(i)}_S\}_{i=1}^n$ into $K$ intervals or neighborhoods $N_S(k)$. $n_S(k)$ denotes
+\noindent Here, $\{N_S(k) = (z_{k-1,S}, z_{k,S}]\}_{k=1}^{K}$ partitions the samples  % chktex 3 chktex 9
+$\{x^{(i)}_S\}_{i=1}^n$ into $K$ intervals or neighborhoods $N_S(k)$. $n_S(k)$ denotes  % chktex 3
 the number of observations in the $k$th interval $N_S(k)$, $k_S(x)$ represents the index
@@
 \noindent to have a mean effect of $0$. For the grid that defines the intervals,
-the quantiles of the empirical distribution of $\{x^{(i)}_S\}_{i=1}^n$ can be
+the quantiles of the empirical distribution of $\{x^{(i)}_S\}_{i=1}^n$ can be  % chktex 3
 used~\cite{apley_visualizing_2020}.\\
@@
 As stated by Molnar et al.~\cite{molnar_relating_2023}, their results also
-apply to conditional variants of the PDP such as ALE. We now make this
+apply to conditional variants of the PDP such as ALE. We now make this  % chktex 13
 definition explicit:
@@
 
-    \noindent where $f^S(x_S, x_C) = \frac{\partial f(x_S, x_C)}{\partial x_S}$ and \text{constant} chosen such that the effect has a mean of $0$ w.r.t. the marginal distribution of $X_S$.
+    \noindent where $f^S(x_S, x_C) = \frac{\partial f(x_S, x_C)}{\partial x_S}$ and \text{constant} chosen such that the effect has a mean of $0$ w.r.t.\ the marginal distribution of $X_S$.
 \end{definition}
@@
 difficult to derive the ground truth feature effects analytically, especially
-for ALE. In these cases, we therefore propose to also estimate those effects by
+for ALE. In these cases, we therefore propose to also estimate those effects by  % chktex 13
 Monte Carlo integration, yielding $\widehat{PDP}_{f,S}(x_S)$ and
@@
 \begin{proof}
-    For better readability, we will omit the subscript $S$ as well as the point $x_S$ and use $X=X_{mc}$ in this proof:
-
-    \begin{align*}
-        \text{MSE}(PDP_{f}, \widehat{PDP}_{\hat f})
-         & = \mathbb{E}_F\mathbb{E}_X[(PDP_f - \widehat{PDP}_{\hat f})^2]                                                                                         \\
-         & = \mathbb{E}_F \mathbb{E}_X[PDP_f^2 - 2PDP_f\widehat{PDP}_{\hat f} + \widehat{PDP}_{\hat f}^2]                                                         \\
-         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\mathbb{E}_X[\widehat{PDP}_{\hat f}^2]                                                      \\
-         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}] + \mathbb{E}_F[\mathbb{E}_X[\widehat{PDP}_{\hat f}]^2] \\
-         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}] + \text{Var}_F(\mathbb{E}_X[\widehat{PDP}_{\hat f}])   \\
-         & \quad + \mathbb{E}_F(\mathbb{E}_X[\widehat{PDP}_{\hat f}])^2                                                                                           \\
-         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] +  Var_F[PDP_{\hat f}] + \mathbb{E}_F[PDP_{\hat f}]^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}]  \\
-         & = (PDP_f - \mathbb{E}_F[PDP_{\hat f}])^2 + \text{Var}_F[PDP_{\hat f}] + \mathbb{E}_F[\text{Var}_X(\widehat{PDP}_{\hat f})]
-    \end{align*}
-
-    \noindent At multiple points, we use the fact that $\mathbb{E}_X[\widehat{PDP}_{\hat f}]
-        = PDP_{\hat f}$ (cf.~\cite{molnar_relating_2023}).
+    Using basic expectation and variance rules, as well as applying the fact that
+    $\mathbb{E}_X[\widehat{PDP}_{\hat f}] = PDP_{\hat f}$ (cf.~\cite{molnar_relating_2023})
+    gives the proof. The full proof is provided in
+    \textsc{Appendix~\ref{app:proof-mse-decomposition}}.
 \end{proof}
@@
 multiple models of the same inducer fitted to $M$ different training data
-sets sampled independently from the DGP. We propose the following estimators:
+sets sampled independently from the DGP. We propose the following estimators:  % chktex 13
 \begin{equation}
@@
     \widehat{\text{Variance}}(x_S; \widehat{PDP}_{\hat f,S}) =
-    \frac{1}{M-1}\sum_{m=1}^M\left(\widehat{PDP}_{\hat f^{(m)},S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat f^{(m)},S}(x_S)\right)^2
+    \frac{1}{M-1}\sum_{m=1}^M{\left(\widehat{PDP}_{\hat f^{(m)},S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat f^{(m)},S}(x_S)\right)}^2
     \label{eq:variance-estimator}
@@
 \begin{equation}
-    \widehat{\text{Variance}}_{MC}(x_S; PDP_f, \widehat{PDP}_f) = \frac{1}{K}\sum_{k=1}^K(PDP_f(x_S) - \widehat{PDP}_f^{(k)}(x_S))^2
+    \widehat{\text{Variance}}_{MC}(x_S; PDP_f, \widehat{PDP}_f) = \frac{1}{K}\sum_{k=1}^K{(PDP_f(x_S) - \widehat{PDP}_f^{(k)}(x_S))}^2
     \label{eq:mc-variance-estimator-groundtruth}
@@
         \widehat{\text{Variance}}_{MC}(x_S; \widehat{PDP}_{\hat{f}}) = \\ \frac{1}{M(K-1)}\sum_{m=1}^M \sum_{k=1}^K
-        \left(\widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S) - \frac{1}{K}\sum_{k=1}^K \widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S)\right)^2
+        {\left(\widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S) - \frac{1}{K}\sum_{k=1}^K \widehat{PDP}_{\hat f^{(m)},S}^{(k)}(x_S)\right)}^2
     \end{split}
@@
           \begin{equation}
-              f_2(\boldsymbol{x}) = 10 \sin(\pi x_1 x_2) + 20(x_3 - \frac{1}{2})^2 + 10 x_4 + 5 x_5
+              f_2(\boldsymbol{x}) = 10 \sin(\pi x_1 x_2) + 20{(x_3 - \frac{1}{2})}^2 + 10 x_4 + 5 x_5
           \end{equation}
@@
           describing wave interference, comprising six independent features. Using
-          the refined sampling strategies from Matsubara et al. \cite{matsubara_rethinking_2024},
+          the refined sampling strategies from Matsubara et al.~\cite{matsubara_rethinking_2024},
           two log-uniformly distributed variables on $[0.1, 10]$, two angles uniformly
@@
 \paragraph{Feature effect estimation.}
-For the trained models, we estimate the feature effects $\widehat{PDP}_{\hat
-        f,S}$ and $\widehat{ALE}_{\hat f,S}$ per feature using
+For the trained models, we estimate the feature effects $\widehat{PDP}_{\hat{f},S}$
+and $\widehat{ALE}_{\hat f,S}$ per feature using
 equations~(\ref{eq:pdp-estimate}) and (\ref{eq:ale-estimate-uncentered}). For
@@
 
+We give the results of the main experiment error decomposition for the SimpleNormalCorrelated
+dataset in \textbf{Fig.\@~\ref{fig:pdp-results-snc}} for PDP and \textbf{Fig.\@~\ref{fig:ale-results-snc}}
+for ALE. The full results for all datasets, models, and features can be found in  % chktex 13
+\textsc{Appendix~\ref{app:additional-results}}.
+
+Generally, we observe that the feature effect error is in the most cases highest when
+the effect is estimated on validation data. For ALE, this observation is more pronounced
+than for PDP. This increase in error is mainly due to an increase in variance (see e.g.,  % chktex 13
+\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}} or \textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-of}}).
+For many models, the feature effect error is lowest when estimated in a cross-validated manner.
+This is slightly more pronounced for overfitting models (OF) than for optimally tuned models (OT),
+and is mainly due to a decrease in variance (see e.g., \textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-of}}
+and \textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}}). Since the variance composes of the model variance
+and the Monte Carlo variance, both could be reduced by cross-validation. Monte Carlo variance
+may decrease as we have slightly more samples overall than in the training set. Conversely,
+the model variance may decrease as it is averaged out through the cross-validation procedure,
+which may especially improve overfitting models that have a higher model variance.
+We investigate these hypotheses in an ablation study in \textsc{Section~\ref{sec:results-variance-decomposition}}.
+
 \begin{figure}[htbp]
@@
         \caption{GAM\_OF}
+        \label{fig:pdp-results-snc-gam-of}  % chktex 24
     \end{subfigure}
@@
         \caption{GAM\_OT}
-    \end{subfigure}
-    \\[10pt]
-    \hfill
+        \label{fig:pdp-results-snc-gam-ot}  % chktex 24
+    \end{subfigure}
+    \\[10pt]
+    \vfill
     \begin{subfigure}[b]{0.49\textwidth}
@@
         \caption{XGBoost\_OF}
+        \label{fig:pdp-results-snc-xgboost-of}  % chktex 24
     \end{subfigure}
@@
         \caption{XGBoost\_OT}
+        \label{fig:pdp-results-snc-xgboost-ot}  % chktex 24
     \end{subfigure}
@@
         \caption{SVM\_OT}
-    \end{subfigure}
-    \caption{Main experiment results for PDP on SimpleNormalCorrelated dataset. \dots}
-    \label{fig:pdp-results-snc}
+        \label{fig:pdp-results-snc-svm-ot}  % chktex 24
+    \end{subfigure}
+    \caption{Main experiment results for PDP on SimpleNormalCorrelated dataset.\dots}
+    \label{fig:pdp-results-snc}  % chktex 24
 \end{figure}
+
+Interestingly, for ALE, we observe for some cases an increased bias when estimating the feature effects on
+validation data and via cross-validation compared to an estimation on training data (see e.g.,
+\textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-ot}} and \textbf{Fig.\@~\ref{fig:ale-results-snc-svm-ot}}).
+Note that this is only the case when errors are generally low, and the bias is still small in absolute terms.
+Since this only occurs for ALE and only for the smaller dataset size (1000 training samples, 250 validation samples),
+we hypothesize that this is due to our approach of choosing the intervals based on theoretical quantiles.
+For the smaller validation sets, we may have intervals with few or no samples, leading to a bias in the
+estimation of the effect. This hypothesis is supported by the fact that the increase in bias only occurs
+in regions with low sample density, which we demonstrate in \textsc{Appendix~\ref{app:main-study-further-investigation}}.
+Thus, an open question remains whether choosing the intervals based on the empirical quantiles mitigates this issue.
 
@@
         \caption{GAM\_OF}
+        \label{fig:ale-results-snc-gam-of}  % chktex 24
     \end{subfigure}
@@
         \caption{GAM\_OT}
-    \end{subfigure}
-    \\[10pt]
-    \hfill
+        \label{fig:ale-results-snc-gam-ot}  % chktex 24
+    \end{subfigure}
+    \\[10pt]
+    \vfill
     \begin{subfigure}[b]{0.49\textwidth}
@@
         \caption{XGBoost\_OF}
+        \label{fig:ale-results-snc-xgboost-of}  % chktex 24
     \end{subfigure}
@@
         \caption{XGBoost\_OT}
+        \label{fig:ale-results-snc-xgboost-ot}  % chktex 24
     \end{subfigure}
@@
         \caption{SVM\_OT}
-    \end{subfigure}
-    \caption{Main experiment results for ALE on SimpleNormalCorrelated dataset. \dots}
-    \label{fig:ale-results-snc}
+        \label{fig:ale-results-snc-svm-ot}  % chktex 24
+    \end{subfigure}
+    \caption{Main experiment results for ALE on SimpleNormalCorrelated dataset.\dots}
+    \label{fig:ale-results-snc}  % chktex 24
 \end{figure}
 
+Further results on Friedman1 and Feynman I.29.16, as well as for the dummy features, largely confirm
+the described findings. For details, we refer to \textsc{Appendix~\ref{app:additional-results}}. We only
+observe one case where the error of ALE is substantially larger when estimated on training data
+due to a high bias. This occurs for the XGBoost\_OF model on the Friedman1 dataset, which we further
+look into in \textsc{Appendix~\ref{app:main-study-further-investigation}}.
+
+\subsection{Ablation Study Results: Variance Decomposition}\label{sec:results-variance-decomposition}
+
+\dots
+
+\subsection{Ablation Study Results: Impact of Dataset Size on Monte Carlo Variance}\label{sec:results-dataset-size}
+
+\clearpage
 \section{Conclusions}\label{sec:conclusion}
@@
 \appendix
+\section{Proof: Decomposition of MSE}\label{app:proof-mse-decomposition}
+
+\begin{proof}
+    Let $X$ denote a random variable for the Monte Carlo samples used to estimate the feature effect.
+    For better readability, we will omit the subscript $S$ as well as the point $x_S$ in this proof:
+
+    \begin{align*}
+        \text{MSE}(PDP_{f}, \widehat{PDP}_{\hat f})
+         & = \mathbb{E}_F\mathbb{E}_X[{(PDP_f - \widehat{PDP}_{\hat f})}^2]                                                             \\
+         & = \mathbb{E}_F \mathbb{E}_X[PDP_f^2 - 2PDP_f\widehat{PDP}_{\hat f} + \widehat{PDP}_{\hat f}^2]                               \\
+         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\mathbb{E}_X[\widehat{PDP}_{\hat f}^2]                            \\
+         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}]                              \\
+         & \quad + \mathbb{E}_F[\mathbb{E}_X{[\widehat{PDP}_{\hat f}]}^2]                                                               \\
+         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}]                              \\
+         & \quad + \text{Var}_F(\mathbb{E}_X[\widehat{PDP}_{\hat f}]) + \mathbb{E}_F{(\mathbb{E}_X[\widehat{PDP}_{\hat f}])}^2          \\
+         & = PDP_f^2 - 2PDP_f\mathbb{E}_F[PDP_{\hat f}] +  Var_F[PDP_{\hat f}]                                                          \\
+         & \quad + \mathbb{E}_F{[PDP_{\hat f}]}^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}]                                    \\
+         & = {(PDP_f - \mathbb{E}_F[PDP_{\hat f}])}^2 + \text{Var}_F[PDP_{\hat f}] + \mathbb{E}_F[\text{Var}_X(\widehat{PDP}_{\hat f})]
+    \end{align*}
+
+    \noindent At multiple points, we use the fact that $\mathbb{E}_X[\widehat{PDP}_{\hat f}]
+        = PDP_{\hat f}$ (cf.~\cite{molnar_relating_2023}).
+        \qed\
+\end{proof}
+
 \section{Proof: Monte Carlo Variance for Groundtruth}\label{app:proof-mc-variance}
 \begin{proof}
-    For better readability, we will omit the subscript $S$ as well as the point $x_S$ and use $X=X_{mc}$ in this proof:
+    Let $X$ denote a random variable for the Monte Carlo samples used to estimate the feature effect.
+    For better readability, we will omit the subscript $S$ as well as the point $x_S$ in this proof:
 
     \begin{align*}
-        \mathbb{E}_F\mathbb{E}_X[(PDP_f - \widehat{PDP}_f)^2] & = \mathbb{E}_F\mathbb{E}_X[PDP_f^2 - 2PDP_f \widehat{PDP}_f + \widehat{PDP}_f^2]                                 \\
+        \text{MSE}(PDP_{f}, \widehat{PDP}_{f}) & = \mathbb{E}_F\mathbb{E}_X[{(PDP_f - \widehat{PDP}_f)}^2]                                                          \\
+                                               & = \mathbb{E}_F\mathbb{E}_X[PDP_f^2 - 2PDP_f \widehat{PDP}_f + \widehat{PDP}_f^2]                                   \\
                                                & = PDP_f^2 - 2PDP_f \mathbb{E}_F\mathbb{E}_X[\widehat{PDP}_f] + \mathbb{E}_F\mathbb{E}_X[\widehat{PDP}_f^2]         \\
-                                                              & = PDP_f^2 - 2PDP_f^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_f] + \mathbb{E}_F[\mathbb{E}_X[\widehat{PDP}_f]^2] \\
+                                               & = PDP_f^2 - 2PDP_f^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_f] + \mathbb{E}_F[\mathbb{E}_X{[\widehat{PDP}_f]}^2] \\
                                                & = PDP_f^2 - 2PDP_f^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_f] + PDP_f^2                                         \\
@@
     that all quantities based on $f$ do not depend on the model distribution $F$.
+    \qed\
 \end{proof}
@@
         \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 20; lam: 15.6807;                                                                                                                                                                       \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 800; gamma: 10;                                                                                                                                                                                \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 917.9061; gamma: 0.0030;                                                                                                                                                                       \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C:\@800; gamma: 10;                                                                                                                                                                                \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C:\@917.9061; gamma: 0.0030;                                                                                                                                                                       \\
         \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1200; max\_depth: 16; learning\_rate: 0.35; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                           \\
@@
         \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 5; lam: 0.0010;                                                                                                                                                                         \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 1000; gamma: 10;                                                                                                                                                                               \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 864.4724; gamma: 0.0085;                                                                                                                                                                       \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C:\@1000; gamma: 10;                                                                                                                                                                               \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C:\@864.4724; gamma: 0.0085;                                                                                                                                                                       \\
         \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1500; max\_depth: 18; learning\_rate: 0.4000; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                         \\
@@
         \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 21; lam: 0.0402;                                                                                                                                                                        \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 1000; gamma: 15;                                                                                                                                                                               \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 917.1949; gamma: 0.2102;                                                                                                                                                                       \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C:\@1000; gamma: 15;                                                                                                                                                                               \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C:\@917.1949; gamma: 0.2102;                                                                                                                                                                       \\
         \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1000; max\_depth: 14; learning\_rate: 0.3; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                            \\
@@
         \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 22; lam: 0.0657;                                                                                                                                                                        \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 1000; gamma: 18;                                                                                                                                                                               \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 901.1903; gamma: 0.2602;                                                                                                                                                                       \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C:\@1000; gamma: 18;                                                                                                                                                                               \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C:\@901.1903; gamma: 0.2602;                                                                                                                                                                       \\
         \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1200; max\_depth: 14; learning\_rate: 0.3; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                            \\
@@
         \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 31; lam: 0.3260;                                                                                                                                                                        \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 200; gamma: 8;                                                                                                                                                                                 \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 11.4317; gamma: 0.1394;                                                                                                                                                                        \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C:\@200; gamma: 8;                                                                                                                                                                                 \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C:\@11.4317; gamma: 0.1394;                                                                                                                                                                        \\
         \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1000; max\_depth: 14; learning\_rate: 0.3; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                            \\
@@
         \textbf{}                               & \textbf{}                         & \textbf{GAM\_OT}     & n\_bases: 32; lam: 0.4500;                                                                                                                                                                        \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C: 400; gamma: 10;                                                                                                                                                                                \\
-        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C: 22.5167; gamma: 0.1114;                                                                                                                                                                        \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OF}     & C:\@400; gamma: 10;                                                                                                                                                                                \\
+        \textbf{}                               & \textbf{}                         & \textbf{SVM\_OT}     & C:\@22.5167; gamma: 0.1114;                                                                                                                                                                        \\
         \textbf{}                               & \textbf{}                         & \textbf{XGBoost\_OF} & n\_estimators: 1000; max\_depth: 14; learning\_rate: 0.3; subsample: 1.0; min\_child\_weight: 1; colsample\_bytree: 1.0; colsample\_bylevel: 1.0; lambda: 0; alpha: 0;                            \\
@@
     \caption{Hyperparameters for the models used in the simulation study}
-    \label{tab:model-hyperparameters}
+    \label{tab:model-hyperparameters}  % chktex 24
 \end{table}
@@
 
-\begin{figure}
-    \includegraphics[width=\textwidth]{img/model_performances.png}
-    \caption{R2-score of the models on training (left) and holdout test data (right), each boxplot aggregates 30 repetitions}
-    \label{fig:model-performance}
-\end{figure}
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{img/model_performance_SimpleNormalCorrelated.png}
+        \caption{Model performances on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{img/model_performance_Friedman1.png}
+        \caption{Model performances on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{img/model_performance_FeynmanI.29.16.png}
+        \caption{Model performances on Feynman I.29.16}
+    \end{subfigure}
+    \caption{R2-scores of the models evaluated on training samples (left) and 10000 holdout test samples (right).
+Each boxplot aggregates 30 repetitions with differing seeds for the data generation (cf.
+\textsc{Section~\ref{sec:experiment-main}}).}
+    \label{fig:model-performance}  % chktex 24
+\end{figure*}
 
@@
 \begin{figure*}[h!]
-    \ContinuedFloat
+    \ContinuedFloat\
     \centering
@@
     \caption{Main experiment results for PDP on SimpleNormalCorrelated dataset. $x_3$ and $x_4$ are dummy variables.}
-    \label{fig:pdp-results-snc-all}
+    \label{fig:pdp-results-snc-all}  % chktex 24
 \end{figure*}
@@
     \caption{Main experiment results for ALE on SimpleNormalCorrelated dataset. $x_3$ and $x_4$ are dummy variables.}
-    \label{fig:ale-results-snc-all}
+    \label{fig:ale-results-snc-all}  % chktex 24
 \end{figure*}
@@
     \centering
-    \ContinuedFloat
+    \ContinuedFloat\
     \begin{subfigure}[b]{\textwidth}
@@
     \caption{Main experiment results for PDP on Friedman1 dataset. $x_6$ and $x_7$ are dummy variables.}
-    \label{fig:pdp-results-friedman1}
+    \label{fig:pdp-results-friedman1}  % chktex 24
 \end{figure*}
@@
     \caption{Main experiment results for ALE on Friedman1 dataset. $x_6$ and $x_7$ are dummy variables.}
-\label{fig:ale-results-friedman1}
+    \label{fig:ale-results-friedman1}  % chktex 24
 \end{figure*}
@@
 
-\begin{figure*}[h!]
+\begin{figure*}[ht!]
     \centering
@@
     \caption{Main experiment results for PDP on FeynmanI.29.16 dataset. $d_1$ and $d_2$ are dummy variables.}
-    \label{fig:pdp-results-feynman}
+    \label{fig:pdp-results-feynman}  % chktex 24
 \end{figure*}
 
-\begin{figure*}[h!]
+\begin{figure*}[ht!]
     \centering
@@
     \end{subfigure}
+    \label{fig:ale-results-feynman}  % chktex 24
     \caption{Main experiment results for ALE on FeynmanI.29.16 dataset. $d_1$ and $d_2$ are dummy variables.}
-    \label{fig:ale-results-feynman}
 \end{figure*}
 
-\end{document}
+\clearpage
+\section{Further Investigation of Main Experiment Results}\label{app:main-experiment-further-investigation}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{img/further_investigation/snc_pdp_XGBoost_OF_1000.png}
+    \caption{$\text{Bias}^2$ and variance per grid point for PDP of XGBoost\_OF on SimpleNormalCorrelated dataset with $n_{train} = 1000$.
+    This example demonstrates the cross-validation reduces the variance mainly in regions where data is scarce, i.e.,
+    at the edges of the feature space. In this regions, the model has only few data points to learn from, and approximations
+    of the feature effect may be less reliable.}
+    \label{fig:further-pdp-xgb-snc}  % chktex 24
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{img/further_investigation/feynman_ale_SVM_OT_1000.png}
+    \caption{$\text{Bias}^2$ and variance per grid point for ALE of SVM\_OT on FeynmanI.29.16 dataset with $n_{train} = 1000$.
+    This example shows that the higher bias on validation data and in the cross-validated scenario for ALE mainly occurs in regions
+    where data is scarce. For $x_1$ and $x_2$, the marginals are log-uniformly distributed, which leads to a lower density of data points
+    in the higher ranges of the features, for which the bias increases. This supports our hypothesis that the bias is due to our
+    theoretical-quantile-based inteval selection, where intervals to the right may empirically contain few to no data points.}
+    \label{fig:further-ale-svm-feynman}  % chktex 24
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{img/further_investigation/friedman1_ale_XGBoost_OF_8000.png}
+    \caption{$\text{Bias}^2$ and variance per grid point for ALE of XGBoost\_OF on Friedman1 dataset with $n_{train} = 8000$.
+    For all features, the bias when estimated on training data is substantially larger than in the other scenarios. The bias
+    is largest mostly on the limits of the feature ranges, and seems to be influenced by the shape of the function.
+    Note that we do not observe this behavior for the corresponding PDP, nor for other datasets and models. 
+    }
+    \label{fig:further-ale-xgb-friedman}  % chktex 24
+\end{figure}
+
+\end{document}  % chktex 17

### 2025-02-11 paper/paper.tex
paper: add chapter and appendix on mc error ablation study results

@@
 
+Finally, we also observe that the Monte Carlo variance --- as expected --- reduced with larger dataset size.
+We investigate this effect further in the next ablation study.
+
 \subsection{Ablation Study Results: Impact of Dataset Size on Monte Carlo Variance}\label{sec:results-dataset-size}
+
+We show the effect of the dataset size on the Monte Carlo variance for the features of the Friedman1 dataset
+for PDP and ALE in \textbf{Fig.\@~\ref{fig:mc-variance}}. The full results are provided in \textsc{Appendix~\ref{app:additional-mc-results}}.
+Naturally, we observe that the Monte Carlo variance decreases with increasing dataset size (except for $x_3$ and $x_4$ for PDP,
+where the variance is generally very small). We also see differences between the features: for more complex features like $x_1$ and $x_2$,
+which include interactions, the Monte Carlo variance is generally higher compared to simpler features like $x_3$ and $x_4$.
+Nonetheless, we observe an interesting difference between PDP and ALE. While they converge to a similar Monte Carlo variance for  % chktex 13
+large dataset sizes, the initial Monte Carlo variance for very small sample sizes is substantially higher for ALE than for PDP.  % chktex 13
+This demonstrates that ALE is more sensitive to the dataset size than PDP. This is most likely due to the fact that ALE  % chktex 13
+estimates effects per interval and thus requires more samples to estimate the effect accurately compared to PDP, which
+uses the full dataset to estimate the effect. Also note that this analysis shows that the Monte Carlo variance introduced
+by estimating the ground truth feature effects in the main study is for most cases negligible compared to the other error sources.
+
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/mean_mc_variance_friedman1_pdp.png}
+        \caption{Mean Monte Carlo variance for PDP}
+        \label{fig:mc-variance-pdp}  % chktex 24
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/mean_mc_variance_friedman1_ale.png}
+        \caption{Mean Monte Carlo variance for ALE}
+        \label{fig:mc-variance-ale}  % chktex 24
+    \end{subfigure}
+    \caption{Mean Monte Carlo variance for PDP and ALE on Friedman1 dataset.
+    For each data size $n_{mc}$, variances are averaged over all grid points.
+    Second linear feature ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability.}
+    \label{fig:mc-variance}  % chktex 24
+\end{figure*}
+
+In addition, a detailed analysis of the Monte Carlo variance per grid point reveals that for PDP,
+the variance mostly depends on the sample density, while for ALE, the variance is also influenced by the
+shape of the underlying function. We provide detailed results in \textsc{Appendix~\ref{app:additional-mc-results}}.
 
@@
 \clearpage
-\section{Further Investigation of Main Experiment Results}\label{app:main-experiment-further-investigation}
+\section{Further Investigation of Main Experiment Results}\label{app:main-study-further-investigation}
 
@@
 
+\clearpage
+\section{Additional Results for Ablation Study: Datasize Impact on Monte Carlo Variance}\label{app:additional-mc-results}
+
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/SNC-all/mean_mc_variance_pdp.png}
+        \caption{Mean Monte Carlo variance on SimpleNormalCorrelated for PDP}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/SNC-all/mean_mc_variance_ale.png}
+        \caption{Mean Monte Carlo variance on SimpleNormalCorrelated for ALE}
+    \end{subfigure}
+    \caption{Mean Monte Carlo variance for PDP and ALE on SimpleNormalCorrelated dataset.
+    For each data size $n_{mc}$, variances are averaged over all grid points.
+    Note that $x_3$ and $x_4$ are dummy features.}
+    \label{fig:mean-mc-variance-snc}  % chktex 24
+\end{figure*}
+
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/mean_mc_variance_pdp.png}
+        \caption{Mean Monte Carlo variance on Friedman1 for PDP}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/mean_mc_variance_ale.png}
+        \caption{Mean Monte Carlo variance on Friedman1 for ALE}
+    \end{subfigure}
+    \caption{Mean Monte Carlo variance for PDP and ALE on Friedman1 dataset.
+    For each data size $n_{mc}$, variances are averaged over all grid points.
+    Note that $x_6$ and $x_7$ are dummy features.}
+    \label{fig:mean-mc-variance-friedman1}  % chktex 24
+\end{figure*}
+
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/SNC-all/mc_variance_pdp.png}
+        \caption{Monte Carlo variance on SimpleNormalCorrelated for PDP}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/SNC-all/mc_variance_ale.png}
+        \caption{Monte Carlo variance on SimpleNormalCorrelated for ALE}
+    \end{subfigure}
+    \caption{Monte Carlo variance for PDP and ALE on SimpleNormalCorrelated dataset.
+    The variances for different data sizes $n_{mc}$ (colors) are shown for each grid
+    point over the range of each feature (x-axis). The ground truth feature effect is
+    shown as dashed line for reference, using a different y-axis scale omitted for
+    clarity. Note that $x_3$ and $x_4$ are dummy features.}
+    \label{fig:mc-variance-snc}  % chktex 24
+\end{figure*}
+
+\begin{figure*}[h!]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/mc_variance_pdp.png}
+        \caption{Monte Carlo variance on Friedman1 for PDP}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/mc_variance_ale.png}
+        \caption{Monte Carlo variance on Friedman1 for ALE}
+    \end{subfigure}
+    \caption{Monte Carlo variance for PDP and ALE on Friedman1 dataset.
+    The variances for different data sizes $n_{mc}$ (colors) are shown for each grid
+    point over the range of each feature (x-axis). The ground truth feature effect is
+    shown as dashed line for reference, using a different y-axis scale omitted for
+    clarity. Note that $x_6$ and $x_7$ are dummy features.}
+    \label{fig:mc-variance-friedman1}  % chktex 24
+\end{figure*}
+
 \end{document}  % chktex 17

### 2025-02-14 current_research_feature_effects/plotting/plots.py
Add ablation study variance decomposition analysis

@@ -307,6 +307,70 @@ def plot_feature_effect_error_table(
     plt.show()
 
 
+def plot_variance_table(
+    df: pd.DataFrame,
+    models: List[str],
+    type: Literal["pdp", "ale"],
+    save_figs: None | Path = None,
+    large_font: bool = False,
+    show_title: bool = True,
+):
+    """
+    Plot a table of feature effect variances for different models.
+
+    Parameters
+    ----------
+    df : pd.DataFrame
+        Dataframe containing the aggregated variances.
+    models : List[str]
+        List of model names.
+    type : Literal[&quot;pdp&quot;, &quot;ale&quot;]
+        Type of feature effect, either 'pdp' or 'ale'.
+    save_figs : None | Path
+        Path to save the figure (default is None).
+    large_font : bool
+        If True, use larger font sizes (default is False).
+    show_title : bool
+        If True, show the title of the plot (default is True).
+    """
+    for model in models:
+        if large_font:
+            _set_fontsize("xlarge")
+        else:
+            _set_fontsize("standard")
+
+        g = sns.FacetGrid(
+            df.loc[df["model"] == model], row="n_train", col="feature", height=3, sharey="row", aspect=0.67
+        )
+        g.map_dataframe(
+            sns.barplot,
+            x="split",
+            y="value",
+            hue="metric",
+            hue_order=["Variance", "Model Variance", "MC Variance"],
+            palette=sns.color_palette("Set2")[2:5],
+        )
+        for ax in g.axes.flat:
+            for bar in ax.patches:
+                if bar.get_height() < 0:
+                    bar.set_hatch('////')
+                    bar.set_edgecolor((0, 0, 0, 0.3))
+        g.set_titles(col_template="${col_name}$", row_template="n_train={row_name}")
+        if show_title:
+            g.fig.suptitle(f"Variance Decomposition {type.upper()} {model}", y=1.02)
+            legend_pos = (0.5, 1.0)
+        else:
+            g.fig.suptitle("", y=1.02)
+            legend_pos = (0.5, 1.05)
+        g.add_legend(loc="upper center", bbox_to_anchor=legend_pos, ncol=3)
+        g.set_ylabels("")
+        plt.tight_layout()
+        if save_figs is not None:
+            g.savefig(save_figs / f"variance_decomposition_{type}_{model}.png", bbox_inches="tight", dpi=300)
+
+    plt.show()
+
+
[…]

### 2025-02-14 current_research_feature_effects/plotting/table_formatting.py

@@ -1,3 +1,6 @@
+from pathlib import Path
+from configparser import ConfigParser
+from typing import Literal
 import pandas as pd
 import numpy as np
 
@@ -16,12 +19,7 @@ def format_dataframe(df: pd.DataFrame) -> pd.DataFrame:
     pd.DataFrame
         Formatted dataframe.
     """
-    pivoted = df.pivot_table(
-        index=["model", "n_train", "split"], columns=["feature", "metric"], values="value"
-    ).reset_index()
-
-    pivoted.columns.name = None
-    df_out = pivoted.groupby(by=["model", "n_train", "split"]).mean()
+    df_out = get_grouped_df(df)
     sorted_cols = [
         (feature, metric)
         for feature in sorted(set(col[0] for col in df_out.columns))
@@ -67,3 +65,141 @@ def highlight_min_feature_metric(data: pd.DataFrame) -> pd.DataFrame:
                 mask.loc[group_mask, col] = is_min
 
     return pd.DataFrame(np.where(mask, "font-weight: bold", ""), index=data.index, columns=data.columns)
+
+
+def get_concatenated_results(
+    dataset: str,
+    effect: Literal["pdp", "ale"],
+    config: ConfigParser,
+    main_config: ConfigParser,
+    main_experiment_path: Path,
+    experiment_path: Path,
+):
+    """
+    Get the concatenated results of the main and ablation experiment
+    for the analysis of variance decomposition.
+
+    Parameters
+    ----------
+    dataset : str
+        Name of the dataset.
+    effect : str
+        Name of the effect.
+    config : configparser.ConfigParser
+        Configuration of the ablation experiment.
+    main_config : configparser.ConfigParser
+        Configuration of the main experiment.
+    main_experiment_path : pathlib.Path
+        Path to the main experiment results.
+    experiment_path : pathlib.Path
+        Path to the ablation experiment results.
+
+    Returns
+    -------
+    pd.DataFrame
+        Concatenated results of the main and ablation experiment.
+    """
+    df_main = pd.read_sql_table(
+        f"{effect}_results",
+        f"sqlite:///{main_experiment_path}/{dataset}/{main_config.get('storage', 'effects_results')}",
+    )
+    df_main_filtered = df_main.loc[(df_main["model"] == "XGBoost_OT") | (df_main["model"] == "XGBoost_OF")]
+    df_ablation = pd.read_sql_table(
[…]

### 2025-02-14 paper/paper.tex
paper: add chapter on variance decomposition ablation study

@@
 equation~(\ref{eq:mc-variance-estimator-model}). Since this study is
-computationally more expensive, we do this only for XGBoost (OT and OF) with
-the same datasets and sizes as in the main experiment.
+computationally more expensive, we do this only for XGBoost (OT and OF), which
+is arguably the most practically relevant model in our set-up. We use the same
+datasets and sizes as in the main experiment.
 
@@
 which may especially improve overfitting models that have a higher model variance.
-We investigate these hypotheses in an ablation study in \textsc{Section~\ref{sec:results-variance-decomposition}}.
+We further investigate these hypotheses in \textsc{Section~\ref{sec:results-variance-decomposition}}.
 
@@
 due to a high bias. This occurs for the XGBoost\_OF model on the Friedman1 dataset, which we further
-look into in \textsc{Appendix~\ref{app:main-study-further-investigation}}.
+look into in \textsc{Appendix~\ref{app:main-study-further-investigation}}. Also note that the we observe
+no substantial differences between scenarios with correlated features (SimpleNormalCorrelated) and
+independent features (Friedman1, Feynman I.29.16).
 
@@
 
-\dots
-
-Finally, we also observe that the Monte Carlo variance --- as expected --- reduced with larger dataset size.
-We investigate this effect further in the next ablation study.
+We use the variance results from the main experiment and the Monte Carlo variance estimated in the ablation study
+to show the decomposition of the feature effect variance into model variance and Monte Carlo variance. The results
+for the Friedman1 dataset are given in \textbf{Fig.\@~\ref{fig:pdp-variance-decomp}} for PDP and
+\textbf{Fig.\@~\ref{fig:ale-variance-decomp}} for ALE. The full results including the other datasets are  % chktex 13
+provided in \textsc{Appendix~\ref{app:additional-variance-decomposition}}.
+
+We observe that the Monte Carlo variance is generally highest when the feature effects are estimated on the smaller
+validation set, which aligns with our expectations.
+Computing the feature effects in a cross-validated manner leads to the lowest Monte Carlo variance with a substantial
+reduction compared to estimation solely on training or validation data.
+
+In addition, we observe that while the model variance is naturally nearly the same for training and validation data,
+it is also reduced in many cases when estimating the feature effects in a
+cross-validated manner. This is especially pronounced for the overfitting models (OF) and is best visible for PDP
+(\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}}). These observations are in line with our previous hypothesis that
+reduces both model variance by averaging over different models as well as Monte Carlo variance by averaging over
+different samples. This is especially beneficial for overfitting models, which have a higher model variance.
+
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}[b]{0.75\textwidth}
+        \includegraphics[width=\textwidth]{img/variance_decomposition_pdp_XGBoost_OF.png}
+        \caption{XGBoost\_OF PDP on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{0.75\textwidth}
+        \includegraphics[width=\textwidth]{img/variance_decomposition_pdp_XGBoost_OT.png}
+        \caption{XGBoost\_OT PDP on Friedman1}
+    \end{subfigure}
+    \caption{Variance decomposition results for PDP on Friedman1 dataset. Variance is taken from the
+    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
+    obtained by subtracting the Monte Carlo variance from the total variance. The second linear feature
+    ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability. Results for XGBoost\_OT
+    on 8000 training samples were not obtained due to excessive computation time.}
+    \label{fig:pdp-variance-decomp}  % chktex 24
+\end{figure}
+
+While the variance due to MC integration is only a minor component of the total variance for PDP, it is the
+dominant component for ALE. Only for feature effect estimation via cross-validation, the Monte Carlo variance  % chktex 13
+is again reduced to a negligible level. This suggests that ALE is more sensitive to the dataset size than PDP,
+which we further investigate in \textsc{Section~\ref{sec:results-dataset-size}}. We also suspect a certain
+amount of inaccuracy in the Monte Carlo variance estimation for ALE, which (a) leads to differences in the
+model variance between training and validation set and (b) causes the model variance to be negative in some cases
+(see \textbf{Fig.\@\ref{fig:ale-variance-decomp}}).
+Naturally, this is not possible, and can most likely be attributed to numerical inaccuracies in the MC variance estimation.
+
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}[b]{0.75\textwidth}
+        \includegraphics[width=\textwidth]{img/variance_decomposition_ale_XGBoost_OF.png}
+        \caption{XGBoost\_OF ALE on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{0.75\textwidth}
+        \includegraphics[width=\textwidth]{img/variance_decomposition_ale_XGBoost_OT.png}
+        \caption{XGBoost\_OT ALE on Friedman1}
+    \end{subfigure}
+    \caption{Variance decomposition results for ALE on Friedman1 dataset. Variance is taken from the
+    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
+    obtained by subtracting the Monte Carlo variance from the total variance. The second linear feature
+    ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability. Results for XGBoost\_OT
+    on 8000 training samples were not obtained due to excessive computation time. Negative variances are
+    due to numerical inaccuracies, marked by the hatch pattern.}
+    \label{fig:ale-variance-decomp}  % chktex 24
+\end{figure}
+
+For both PDP and ALE, we find that the Monte Carlo variance, as expected, reduces with larger dataset size.
+This becomes clear both when comparing training and validation data and when comparing the two dataset sizes
+(1000 and 8000 samples). To analyze this not only for these few dataset sizes, we dedicate a separate
+ablation study to investigate this further in the next section.
 
@@
     \caption{R2-scores of the models evaluated on training samples (left) and 10000 holdout test samples (right).
-Each boxplot aggregates 30 repetitions with differing seeds for the data generation (cf.
-\textsc{Section~\ref{sec:experiment-main}}).}
+Each boxplot aggregates 30 repetitions with differing seeds for the data generation (cf.\ \textsc{Section~\ref{sec:experiment-main}}).}
     \label{fig:model-performance}  % chktex 24
@@
 \clearpage
+\section{Additional Results for Ablation Study: Variance Decomposition}\label{app:additional-variance-decomposition}
+
+\subsection{SimpleNormalCorrelated Dataset}
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}[b]{0.75\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC-all/variance_decomposition_pdp_XGBoost_OF.png}
+        \caption{XGBoost\_OF PDP on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{0.75\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC-all/variance_decomposition_pdp_XGBoost_OT.png}
+        \caption{XGBoost\_OT PDP on SimpleNormalCorrelated}
+    \end{subfigure}
+    \caption{Variance decomposition results for PDP on SimpleNormalCorrelated dataset. Variance is taken from the
+    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
+    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
+    on 8000 training samples were not obtained due to excessive computation time.}
+    \label{fig:pdp-variance-decomp-snc}  % chktex 24
+\end{figure}
+
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}[b]{0.75\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC-all/variance_decomposition_ale_XGBoost_OF.png}
+        \caption{XGBoost\_OF ALE on SimpleNormalCorrelated}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{0.75\textwidth}
+        \includegraphics[width=\textwidth]{img/SNC-all/variance_decomposition_ale_XGBoost_OT.png}
+        \caption{XGBoost\_OT ALE on SimpleNormalCorrelated}
+    \end{subfigure}
+    \caption{Variance decomposition results for PDP on SimpleNormalCorrelated dataset. Variance is taken from the
+    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
+    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
+    on 8000 training samples were not obtained due to excessive computation time. Negative variances are
+    due to numerical inaccuracies, marked by the hatch pattern.}
+    \label{fig:ale-variance-decomp-snc}  % chktex 24
+\end{figure}
+
+\clearpage
+\subsection{Friedman1 Dataset}
+
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \includegraphics[width=\textwidth]{img/Friedman1-all/variance_decomposition_pdp_XGBoost_OF.png}
+        \caption{XGBoost\_OF PDP on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \includegraphics[width=\textwidth]{img/Friedman1-all/variance_decomposition_pdp_XGBoost_OT.png}
+        \caption{XGBoost\_OT PDP on Friedman1}
+    \end{subfigure}
+    \caption{Variance decomposition results for PDP on Friedman1 dataset. Variance is taken from the
+    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
+    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
+    on 8000 training samples were not obtained due to excessive computation time.}
+    \label{fig:pdp-variance-decomp-friedman1}  % chktex 24
+\end{figure}
+
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}[b]{\textwidth}
+        \includegraphics[width=\textwidth]{img/Friedman1-all/variance_decomposition_ale_XGBoost_OF.png}
+        \caption{XGBoost\_OF ALE on Friedman1}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{\textwidth}
+        \includegraphics[width=\textwidth]{img/Friedman1-all/variance_decomposition_ale_XGBoost_OT.png}
+        \caption{XGBoost\_OT ALE on Friedman1}
+    \end{subfigure}
+    \caption{Variance decomposition results for PDP on Friedman1 dataset. Variance is taken from the
+    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
+    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
+    on 8000 training samples were not obtained due to excessive computation time. Negative variances are
+    due to numerical inaccuracies, marked by the hatch pattern.}
+    \label{fig:ale-variance-decomp-friedman1}  % chktex 24
+\end{figure}
+
+\clearpage
+\subsection{Feynman I.29.16 Dataset}
+
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/variance_decomposition_pdp_XGBoost_OF.png}
+        \caption{XGBoost\_OF PDP on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/variance_decomposition_pdp_XGBoost_OT.png}
+        \caption{XGBoost\_OT PDP on FeynmanI.29.16}
+    \end{subfigure}
+    \caption{Variance decomposition results for PDP on FeynmanI.29.16 dataset. Variance is taken from the
+    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
+    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
+    on 8000 training samples were not obtained due to excessive computation time.}
+    \label{fig:pdp-variance-decomp-feynman}  % chktex 24
+\end{figure}
+
+\begin{figure}[htbp]
+    \centering
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/variance_decomposition_ale_XGBoost_OF.png}
+        \caption{XGBoost\_OF ALE on FeynmanI.29.16}
+    \end{subfigure}
+    \\[10pt]
+    \vfill
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/variance_decomposition_ale_XGBoost_OT.png}
+        \caption{XGBoost\_OT ALE on FeynmanI.29.16}
+    \end{subfigure}
+    \caption{Variance decomposition results for PDP on FeynmanI.29.16 dataset. Variance is taken from the
+    main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
+    obtained by subtracting the Monte Carlo variance from the total variance. Results for XGBoost\_OT
+    on 8000 training samples were not obtained due to excessive computation time. Negative variances are
+    due to numerical inaccuracies, marked by the hatch pattern.}
+    \label{fig:ale-variance-decomp-feynman}  % chktex 24
+\end{figure}
+
+\clearpage
 \section{Additional Results for Ablation Study: Datasize Impact on Monte Carlo Variance}\label{app:additional-mc-results}
+
+\subsection{Aggregated Results}
 
@@
 \end{figure*}
+
+\clearpage
+\subsection{Results per Grid Point}
 

### 2025-02-15 paper/paper.tex
paper: add conclusion chapter and abstract

@@
 \usepackage{tabularx}
+\usepackage{hyperref}
 \usepackage{ragged2e}
@@
 \begin{abstract}
-    The abstract should briefly summarize the contents of the paper in
-    150--250 words.
-
-    \keywords{Explainable AI  \and Feature effects \and Partial dependence plot \and Accumulated local effects}
+    As Machine Learning models are increasingly employed in critical domains
+    like healthcare and finance, ensuring accurate model explanations by
+    Interpretable Machine Learning methods becomes crucial to avoid misguided
+    decisions. The choice of dataset for estimating feature effects --- whether
+    to use training data, validation data, or even employ cross-validation ---
+    is one important fundamental question in this context that remains largely
+    unexplored.
+    Through a comprehensive simulation study, we investigate this question for
+    Partial Dependence Plots (PDP) and Accumulated Local Effects (ALE), decomposing
+    their error into bias and variance components. We extend previous work by
+    providing a theoretical framework for error decomposition and contribute
+    a systematic overview of test functions for simulation studies in Interpretable ML.  % chktex 13
+    Our simulation study results validate the common practice of using training
+    data, showing no evidence against it. We find that larger datasets for estimation
+    are generally preferable due to reduced Monte Carlo variance, with ALE being
+    particularly sensitive to dataset size. Cross-validation emerges as a promising
+    approach, reducing both model and Monte Carlo variance, which is especially
+    beneficial for overfitting models.
+    Our findings enable practitioners to make informed decisions about dataset
+    choice for feature effect estimation while establishing a foundation for
+    future research in this area.\footnote{Our entire implementation and results
+    are available at \url{https://github.com/timo282/current-research-feature-effects}.}
+
+    \keywords{Interpretable Machine Learning  \and Feature effects \and Partial dependence plot \and Accumulated local effects}
 \end{abstract}
@@
 systems that intrinsically do not allow insight into their internal reasoning,
-making it often impossible to explain their decisions. However, this can be
-problematic in many critical domains and applications, such as the healthcare,
-legal, or finance sectors, where decisions must be transparent and
+making it often impossible to explain their decisions. This can be
+problematic in many critical domains and applications, such as healthcare,
+legal, or finance, where decisions must be transparent and
 accountable~\cite{adadi_peeking_2018}.
 
-Interpretability in ML is crucial to enhance
-trust~\cite{ribeiro_why_2016,teach_analysis_1981}, address potential
-biases~\cite{guidotti_survey_2019}, fairness and ethical concerns
-~\cite{lipton_mythos_2018}, and ensure compliance with regulations such as the
+The need for interpretability in ML stems from multiple factors: building trust in
+model predictions~\cite{ribeiro_why_2016,teach_analysis_1981},  identifying and mitigating
+potential biases~\cite{guidotti_survey_2019}, addressing fairness and ethical concerns
+~\cite{lipton_mythos_2018}, and ensuring compliance with regulations like the
 EU's General Data Protection Regulation (GDPR)~\cite{gdpr2016} and AI
 Act~\cite{euaia2024}. To address these challenges, the field of Explainable AI
-/ Interpretable ML has emerged~\cite{adadi_peeking_2018}. In the last years, a
-wide variety of methods has been developed, including model-specific and
+/ Interpretable ML has emerged~\cite{adadi_peeking_2018}, develping a
+wide variety of methods to explain ML models. This includes both model-specific and
 model-agnostic approaches ranging from local feature attributions to global
-feature importances and effects\footnote{For an overview of Explainable AI
+feature importances and effects\footnote{For comprehensive overviews of Explainable AI
     methods, see
@@
 Due to the severity of many applications, it is crucial to utilize these
-explainability methods correctly to avoid misleading or incorrect conclusions.
+interpretability methods correctly to avoid misleading or incorrect conclusions.
 In general, there are many pitfalls to be aware of~\cite{molnar_general_2022},
-including whether to compute explanations in-sample, i.e.\ on training data, or
+including the question whether to compute explanations in-sample, i.e.\ on training data, or
 out-of-sample, which we refer to as validation data in the following. Existing
 works have studied the implications of this choice for many methods, including
-studies on \textit{Permutation Feature Importance (PFI) } (e.g.,
-in~\cite{molnar_general_2022}), \textit{Mean Decrease in Impurity (MDI)} of
+studies on \textit{Permutation Feature Importance (PFI) }~\cite{molnar_general_2022},
+\textit{Mean Decrease in Impurity (MDI)} of
 Random Forests~\cite{loecher_debiasing_2022}, or \textit{SHAP}
@@
 computational constraints. Although PDP and ALE are not based on generalization
-error like PFI, it is not studied if and how they are affected by overfitting.
-In addition, larger datasets may improve the feature effect estimates, but also
-increase computation times.\\
-
-\noindent In this paper, we aim to answer the largely unaddressed, fundamental
+error like PFI, it is not studied if and how they are affected by overfitting
+when estimated on training data. In addition, larger datasets may improve the
+feature effect estimates, but also increase computation times.\\
+
+\noindent In this paper, we aim to answer this largely unaddressed, fundamental
 methodological question of whether to use training or validation data to
@@
 
-\noindent These contributions have several important implications: Firstly, our empirically
+\noindent These contributions have several important implications: First, our empirically
 grounded recommendations enable practitioners and researchers to make informed
-decisions about which data to choose for feature effect computation, helping to
+decisions about which data to choose for feature effect estimation, helping to
 understand potential implications of their choices. Additionally, our framework
@@
 violated, the method may produce unrealistic data points outside the underlying
-joint distribution of the data. This extrapolation issue can cause misleading
-interpretations~\cite{molnar_interpretable_2022,molnar_general_2022}.
+joint data distribution of the data, which can cause misleading
+interpretations (``extrapolation issue'')~\cite{molnar_interpretable_2022,molnar_general_2022}.
 
@@
 effects~\cite{hooker_discovering_2004}. Recent extensions have further
-addressed limitations of these classical methods. For example, \textit{Robust
+addressed limitations of the methods above: \textit{Robust
     and Heterogeneity-aware ALE (RHALE)}~\cite{gkolemis_rhale_2023} adds
-heterogeneity quantification to ALE and improving robustness, or
+heterogeneity quantification to ALE and improves robustness.
 \textit{Accumulated Total Derivative Effect (ATDEV)} plots can be decomposed
@@
 proposed bootstrap-based
-approaches\cite{esselman_landscape_2015,grange_using_2019}. While the former
+approaches~\cite{esselman_landscape_2015,grange_using_2019}. While the former
 are not model-agnostic, the latter often capture only the variance introduced
-by Monte Carlo approximation. Nonetheless, the model variance is another source
+by Monte Carlo approximation. However, the model variance is another source
 of uncertainty in PDPs and ALEs, and accounting for it requires multiple model
@@
 \paragraph{Feature effect estimation.}
-For the trained models, we estimate the feature effects $\widehat{PDP}_{\hat{f},S}$
-and $\widehat{ALE}_{\hat f,S}$ per feature using
+For the trained models, we estimate the feature effects
+$\widehat{PDP}_{\hat{f},S}$ and $\widehat{ALE}_{\hat f,S}$ per feature using
 equations~(\ref{eq:pdp-estimate}) and (\ref{eq:ale-estimate-uncentered}). For
@@
 
-We give the results of the main experiment error decomposition for the SimpleNormalCorrelated
-dataset in \textbf{Fig.\@~\ref{fig:pdp-results-snc}} for PDP and \textbf{Fig.\@~\ref{fig:ale-results-snc}}
-for ALE. The full results for all datasets, models, and features can be found in  % chktex 13
+We give the results of the main experiment error decomposition for the
+SimpleNormalCorrelated dataset in \textbf{Fig.\@~\ref{fig:pdp-results-snc}} for
+PDP and \textbf{Fig.\@~\ref{fig:ale-results-snc}} for ALE. The full results for  % chktex 13
+all datasets, models, and features can be found in
 \textsc{Appendix~\ref{app:additional-results}}.
 
-Generally, we observe that the feature effect error is in the most cases highest when
-the effect is estimated on validation data. For ALE, this observation is more pronounced
-than for PDP. This increase in error is mainly due to an increase in variance (see e.g.,  % chktex 13
-\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}} or \textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-of}}).
-For many models, the feature effect error is lowest when estimated in a cross-validated manner.
-This is slightly more pronounced for overfitting models (OF) than for optimally tuned models (OT),
-and is mainly due to a decrease in variance (see e.g., \textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-of}}
-and \textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}}). Since the variance composes of the model variance
-and the Monte Carlo variance, both could be reduced by cross-validation. Monte Carlo variance
-may decrease as we have slightly more samples overall than in the training set. Conversely,
-the model variance may decrease as it is averaged out through the cross-validation procedure,
-which may especially improve overfitting models that have a higher model variance.
-We further investigate these hypotheses in \textsc{Section~\ref{sec:results-variance-decomposition}}.
+Generally, we observe that the feature effect error is in the most cases
+highest when the effect is estimated on validation data. For ALE, this
+observation is more pronounced than for PDP. This increase in error is mainly  % chktex 13
+due to an increase in variance (see e.g.,
+\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}} or
+\textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-of}}). For many models, the
+feature effect error is lowest when estimated in a cross-validated manner. This
+is slightly more pronounced for overfitting models (OF) than for optimally
+tuned models (OT), and is mainly due to a decrease in variance (see e.g.,
+\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-of}} and
+\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}}). Since the variance composes
+of the model variance and the Monte Carlo variance, both could be reduced by
+cross-validation. Monte Carlo variance may decrease as we have slightly more
+samples overall compared to the training set. Conversely, the model variance
+may decrease as it is averaged out through the cross-validation procedure,
+which may especially improve overfitting models that have a higher model
+variance. We further investigate these hypotheses in
+\textsc{Section~\ref{sec:results-variance-decomposition}}.
 
@@
 
-Interestingly, for ALE, we observe for some cases an increased bias when estimating the feature effects on
-validation data and via cross-validation compared to an estimation on training data (see e.g.,
-\textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-ot}} and \textbf{Fig.\@~\ref{fig:ale-results-snc-svm-ot}}).
-Note that this is only the case when errors are generally low, and the bias is still small in absolute terms.
-Since this only occurs for ALE and only for the smaller dataset size (1000 training samples, 250 validation samples),
-we hypothesize that this is due to our approach of choosing the intervals based on theoretical quantiles.
-For the smaller validation sets, we may have intervals with few or no samples, leading to a bias in the
-estimation of the effect. This hypothesis is supported by the fact that the increase in bias only occurs
-in regions with low sample density, which we demonstrate in \textsc{Appendix~\ref{app:main-study-further-investigation}}.
-Thus, an open question remains whether choosing the intervals based on the empirical quantiles mitigates this issue.
+Interestingly, for ALE, we observe for some cases an increased bias when
+estimating the feature effects on validation data and via cross-validation
+compared to an estimation on training data (see e.g.,
+\textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-ot}} and
+\textbf{Fig.\@~\ref{fig:ale-results-snc-svm-ot}}). Note that this is only the
+case when errors are generally low, and the bias is still small in absolute
+terms. Since this only occurs for ALE and only for the smaller dataset size
+(1000 training samples, 250 validation samples), we hypothesize that this is
+due to our approach of choosing the intervals based on theoretical quantiles.
+For the smaller validation sets, we may have intervals with few or no samples,
+leading to a bias in the estimation of the effect. This hypothesis is supported
+by the fact that the increase in bias only occurs in regions with low sample
+density, which we demonstrate in
+\textsc{Appendix~\ref{app:main-study-further-investigation}}. Thus, an open
+question remains whether choosing the intervals based on the empirical
+quantiles mitigates this issue.
 
@@
 
-Further results on Friedman1 and Feynman I.29.16, as well as for the dummy features, largely confirm
-the described findings. For details, we refer to \textsc{Appendix~\ref{app:additional-results}}. We only
-observe one case where the error of ALE is substantially larger when estimated on training data
-due to a high bias. This occurs for the XGBoost\_OF model on the Friedman1 dataset, which we further
-look into in \textsc{Appendix~\ref{app:main-study-further-investigation}}. Also note that the we observe
-no substantial differences between scenarios with correlated features (SimpleNormalCorrelated) and
-independent features (Friedman1, Feynman I.29.16).
+Overall, we find no evidence against the currently dominant practice of
+estimating feature effects on training data. Further results on Friedman1 and
+Feynman I.29.16, as well as for the dummy features, largely confirm our
+described findings. For details, we refer to
+\textsc{Appendix~\ref{app:additional-results}}. This also means that we observe
+no substantial differences between scenarios with correlated features
+(SimpleNormalCorrelated) and independent features (Friedman1, Feynman I.29.16).
+Similarly, the more real-world relevant Feynman I.29.16 dataset shows results
+in line with the other datasets. Note that we only observe one case where the
+error of ALE is substantially larger when estimated on training data due to a
+high bias. This occurs for the XGBoost\_OF model on the Friedman1 dataset,
+which we further look into in
+\textsc{Appendix~\ref{app:main-study-further-investigation}}.
 
@@
 
-We use the variance results from the main experiment and the Monte Carlo variance estimated in the ablation study
-to show the decomposition of the feature effect variance into model variance and Monte Carlo variance. The results
-for the Friedman1 dataset are given in \textbf{Fig.\@~\ref{fig:pdp-variance-decomp}} for PDP and
-\textbf{Fig.\@~\ref{fig:ale-variance-decomp}} for ALE. The full results including the other datasets are  % chktex 13
-provided in \textsc{Appendix~\ref{app:additional-variance-decomposition}}.
-
-We observe that the Monte Carlo variance is generally highest when the feature effects are estimated on the smaller
-validation set, which aligns with our expectations.
-Computing the feature effects in a cross-validated manner leads to the lowest Monte Carlo variance with a substantial
-reduction compared to estimation solely on training or validation data.
-
-In addition, we observe that while the model variance is naturally nearly the same for training and validation data,
-it is also reduced in many cases when estimating the feature effects in a
-cross-validated manner. This is especially pronounced for the overfitting models (OF) and is best visible for PDP
-(\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}}). These observations are in line with our previous hypothesis that
-reduces both model variance by averaging over different models as well as Monte Carlo variance by averaging over
-different samples. This is especially beneficial for overfitting models, which have a higher model variance.
+We use the variance results from the main experiment and the Monte Carlo
+variance estimated in the ablation study to show the decomposition of the
+feature effect variance into model variance and Monte Carlo variance. The
+results for the Friedman1 dataset are given in
+\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}} for PDP and
+\textbf{Fig.\@~\ref{fig:ale-variance-decomp}} for ALE. Note that we now  % chktex 13
+consider the Friedman1 dataset as it provides a wider variety of different
+feature effects. The full results including the other datasets are provided in
+\textsc{Appendix~\ref{app:additional-variance-decomposition}}.
+
+We observe that the Monte Carlo variance is generally highest when the feature
+effects are estimated on the smaller validation set, which aligns with our
+expectations. Computing the feature effects in a cross-validated manner leads
+to the lowest Monte Carlo variance with a substantial reduction compared to
+estimation solely on training or validation data.
+
+In addition, we observe that the model variance is also reduced in many cases when
+estimating the feature effects in a cross-validated manner compared to an estimation
+on training or validation data, where model variance is (roughly) the same. The observed
+reduction through cross-validation is especially
+pronounced for the overfitting models (OF) and best visible for PDP
+(\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}}). These observations are in line
+with our previous hypothesis that cross-validation reduces both model variance by averaging over
+different models as well as Monte Carlo variance by averaging over different
+samples. This is especially beneficial for overfitting models, which have a
+higher model variance.
 
@@
 
-While the variance due to MC integration is only a minor component of the total variance for PDP, it is the
-dominant component for ALE. Only for feature effect estimation via cross-validation, the Monte Carlo variance  % chktex 13
-is again reduced to a negligible level. This suggests that ALE is more sensitive to the dataset size than PDP,
-which we further investigate in \textsc{Section~\ref{sec:results-dataset-size}}. We also suspect a certain
-amount of inaccuracy in the Monte Carlo variance estimation for ALE, which (a) leads to differences in the
-model variance between training and validation set and (b) causes the model variance to be negative in some cases
-(see \textbf{Fig.\@\ref{fig:ale-variance-decomp}}).
-Naturally, this is not possible, and can most likely be attributed to numerical inaccuracies in the MC variance estimation.
+While the variance due to MC integration is only a minor component of the total
+variance for PDP, it is the dominant component for ALE. Only for feature effect  % chktex 13
+estimation via cross-validation, the Monte Carlo variance is again reduced to a
+negligible level. This suggests that ALE is more sensitive to the dataset size
+than PDP, which we further investigate in
+\textsc{Section~\ref{sec:results-dataset-size}}. We also suspect a certain
+amount of inaccuracy in the Monte Carlo variance estimation for ALE, which (a)
+leads to differences in the model variance between training and validation set
+and (b) causes the model variance to become negative in some cases (see
+\textbf{Fig.\@\ref{fig:ale-variance-decomp}}). Naturally, this is not possible
+and can most likely be attributed to numerical inaccuracies in the MC variance
+estimation.
 
@@
 
-For both PDP and ALE, we find that the Monte Carlo variance, as expected, reduces with larger dataset size.
-This becomes clear both when comparing training and validation data and when comparing the two dataset sizes
-(1000 and 8000 samples). To analyze this not only for these few dataset sizes, we dedicate a separate
-ablation study to investigate this further in the next section.
+For both PDP and ALE, we find that the Monte Carlo variance, as expected,
+reduces with larger dataset size. This becomes clear both when comparing
+training and validation data and when comparing the two dataset sizes (1000 and
+8000 samples). To analyze this not only for those few dataset sizes, we
+dedicate a separate ablation study to investigate this further in the next
+section.
 
@@
 
-We show the effect of the dataset size on the Monte Carlo variance for the features of the Friedman1 dataset
-for PDP and ALE in \textbf{Fig.\@~\ref{fig:mc-variance}}. The full results are provided in \textsc{Appendix~\ref{app:additional-mc-results}}.
-Naturally, we observe that the Monte Carlo variance decreases with increasing dataset size (except for $x_3$ and $x_4$ for PDP,
-where the variance is generally very small). We also see differences between the features: for more complex features like $x_1$ and $x_2$,
-which include interactions, the Monte Carlo variance is generally higher compared to simpler features like $x_3$ and $x_4$.
-Nonetheless, we observe an interesting difference between PDP and ALE. While they converge to a similar Monte Carlo variance for  % chktex 13
-large dataset sizes, the initial Monte Carlo variance for very small sample sizes is substantially higher for ALE than for PDP.  % chktex 13
-This demonstrates that ALE is more sensitive to the dataset size than PDP. This is most likely due to the fact that ALE  % chktex 13
-estimates effects per interval and thus requires more samples to estimate the effect accurately compared to PDP, which
-uses the full dataset to estimate the effect. Also note that this analysis shows that the Monte Carlo variance introduced
-by estimating the ground truth feature effects in the main study is for most cases negligible compared to the other error sources.
+We show the effect of the dataset size on the Monte Carlo variance for the
+features of the Friedman1 dataset for PDP and ALE in
+\textbf{Fig.\@~\ref{fig:mc-variance}}. The full results are provided in
+\textsc{Appendix~\ref{app:additional-mc-results}}. As expected, we observe that
+the Monte Carlo variance decreases with increasing dataset size (except for
+$x_3$ and $x_4$ for PDP, where the variance is generally very small). We
+find differences between the features: for more complex features like $x_1$ and
+$x_2$, which include interactions, the Monte Carlo variance is generally higher
+compared to simpler features like $x_3$ and $x_4$. Nonetheless, we observe an
+interesting difference between PDP and ALE. While they converge to a similar  % chktex 13
+Monte Carlo variance for large dataset sizes, the initial Monte Carlo variance
+for very small sample sizes is substantially higher for ALE than for PDP. This  % chktex 13
+demonstrates that ALE is more sensitive to the dataset size than PDP. This is  % chktex 13
+most likely due to the fact that ALE estimates effects per interval and thus
+requires more samples to estimate the effect accurately compared to PDP, which
+uses the full dataset to estimate the effect. Note that this analysis
+also proves that the Monte Carlo variance introduced by estimating the ground truth
+feature effects in the main study is for most cases negligible compared to the
+other error sources.
 
@@
 
-In addition, a detailed analysis of the Monte Carlo variance per grid point reveals that for PDP,
-the variance mostly depends on the sample density, while for ALE, the variance is also influenced by the
-shape of the underlying function. We provide detailed results in \textsc{Appendix~\ref{app:additional-mc-results}}.
+In addition, a detailed analysis of the Monte Carlo variance per grid point
+reveals that for PDP, the variance mostly depends on the sample density, while
+for ALE, the variance is also influenced by the shape of the underlying
+function. We provide detailed results in
+\textsc{Appendix~\ref{app:additional-mc-results}}.
 
@@
 
+Summarizing our main contributions, we (1) empirically investigated the
+impact of dataset choice for feature effect estimation. We (2) based our study on
+a theoretical framework that decomposes the feature effect error into different
+components and introduces estimators for each of them. We (3) conducted a
+review of test functions for simulation studies and build the choices for
+our study on these.
+
+Our comprehensive simulation study provided several key insights into
+the choice of dataset for feature effect estimation. First and foremost,
+we find no general evidence that speaks against the current predominant practice
+of using training data for computing feature effects. This is an important validation
+of existing approaches in the literature.
+
+More specifically, our results demonstrated that feature effect errors are
+generally lower when more data is available for the estimation, which is usually
+the case for the training set. This is primarily due to reduced Monte Carlo variance,
+as shown in our ablation studies. The effect is particularly pronounced for ALE plots,
+which exhibit higher sensitivity to dataset size compared to PDPs, especially for smaller samples,
+due to the interval-based estimation. This suggests that for ALE plots, it is even more
+important to use a sufficently large dataset for accurate estimation. Based on these
+findings, we generally recommend using the largest available dataset for feature effect
+estimation, as long as this is computationally feasible.
+
+An interesting finding emerged regarding cross-validation for feature effect estimation.
+Our main study suggests that cross-validation is preferable in many scenarios,
+leading to lower overall errors. The ablation study reveals that this improvement stems from two distinct mechanisms:
+(1) reduced Monte Carlo variance, and (2) reduced model variance through the averaging of estimates from different models.
+This dual benefit is particularly advantageous for overfitting models, which typically suffer
+from higher variance. Therefore, we recommend using cross-validation for feature effect estimation
+when overfitting is a serious concern (e.g., due to a large difference between training
+and validation performance). Estimating feature effects for a single model and dataset
+or using cross-validation also becomes a trade-off between computation time and accurate
+feature effect estimation, as cross-validation comes with computational overhead depending
+on the number of folds.
+
+We also observed that the correlation structure between features appears to have limited impact on our findings
+and are largely consistent across our set of different, diverse datasets.
+This suggests that our recommendations are robust across different data scenarios,
+including those with complex feature interactions, correlations, 
+and real-world inspired relationships as in the Feynman dataset.
+
 \subsection{Limitations \& Future Work}
 
+While our study provides valuable insights, several limitations and opportunities
+for future research remain. One methodological limitation lies in our choice of using
+theoretical quantiles for the grid points in feature effect estimation to make feature
+effects comparable. However, this can lead to empty or sparsely populated intervals, particularly for smaller
+datasets or in regions of low data density. Moreover, this approach has limited real-world
+applicability, as we usually do not now the theoretical but only the empirical quantiles
+(which should be the same in expectation).
+
+Furthermore, the observed instabilities in variance estimation for the Monte Carlo variance,
+particularly of ALE plots, indicate room for methodological improvements. Future research may
+focus on developing more robust estimation procedures for decomposing variance components,
+before applying these estimators to address other research questions.
+
+Another limitation of our study lies in the pre-selection of (optimal) model hyperparameters.
+In practice, these parameters need to be tuned typically using the same dataset as for feature
+effect estimation. This introduces a potential post-selection bias requiring for more sophisticated
+methods, especially when estimating the feature effects in a cross-validated manner, such as
+the use of nested cross-validation.
+
+Further, it is important to note that our study primarily adopts a ``true to the data'' perspective
+when defining feature effect errors, focusing on explaining the underlying data-generating process
+rather than the model's internal representation. While this choice aligns with many practical
+applications, there are also use cases for which ``true to the model'' interpretations might be 
+preferred, which is not considered in our analysis.
+% also: data choice only with regard to feature effect error, not uncertainty estimation
+
+Future research could also investigate the impact of data shifts between training and validation sets.
+While our focused considered overfitting models, future work could explore the impact of data shifts
+between training and validation sets and their impact on feature effect estimation, analyzing the
+robustness of the estimation under distribution shifts.
+
+Moreover, while our cross-validation strategy showed promising improvements over
+conventional estimation on a single dataset, this approach could be enhanced further.
+Future research could explore strategies that utilize more data per fold, such as using the
+training set in each fold for feature effect estimation (and use the validation part only for
+perfomance estimation) or combining training and validation splits. Alternatively, bootstrapping
+strategies similar to the approaches pursued by Molnar et al.~\cite{molnar_relating_2023} could be
+explored.
+
+Following the observations of Loecher~\cite{loecher_debiasing_2024} that hat MDI and SHAP show
+a bias toward high-entropy variables for Random Forests, future work might also investigate if and how
+feature entropy and cardinality influence feature effect estimation.
+
+While we aimed to incorporate real-world relevance through physics-inspired datasets, further
+work is needed to fully capture the complexity of practical applications. In practice, datasets
+have features with varying and also just partial effect strengths, different noise levels,
+higher-order interactions, and diverse marginal distributions. Future work could focus on
+developing more realistic datasets capturing these aspects, to further bridge the gap between
+research, such as simulation studies, and practical applications.
+
+These limitations and future directions highlight the rich opportunities for further research
+in this area to obtain practically relevant insights.
+
 \begin{credits}
-    \subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is used for
-    general acknowledgments, for example: This study was funded by X (grant number
-    Y).
-
+    % \subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is used for
+    % general acknowledgments, for example: This study was funded by X (grant number
+    % Y).
     \subsubsection{\discintname}

### 2025-02-16 current_research_feature_effects/plotting/plots.py
Add plot for ground truth feature effects

@@ -5,14 +5,9 @@ from pathlib import Path
 import pandas as pd
 import numpy as np
 import matplotlib.pyplot as plt
-
-# from matplotlib import transforms
 import matplotlib.colors as colors
 import matplotlib.cm as cm
 import seaborn as sns
-
-# from sklearn.base import BaseEstimator
-# from scipy.stats import pearsonr, spearmanr
 
 from current_research_feature_effects.feature_effects import FeatureEffect, compute_theoretical_effects
 from current_research_feature_effects.data_generating.data_generation import Groundtruth
@@ -21,8 +16,6 @@ from current_research_feature_effects.plotting.utils import (
     get_boxplot_style,
     get_feature_effect_plot_style,
 )
-
-# from current_research_feature_effects.feature_effects import compute_pdps, compute_ales
 
 
 def plot_feature_effect(effect: FeatureEffect | List[FeatureEffect], labels: List[str] = None) -> plt.Figure:
@@ -67,41 +60,71 @@ def plot_feature_effect(effect: FeatureEffect | List[FeatureEffect], labels: Lis
     return fig
 
 
-def plot_groundtruth(groundtruth: Groundtruth, grid_points: Dict[str, np.ndarray], effect_type: Literal["pdp", "ale"]):
+def plot_feature_effect_vs_groundtruth(
+    effect: FeatureEffect,
+    groundtruth: FeatureEffect,
+    feature_names: List[str],
+    large_font: bool = False,
+    save_figs: None | Path = None,
+) -> plt.Figure:
     """
-    Plot the theoretical groundtruth feature effects.
+    Plot feature effects against groundtruth for a given feature effect object.
 
     Parameters
     ----------
-    groundtruth : Groundtruth
-        The Groundtruth object to be plotted.
-    grid_points : Dict[str, np.ndarray]
-        A dictionary containing the grid points for each feature.
-    effect_type : {'pdp', 'ale'}
-        The type of effect to plot, either 'pdp' or 'ale'.
+    effect : FeatureEffect
+        The FeatureEffect object to be plotted.
+    groundtruth : FeatureEffect
+        The FeatureEffect object representing the groundtruth.
+    feature_names : List[str]
+        A list of feature names to plot.
+    large_font : bool, optional
+        If True, use larger font sizes (default is False).
+    save_figs : None | Path
+        Path to save the figure (default is None).
 
     Returns
     -------
     plt.Figure
[…]

### 2025-02-16 paper/paper.tex
paper: add groundtruth feature effects appendix

@@
 Act~\cite{euaia2024}. To address these challenges, the field of Explainable AI
-/ Interpretable ML has emerged~\cite{adadi_peeking_2018}, develping a
+/ Interpretable ML has emerged~\cite{adadi_peeking_2018}, developing a
 wide variety of methods to explain ML models. This includes both model-specific and
@@
 when estimated on training data. In addition, larger datasets may improve the
-feature effect estimates, but also increase computation times.\\
+feature effect estimates but also increase computation times.\\
 
@@
 understand potential implications of their choices. Additionally, our framework
-for evaluating feature effects and systematic collection of test functions
+for evaluating feature effects and our systematic collection of test functions
 provide a foundation for future research in Interpretable ML, with the latter
@@
 \clearpage
+\section{Derivation and Visualization of Analytical Ground Truth Feature Effects}\label{app:ground-truth-feature-effects}
+In the following, we derive the analytical ground truth PDPs and ALEs for the
+considered datasets. In addition, we provide visualizations of the derived
+feature effects and compare them to the estimated feature effects as gut check.
+
+\subsection{SimpleNormalCorrelated Dataset}
+
+\begin{figure*}[h!]
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/SNC-all/pdp_groundtruth_comparison.png}
+        \caption{PDP ground truth of SimpleNormalCorrelated}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/SNC-all/ale_groundtruth_comparison.png}
+        \caption{ALE ground truth of SimpleNormalCorrelated}
+    \end{subfigure}
+    \caption{Ground truth feature effects for SimpleNormalCorrelated dataset. The green line
+    represents the analytically derived ground truth, the orange line the estimated feature
+    effect on the ground truth function using 10000 Monte Carlo samples. Note that both
+    are centered that the average effect at the quantile-based grid points is 0.
+    Since $x_3$ and $x_4$, are dummy variables, they have a constant effect of 0 and are
+    thus omitted.}
+    \label{fig:snc-groundtruth}  % chktex 24
+\end{figure*}
+
+\noindent For SimpleNormalCorrelated, we use normally distributed features with $\mu_1 = 0$,
+$\mu_2 = 0$, $\sigma_1 = 1$, $\sigma_2 = 1$, and correlation
+$\rho = 0.9$. We denote
+arbitrary constants with $C$ and $C^\prime$, respectively.
+
+\begin{align*}
+    PDP_{f,1}(x_1) &= \mathbb{E}_{X_2}[x_1 + \frac{1}{2}X_2^2 + x_1 X_2] \\
+                 &= x_1 + \frac{1}{2} \mathbb{E}_{X_2}[X_2^2] + x_1 \mathbb{E}_{X_2}[X_2] \\
+                 &= x_1 + \frac{1}{2} (Var_{X_2}(X_2) + \mathbb{E}_{X_2}[X_2]^2) + x_1 \mathbb{E}_{X_2}[X_2] \\
+                 &= x_1 + \frac{1}{2} (\sigma_2^2 + \mu_2^2) + x_1 \mu_2 \\
+                 &= x_1 + \frac{1}{2} \\
+                 &= x_1 + C_1
+\end{align*}
+
+\begin{align*}
+    PDP_{f,2}(x_2) &= \mathbb{E}_{X_1}[X_1 + \frac{1}{2}x_2^2 + X_1 x_2] \\
+                    &= \mathbb{E}_{X_1}[X_1] + \frac{1}{2}x_2^2 + \mathbb{E}_{X_1}[X_1] x_2 \\
+                    &= \mu_1 + \frac{1}{2}x_2^2 + \mu_1 x_2 \\
+                    &= \frac{1}{2}x_2^2
+\end{align*}
+
+\begin{align*}
+    ALE_{f,1}(x_1) &= \int_{x_{min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[\frac{\partial}{\partial X_1} X_1 + 0.5 X_2^2 + X_1X_2 \Big| X_1=z_1\right] dz_1 - C_1 \\
+    &= \int_{x_{min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[1 + X_2 \Big| X_1=z_1\right] dz_1 - C_1 \\
+    &(\text{using } X_2 = \beta_{2|1}(X_1-\mu_1) + \varepsilon_1 + \mu_2\text{, where } \beta_{2|1}=\rho \frac{\sigma_2}{\sigma_1}) \\
+    &= \int_{x_{min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[1 + \beta_{2|1}(X_1-\mu_1) + + \varepsilon_1 + \mu_2 \Big| X_1=z_1\right] dz_1 - C_1 \\
+    &= \int_{x_{min,1}}^{x_1} 1 + \beta_{2|1}(z_1-\mu_1) + \mu_2 \,dz_1 - C_1 \\
+    &= [z_1 + \frac{1}{2}\beta_{2|1}(z_1-\mu_1)^2 + \mu_2 z_1]_{x_{min,1}}^{x_1} - C_1 \\
+    &= x_1 + \frac{\beta_{2|1}}{2}(x_1-\mu_1)^2 + \mu_2 x_1 \\
+    & \quad - x_{min,1} - \frac{\beta_{2|1}}{2}(x_{min,1}-\mu_1)^2 - \mu_2 x_{min,1} - C_1 \\
+    % &= (1 + \mu_2) \cdot (x_1 - x_{min,1}) + \frac{\beta_{2|1}}{2} \cdot ((x_1-\mu_1)^2 - (x_{min,1}-\mu_1)^2) - C \\
+    % &= (1 + \mu_2) \cdot (x_1 - x_{min,1}) + \frac{\beta_{2|1}}{2} \cdot (x_1^2 - 2\mu_1 x_1 - x_{min,1}^2 + 2\mu_1 x_{min,1}) - C \\
+    &= (1 + \mu_2) x_1 - (1 + \mu_2) x_{min,1} + \frac{\beta_{2|1}}{2} x_1^2 - \frac{\beta_{2|1}}{2} x_{min,1}^2 \\
+    & \quad - \mu_1 \beta_{2|1} x_1 - \mu_1 \beta_{2|1}  x_{min,1} - C_1 \\
+    &= \frac{\beta_{2|1}}{2} x_1^2 + (1 + \mu_2 - \mu_1 \beta_{2|1}) x_1 - C_1^\prime \\
+    &= 0.45 x_1^2 + x_1 - C_1^\prime
+\end{align*}
+\\
+
+\noindent With anologous calculations, we get:
+
+\begin{align*}
+    ALE_{f,2}(x_2) &= (\frac{1}{2} + \mu_1 + \frac{\beta_{1|2}}{2}) x_2^2 - \beta_{1|2} \mu_2 x_2 - C_2^\prime \\
+    &= 0.95 x_2^2 - C_2^\prime
+\end{align*}
+
+\subsection{Friedman1 Dataset}
+
+For Friedman1, we use independent uniformly distributed features between 0
+and 1. Thus, all features $S$ have $\mu_S = \mathbb{E}[X_S]=\frac{a+b}{2}=0.5$ and
+$\sigma_S^2 = Var[X]=\frac{(b-a)^2}{12}=\frac{1}{12}$.
+Si denotes the sine integral function, $C_S$ and $C_S^\prime$ are arbitrary constants.
+
+\begin{align*}
+    PDP_{f,1}(x_1) &= \mathbb{E}_{-X_1}[10\sin(\pi x_1X_2) + 20(X_3-\frac{1}{2})^2 + 10X_4 + 5X_5] \\
+    &= 10 \mathbb{E}_{X_2}[\sin(\pi x_1X_2)] + 20(\mathbb{E}_{X_3}[X_3^2] - \mathbb{E}_{X_3}[X_3] + \frac{1}{4}) + 10 \mathbb{E}_{X_4}[X_4] \\
+    &\quad + 5 \mathbb{E}_{X_5}[X_5]\\
+    &= 10 \cdot \int_{a}^{b} \sin(\pi x_1x_2)dx_2 + 20\cdot(\mu_3^2 + \sigma_3^2 - \mu_3 + \frac{1}{4}) + 10\mu_4 + 5\mu_5 \\
+    &= 10 \cdot \frac{-\cos(\pi bx_1) + \cos(\pi ax_1)}{\pi x_1} + 20(\mu_3^2 + \sigma_3^2 - \mu_3) + 5 + 10\mu_4 + 5\mu_5 \\
+    &= 10 \cdot \frac{-\cos(\pi x_1) + 1}{\pi x_1} + C_1 \\
+\end{align*}
+
+\noindent Analogously, we get for $x_2$:
+
+\begin{align*}
+    PDP_{f,2}(x_2) &= 10 \cdot \frac{-\cos(\pi x_2) + 1}{\pi x_2} + C_2
+\end{align*}
+
+\begin{align*}
+    PDP_{f,3}(x_3) &= \mathbb{E}_{-X_3}[10\sin(\pi X_1X_2) + 20(x_3-\frac{1}{2})^2 + 10X_4 + 5X_4] \\
+    &= 10\int_{a_1}^{b_1}\int_{a_2}^{b_2}\sin(\pi x_1x_2)dx_2dx_1 + 20(x_3-\frac{1}{2})^2 + 10\mu_4 + 5\mu_5 \\
+    &= \frac{10}{\pi}[Si(\pi b_1a_2) - Si(\pi a_1a_2) - Si(\pi b_1b_2) + Si(\pi a_1b_2)] \\
+    &\quad + 20x_3^2 - 20x_3 + 5 + 10\mu_4 + 5\mu_5 \\
+    &= 20x_3^2 - 20x_3 + C_3 \\
+\end{align*}
+
+\noindent Since Si is not defined at 0, we use a small $\epsilon$ for
+computational stability.\newline
+Analogously, we get for $x_4$, $x_5$:
+
+\begin{align*}
+    PDP_{f,4}(x_4) &= 10x_4 + C_4
+\end{align*}
+
+\begin{align*}
+    PDP_{f,2}(x_2) &= 5x_5 + C_5
+\end{align*}
+
+\begin{align*}
+    ALE_{f,1}(x_1) &= \dots
+\end{align*}
+
+\begin{align*}
+    ALE_{f,2}(x_2) &= \dots
+\end{align*}
+
+\begin{align*}
+    ALE_{f,3}(x_3) &= \int_{x_{3,\min}}^{x_3} \mathbb{E}_{-X_3|X_3}\Bigg[\frac{\partial}{\partial x_3}(10\sin(\pi X_1X_2) + 20(X_3-\frac{1}{2})^2 + 10X_4 \Bigg. \\
+    &\qquad\qquad\qquad\qquad\qquad \Bigg. + 5X_5) \big|X_3=z_3\Bigg]dz_3 - C_3 \\
+    &= \int_{x_{3,\min}}^{x_3} \mathbb{E}_{-X_3|X_3}\left[40z_3-20\right]dz_3 - C_3 \\
+    &= \int_{x_{3,\min}}^{x_3} 40z_3-20dz_3 - C_3 \\
+    &= \left[20z_3^2 - 20z_3\right]_{x_{3,\min}}^{x_3} - C_3 \\
+    &= 20x_3^2 - 20x_3 - 20x_{3,\min}^2 + 20x_{3,\min} - C_3 \\
+    &= 20x_3^2 - 20x_3 - C_3^\prime
+\end{align*}
+
+\begin{align*}
+    ALE_{f,4}(x_4) &= \int_{x_{4,\min}}^{x_4} \mathbb{E}_{-X_4|X_4}\Bigg[\frac{\partial}{\partial x_4}(10\sin(\pi X_1X_2) + 20(X_3-\frac{1}{2})^2 + 10X_4 \Bigg. \\
+    &\qquad\qquad\qquad\qquad\qquad \Bigg. {} + 5X_5) \big|X_4=z_4\Bigg]dz_4 - C_4 \\
+    &= \int_{x_{4,\min}}^{x_4} \mathbb{E}_{-x_4|x_4}\left[10\right]dz_4 - C_4 \\
+    &= 10x_4 - 10x_{4,\min} - C_4 \\
+    &= 10x_4 - C_4^\prime
+\end{align*}
+
+\noindent Analogously, we get for $x_5$:
+
+\begin{align*}
+    ALE_{f,5}(x_5) &= 5x_5 - C_5^\prime
+\end{align*}
+
+\begin{figure*}[h!]
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/pdp_groundtruth_comparison.png}
+        \caption{PDP ground truth of Friedman1}
+    \end{subfigure}
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/ale_groundtruth_comparison.png}
+        \caption{ALE ground truth of Friedman1}
+    \end{subfigure}
+    \caption{Ground truth feature effects for Friedman1 dataset. The green line
+    represents the analytically derived ground truth, the orange line the estimated feature
+    effect on the ground truth function using 10000 Monte Carlo samples. Note that both
+    are centered that the average effect at the quantile-based grid points is 0.
+    Since $x_6$ and $x_7$, are dummy variables, they have a constant effect of 0 and are
+    thus omitted.}
+    \label{fig:friedman1-groundtruth}  % chktex 24
+\end{figure*}
+
+\subsection{FeynmanI.29.16 Dataset}
+
+For the FeynmanI.29.16 dataset, it is not possible to derive the ground truth
+feature effects analytically. Nonetheless, we give a visualization of the
+estimated ground truth feature effects for illustration purposes in
+\textbf{Fig.~\ref{fig:feynman-groundtruth}}.
+
+\begin{figure*}[h!]
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/pdp_groundtruth_comparison.png}
+        \caption{PDP ground truth of FeynmanI.29.16}
+    \end{subfigure}
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/FeynmanI.29.16-all/ale_groundtruth_comparison.png}
+        \caption{ALE ground truth of FeynmanI.29.16}
+    \end{subfigure}
+    \caption{Ground truth feature effects for FeynmanI.29.16 dataset. The green line
+    represents the analytically derived ground truth, the orange line the estimated feature
+    effect on the ground truth function using 10000 Monte Carlo samples. Note that both
+    are centered that the average effect at the quantile-based grid points is 0.
+    Since $d_1$ and $d_2$, are dummy variables, they have a constant effect of 0 and are
+    thus omitted.}
+    \label{fig:feynman-groundtruth}  % chktex 24
+\end{figure*}
+
+\clearpage
 \section{Additional Main Experiment Results}\label{app:additional-results}

### 2025-02-19 paper/paper.tex
paper: refine chapters 1-4

@@
 
-In this paper, we focus on PDP and ALE and refer to them when speaking of
+In the following, we focus on PDP and ALE and refer to them when speaking of
 feature effects.
@@
 following, we synthesize commonly used test functions across various domains,
-providing structured guidance for researchers --- particularly those in
-Interpretable ML --- in selecting appropriate test functions for their
-simulation studies. By examining test functions from different fields and their
-purposes of application, we aim to facilitate experimental design decisions
-rather than giving an exhaustive overview.
+providing structured guidance for researchers, particularly those in
+Interpretable ML, in selecting appropriate test functions to facilitate
+designing simulation studies.
 
@@
 continuous optimization algorithms through artificial test
-functions~\cite{hansen_coco_2016}. These classical test function suites are
-well-established in optimization and may also serve as a basis for
-Interpretable ML researchers. However, the ability of these artifical test
-functions to represent complex real-world behavior is often
+functions~\cite{hansen_coco_2016}. While being
+well-established in optimization, they may also serve as a basis for
+Interpretable ML researchers. However, the ability of these artificial test
+functions to represent complex real-world behavior is
 limited~\cite{zaefferer_simulation-based_2017}.
@@
 with the \textit{Feynman Symbolic Regression Database
-    (FSReD)}~\cite{udrescu_ai_2020} being a prominent example. FSReD comprises 100
+    (FSReD)}~\cite{udrescu_ai_2020} as a prominent example. FSReD comprises 100
 physics equations from the seminal \textit{Feynman's Lectures on Physics
@@
 physics texts. These equations span diverse physics domains and involve a
-varying number of variables and various elementary functions such as arithmetic
-operations, trigonometric functions, and exponentials. Tabular datasets are
+varying number of variables and various elementary functions. Tabular datasets are
 generated through random sampling from defined value ranges.
@@
 Matsubara~et~al.~\cite{matsubara_rethinking_2024} addressed several limitations
-of the original FSReD. They introduce a three-tiered categorization of problems
-(easy, medium, hard) based on their complexity, incorporate dummy variables to
-simulate irrelevant features, and implement more realistic sampling ranges and
+of the original FSReD by introducing a three-tiered categorization of problems
+based on their complexity, incorporating dummy variables to
+simulate irrelevant features, and implementing more realistic sampling ranges and
 strategies. Detailed specifications for all formulas, including their sampling
@@
 
-While initially developed for symbolic regression tasks, many Feynman equations
-may serve as suitable test functions for simulation studies in Interpretable
+While originally designed for symbolic regression tasks, many Feynman equations
+serve as suitable test functions for simulation studies in Interpretable
 ML. Their basis in physical principles provides real-world relevance, though  % chktex 13
@@
 functions designed to evaluate specific aspects of interpretability methods.
-
 Goldstein et al.~\cite{goldstein_peeking_2015} used several simple test
@@
 absence of interactions, simple interactions to reveal heterogeneity that might
-be obscured by averaging procedures such as PDPs, and a specially designed
+be obscured by PDPs, and a specially designed
 function with an empty quadrant for assessing extrapolation behavior.
 
-Similarly, Liu et al.~\cite{liu_model_2018} focus on simple functions before
-progressing to more complex ones. They begin with basic two-variable scenarios
---- using additive functions, interaction functions, and combinations thereof
---- and examine these under both independent and correlated feature conditions
-to compare various feature effect methods. The advantage of these simple test
-functions is that solutions (e.g., feature effects) can also be computed
-analytically, and that they allow for deeper and more fine-grained analysis of
+Similarly, Liu et al.~\cite{liu_model_2018} focused on simple functions before
+increasing complexity. They started with basic two-variable scenarios
+--- additive functions, interaction functions, and combinations thereof
+--- and examined these under both independent and correlated features,
+comparing various feature effect methods. Advantages of these simple test
+functions are that solutions (e.g., feature effects) can be computed
+analytically, and that they allow for more fine-grained analyses of
 individual aspects.
@@
 of interactions with different orders, strengths, non-linearities, and
-overlaps. While this makes them particularly valuable for interaction
+overlaps. While being particularly valuable for interaction 
 detection, they are also useful for evaluating other interpretability methods
@@
 features or interactions, the need for analytical solutions for validation, as
-well as the relevance to real-world applications in the domain of interest.
-
-\section{Background}\label{sec:background}
+well as the relevance to real-world applications.
+
+\section{Theoretical Framework: Feature Effects and Error Decomposition}\label{sec:background}
 
@@
 
-The \textit{Partial Dependence Plot (PDP)} by
-Friedman~\cite{friedman_greedy_2001} describes the marginal effect of one or
+The \textit{Partial Dependence Plot (PDP)}~\cite{friedman_greedy_2001} describes the marginal effect of one or
 two features on the prediction of a model $\hat f$. For a feature set $X_S$
@@
 \begin{equation}
-    PDP_{\hat f, S} = \mathbb{E}_{X_C}[\hat{f}(x_S, X_C)] = \int f(x_S, x_C)d\mathbb{P}(x_C),
+    PDP_{\hat f, S} = \mathbb{E}_{X_C}[\hat{f}(x_S, X_C)] = \int \hat f(x_S, x_C)d\mathbb{P}(x_C),
 \end{equation}
@@
 \noindent Here, $x_C^{(i)}$ are the actual complement feature values from the dataset of $n$ instances.
-To plot this function, a grid of $G$ grid points
-$\{(x_S^{(g)}, \widehat{PD}_{\hat f, S}(x_S^{(g)}))\}_{g=1}^G$ can be used~\cite{molnar_relating_2023}.  % chktex 3
+To plot the function, a grid of $G$ grid points
+$\{(x_S^{(g)}, \widehat{PDP}_{\hat f, S}(x_S^{(g)}))\}_{g=1}^G$ can be used~\cite{molnar_relating_2023}.  % chktex 3
 Molnar et al.~\cite{molnar_general_2022} recommend using quantile-based over equidistant grids.\\
 
-\noindent The \textit{Accumulated Local Effects (ALE)} plot is an alternative to the PDP that
-solves the extrapolation issue~\cite{apley_visualizing_2020}. Using the notation above,
-for $|S|=1$, the ALE plot is defined as
+\newpage
+\noindent The \textit{Accumulated Local Effects (ALE)} plot~\cite{apley_visualizing_2020}
+for $|S|=1$ is defined as
 
@@
 \noindent where $\hat{f}^S(x_S, x_C) = \frac{\partial \hat{f}(x_S, x_C)}{\partial x_S}$.
-The constant is chosen so that $\widehat{ALE}_{\hat f,S}(X_S)$ is centered with a mean of $0$
+The constant is chosen so that $ALE_{\hat f,S}(X_S)$ is centered with a mean of $0$
 w.r.t.\ the marginal distribution of $X_S$. The uncentered ALE can be estimated by
@@
 $\{x^{(i)}_S\}_{i=1}^n$ into $K$ intervals or neighborhoods $N_S(k)$. $n_S(k)$ denotes  % chktex 3
-the number of observations in the $k$th interval $N_S(k)$, $k_S(x)$ represents the index
-of the interval to which a particular value $x$ of feature $x_S$ belongs.
+the number of observations in the $k$th interval, $k_S(x)$ the index
+of the interval to which a particular value $x$ of feature $S$ belongs.
 The uncentered ALE is centered by
@@
 the quantiles of the empirical distribution of $\{x^{(i)}_S\}_{i=1}^n$ can be  % chktex 3
-used~\cite{apley_visualizing_2020}.\\
+used~\cite{apley_visualizing_2020}.
 
@@
 
-To quantify the error of a computed feature effect, a ``ground truth'' needs to
+\paragraph{Feature effect ground truth.} To quantify the error of a computed
+feature effect, a ``ground truth'' needs to
 be defined first. We follow the approach of Molnar et
@@
 underlying ground truth function $f$ (instead of model $\hat f$).
-
 For PDP, we can directly use the definition of Molnar et
@@
 
-As stated by Molnar et al.~\cite{molnar_relating_2023}, their results also
+\noindent As stated by Molnar et al.~\cite{molnar_relating_2023}, their results also
 apply to conditional variants of the PDP such as ALE. We now make this  % chktex 13
-definition explicit:
+definition explicit.
+
+\newpage
 
@@
 %(cf.~\cite{molnar_relating_2023}). 
-Note that different ground truth effects may also be defined, and our choices
+\noindent Note that different ground truth feature effects can also be defined, and our choices
 come with certain implications and limitations, such as omitting the
-\textit{aggregation bias}\footnote{for details,
-    see~\cite{herbinger_repid_2022}}~\cite{mehrabi_survey_2021}.
-
-With more complex ground truth functions $f$, it may become increasingly
-difficult to derive the ground truth feature effects analytically, especially
-for ALE. In these cases, we therefore propose to also estimate those effects by  % chktex 13
+\textit{aggregation bias}~\cite{herbinger_repid_2022,mehrabi_survey_2021}.
+
+With more complex ground truth functions $f$, it becomes often infeasible to
+derive the ground truth feature effects analytically. For these cases, we
+therefore propose to also estimate those effects by  % chktex 13
 Monte Carlo integration, yielding $\widehat{PDP}_{f,S}(x_S)$ and
-$\widehat{ALE}_{f,S}(x_S)$ (obtained by plugging in $f$ instead of $\hat f$
+$\widehat{ALE}_{f,S}(x_S)$ (obtained by plugging $f$ instead of $\hat f$
 into the estimators in equations (\ref{eq:pdp-estimate}) and
-(\ref{eq:ale-estimate-uncentered}) / (\ref{eq:ale-estimate-centered})).\\
-
-\noindent Summarizing, we have now defined four quantities per feature effect:
+(\ref{eq:ale-estimate-uncentered}) / (\ref{eq:ale-estimate-centered})).
+
+\paragraph{Feature effect error decomposition.} Summarizing, we have now defined four quantities per feature effect:
 $PDP_{f,S}$ $\widehat{PDP}_{f,S}$, $PDP_{\hat f,S}$, and $\widehat{PDP}_{\hat f,S}$
-(analogue for ALE). We can now define different errors between each of these
-quantities. In this paper, we focus on the MSE, as it can be decomposed into
-bias and variance (see e.g.~\cite{geman_neural_1992}).
-Taking, for example, $PDP_{f,S}$ as ground truth, we can define the MSE of
-$PDP_{\hat f,S}$ at a point $x_S$ as follows~\cite{molnar_relating_2023}:
+(analogously for ALE). Between these quantities, different error metrics can because
+defined. In this paper, we focus on MSE as it can be decomposed into
+bias and variance~\cite{geman_neural_1992}.
+When taking $PDP_{f,S}$ as ground truth, the MSE of $PDP_{\hat f,S}$ at
+point $x_S$ can be formulated as follows~\cite{molnar_relating_2023}:
 
@@
 Since we usually cannot determine $PDP_{\hat f,S}$, we need to estimate it by
-Monte Carlo integration, yielding $\widehat{PDP}_{\hat f,S}(x_S)$. This,
-however, introduces an additional variance term (we use the random variable
-$X_{mc}$ for the Monte Carlo samples (e.g., training or validation data)):
+Monte Carlo integration, which introduces an additional variance term (we use the random variable
+$X$ for the Monte Carlo samples):
 
@@
         = \underbrace{{(PDP_{f,S}(x_S) - \mathbb{E}_F[PDP_{\hat{f}, S}(x_S)])}^2}_{Bias^2} \\
-        + \underbrace{\text{Var}_F[PDP_{\hat{f},S}(x_S)]}_{Variance} + \underbrace{\mathbb{E}_F\text{Var}_{X_{mc}}[\widehat{PDP}_{\hat{f},S}(x_S)]}_{MC-Variance}
+        + \underbrace{\text{Var}_F[PDP_{\hat{f},S}(x_S)]}_{Variance} + \underbrace{\mathbb{E}_F\text{Var}_{X}[\widehat{PDP}_{\hat{f},S}(x_S)]}_{MC-Variance}
     \end{split}
+    \label{eq:mse-decomposition}
 \end{equation}
@@
 distribution $F$. Similarly, one could use the estimate $\widehat{PDP}_{f,S}$
-as groundtruth, introducing an additional variance term $Var_{X_{mc2}}$when
-estimated on different MC sample (proof in
+as ground truth, introducing an additional variance term when
+estimated with Monte Carlo samples (proof in
 \textsc{Appendix~\ref{app:proof-mc-variance}}).
-An overview of all error terms in this chain can be found in
-\textbf{Fig.\@~\ref{fig:error-graph}}.\\
+An overview of all error terms is given in
+\textbf{Fig.\@~\ref{fig:error-graph}}.
 
@@
     \includegraphics[width=\textwidth]{img/error_graph.pdf}
-    \caption{Error chain in feature effect estimation
-        (modified, original version can be found in~\cite{molnar_relating_2023})}\label{fig:error-graph}
+    \caption{Error chain in feature effect estimation. The deviation of the estimated
+        feature effect from the ground truth can be decomposed into bias, model variance,
+        and Monte Carlo variance. If the ground truth is estimated by Monte Carlo
+        integration, an additional Monte Carlo variance term arises. This graphic
+        builds on~\cite{molnar_relating_2023}.}\label{fig:error-graph}
 \end{figure}
 
-\noindent $\mathbb{E}_F$ and $\text{Var}_F$ could be estimated by averaging over
+\paragraph{Feature effect error estimation.} Next, we aim to estimate these error terms.
+$\mathbb{E}_F$ and $\text{Var}_F$ can be estimated by averaging over
 multiple models of the same inducer fitted to $M$ different training data
-sets sampled independently from the DGP. We propose the following estimators:  % chktex 13
+sets, each independently sampled from the DGP. We propose the following estimators:  % chktex 13
+
 \begin{equation}
@@
 \begin{equation}
-    \widehat{\text{Variance}}(x_S; \widehat{PDP}_{\hat f,S}) =
+    \begin{split}
+    \widehat{\text{Variance}}(x_S; \widehat{PDP}_{\hat f,S}) = \\
     \frac{1}{M-1}\sum_{m=1}^M{\left(\widehat{PDP}_{\hat f^{(m)},S}(x_S) - \frac{1}{M}\sum_{m=1}^M \widehat{PDP}_{\hat f^{(m)},S}(x_S)\right)}^2
+    \end{split}
     \label{eq:variance-estimator}
@@
 
-\noindent Note that these are similar to the approach in~\cite{molnar_relating_2023},
-but we do not specify which data points to use for MC integration. The
-variance captures both the variance in the model fits and the variance due to
-MC integration. To estimate the MC variance, we propose the following estimators:
+\noindent The variance estimator is the same as in~\cite{molnar_relating_2023},
+capturing both the variance in the model fits and the variance due to
+MC integration. For the Monte Carlo variance of estimating the ground truth by
+Monte Carlo integration, i.e.\ omitting the model, we propose the following estimator:
 \begin{equation}
@@
 
-\noindent and
+\noindent To estimate the Monte Carlo variance for feature effects of models
+given in equation~\ref{eq:mse-decomposition}, we have to consider multiple
+Monte Carlo iterations for multiple model refits, yielding the following estimator:
+
 \begin{equation}
@@
 
-\noindent For more convenient analysis of the errors, one could also
-aggregate them over the marginal distribution of $X_S$ (e.g.,
-estimated by averaging over the grid points if chosen appropriately)
-to obtain a single error measure per feature effect.
-
-While our definitions are based on the PDP, they can be directly applied to the
-ALE as well.
+\noindent While these estimators are based on the PDP, they
+can be defined analogously for ALE.  % chktex 13
+% For a more convenient analysis of the errors, we suggest
+% an aggregation of the error terms over the marginal distribution of $X_S$
+% (e.g., by averaging over the grid points if chosen appropriately)
+% to obtain a single error measure per feature effect.
+It is important to note that most of these estimators are rather
+impractical as they require knowledge of the ground truth. Nonetheless,
+we rather use them to understand the behavior of the feature
+effect error components under different dataset choices. This
+distinguishes our approach from previous works such as~\cite{molnar_relating_2023},
+which mostly focus on providing practical variance estimates to provide
+confidence intervals for uncertainty quantification in feature effects.
 
@@
 
-To address the research question of which data to choose to estimate feature
-effects, we conduct a comprehensive simulation study. For different models,
+To address the central research question of this paper, which data to
+choose for feature effect estimation, we conduct a comprehensive simulation
+study. For different models,
 datasets, and dataset sizes, we estimate the feature effects PDP and ALE on
 training data, validation data, and in a cross-validated manner. We then
-compute the feature effect error as MSE with respect to the ground truth
-feature effects and decompose it into bias and variance components. By
+compute the feature effect error as MSE against the ground truth
+feature effects, and decompose it into bias and variance components. By
 comparing these error terms across the different estimation strategies
@@
 insights into the implications of the data choice on feature effect estimation.
-For deeper analysis, we conduct two ablation studies: one decomposing the
-feature effect variance into model variance and Monte Carlo variance, and
-another examining the impact of the dataset size on the Monte Carlo variance.
+To deepen our analysis, we conduct two ablation studies: (1)~decomposing feature
+effect variance into model variance and Monte Carlo variance, and (2)~examining
+the influence of dataset size on Monte Carlo variance.
 In the following, we describe our experimental set-up for the main simulation
@@
           standard-normally distributed features, where the first two
-          features exhibit strong correlation ($\rho = 0.9$) while the others are
-          independent dummy variables. The target variable is given by this simple
-          formula:
+          are strongly correlated ($\rho = 0.9$) while the others are
+          independent dummy variables. The target variable is given by:
           \begin{equation}
@@
           \end{equation}
-          This test function is inspired by~\cite{liu_model_2018} and aims
-          to focus on the impact of correlation and interactions.
+          This test function, inspired by~\cite{liu_model_2018}, focuses
+          on investigating the impact of correlation and interactions.
     \item \textbf{Friedman1} implements the classical Friedman1 benchmark
           function~\cite{breiman_bagging_1996,friedman_multivariate_1991}
-          with seven uniformly distributed features between 0 and 1, all mutually
-          independent:
+          with seven mutually independent, uniformly distributed features
+          between 0 and 1:
           \begin{equation}
@@
           \end{equation}
-          This dataset includes a mix of linear and different non-linear effects with
+          This dataset includes a mix of linear and non-linear effects with
           interactions.
     \item \textbf{Feynman I.29.16} is based on the Feynman equation I.29.16
-          describing wave interference, comprising six independent features. Using
+          describing wave interference, comprising six independent features. Following
           the refined sampling strategies from Matsubara et al.~\cite{matsubara_rethinking_2024},
-          two log-uniformly distributed variables on $[0.1, 10]$, two angles uniformly
+          we use two log-uniformly distributed variables on $[0.1, 10]$, two angles uniformly
           distributed over $[0, 2\pi]$, and two uniformly distributed dummy variables on [0,1]:
@@
           \end{equation}
-          This dataset is designed to reflect a physics-based relationship for more real-world relevance.
+          This dataset is designed to reflect a physics-based relationship for enhanced real-world relevance.
 
@@
 
-\noindent To generate the datasets, a standard normally distributed noise term $\epsilon$ is added to each function,
-scaled by a signal-to-noise ratio of five\footnote{To determine the factor by which the noise is multiplied,
-    the standard deviation of the signal is computed over 100'000 randomly drawn samples of $y$ and divided by
-    the signal-to-noise ratio of five.}. For each test function, we consider two dataset sizes:
+\noindent To generate the datasets, we add a standard-normally distributed noise term $\epsilon$ to each function,
+scaled by a signal-to-noise ratio of five\footnote{To determine the noise scaling factor,
+    we compute the standard deviation of the signal over 100,000 randomly drawn samples of $y$ and
+    divide by the signal-to-noise ratio of five.}. For each test function, we consider two dataset sizes:
 $(n_{train}=1000, n_{val}=250)$ and $(n_{train}=8000, n_{val}=2000)$.
@@
 each fold fit the model on four folds, and estimate the feature effects on the
-remaining fold. We then average the feature effects over all folds. We also
-estimate the ground truth feature effects $\widehat{PDP}_{f,S}$ and
-$\widehat{ALE}_{f,S}$ since the complexity of the Feynman equation makes
-analytical computation infeasible. For the Monte Carlo integration, we use
-10000 samples additionally sampled from the DGP, which introduces only a
-negligible additional variance term as we show in (\dots) We use 100 grid
+remaining fold. We then average the feature effects point-wise over all folds.
+
+We also estimate the ground truth feature effects $\widehat{PDP}_{f,S}$ and
+$\widehat{ALE}_{f,S}$ as the complexity of the Feynman equation makes
+an analytical computation infeasible. For their Monte Carlo integration, we use
+10,000 samples additionally drawn from the DGP, which introduces only a
+negligible additional variance as we show in (Section~\ref{sec:results-dataset-size}). We use 100 grid
 points, defined by the quantiles of the theoretical distribution of $X_S$ to
@@
 omit the first and the last grid point after estimation to avoid boundary
-effects especially occuring in ALE plots.
+effects particularly occuring in ALE plots.
 
 \paragraph{Feature effect errors.} To quantify the error of the estimated feature effects, we compute the MSE of
-the estimated model feature effects with respect to the estimated ground truth
-features effects, as well as their bias and variance. For that, we use the
+the estimated model feature effects against the estimated ground truth effects,
+as well as their bias and variance. We use the
 estimators defined in equations~(\ref{eq:mse-estimator}) to
@@
 DGP, refitting the models, and estimating the feature effects. To aggregate the
-errors into a single measure per feature effect, we average the errors over the
+errors into a single number per feature effect, we average them over the
 marginal distribution of $X_S$ by averaging over the grid points.
@@
 
-\paragraph{Variance decomposition study.} The variance of the feature effects estimated with the procedure described in
-the main experiment \textsc{Section~\ref{sec:experiment-main}} consists of two
-components: the model variance and the variance due to the Monte Carlo
-integration. To disentangle these two sources of variance, we aim to
-additionally estimate the Monte Carlo variance using the estimator in
-equation~(\ref{eq:mc-variance-estimator-model}). This allows us to substract it
-from the total variance estimated in the main experiment and obtain the model
-variance, giving more nuanced insights into the implications of the data choice
-on feature effect estimation. To obtain this estimate, we maintain the set-up
-of the main experiment but for each trained model $\hat f^{(m)}$ (plus five
-models from CV $\hat f^{(m,1)} - \hat f^{(m,5)}$), we draw $K=30$ new training
-and validation sets from the DGP and estimate the feature effects on these sets
-plus in a cross-validated manner with the already fitted models. This allows us
-to estimate the Monte Carlo variance as in
-equation~(\ref{eq:mc-variance-estimator-model}). Since this study is
-computationally more expensive, we do this only for XGBoost (OT and OF), which
-is arguably the most practically relevant model in our set-up. We use the same
-datasets and sizes as in the main experiment.
-
-\paragraph{Impact of dataset size on Monte Carlo variance.} To investigate the impact of the dataset size on the Monte Carlo variance, we
-estimate the Monte Carlo variance between the analytical ground truth feature
-effects and the estimated ground truth feature effects on different dataset
-sizes. As estimator for the Monte Carlo variance, we use
-equation~(\ref{eq:mc-variance-estimator-groundtruth}). With this approach, we
-entirely omit the model and thus eliminate other error sources such as model
-bias and variance (as with a perfect model fit). To estimate the ground truth
-feature effects, we use 50 different dataset sizes ranging from $10^1$ to
-$10^6$ on a logarithmic scale. Note that we only consider the
-SimpleNormalCorrelated and Friedman1 datasets for this study as an analytical
-computation of the FeynmanI.29.16 feature effects is infeasible. For other
-details, we maintain the set-up of the main experiment.
+\paragraph{Variance decomposition study.} The variance of the feature effects
+estimated in the main experiment \textsc{Section~\ref{sec:experiment-main}} consists of two
+components: model variance and variance due to the Monte Carlo
+integration. To disentangle these two sources of variance, we estimate the
+Monte Carlo variance using the estimator in
+equation~(\ref{eq:mc-variance-estimator-model}). By subtracting this from the
+total variance obtained in the main experiment, we derive the model variance,
+providing more nuanced insights into the implications of data choice in feature
+effect estimation.
+
+For this analysis, we maintain the set-up of the main experiment with one key
+modification: for each trained model $\hat f^{(m)}$ (plus the five
+models $\hat f^{(m,1)} - \hat f^{(m,5)}$ trained in cross-validation), we draw $K=30$ new training
+and validation sets from the DGP. We then estimate feature effects on these  % chktex 13
+sets and in a cross-validated manner using the already fitted models. This approach
+allows us to estimate the Monte Carlo variance.
+
+Due to the computational intensity of this analysis, we focus only on XGBoost models
+(both OT and OF variants), which represent the arguably most practically relevant models
+in our setup.
+
+\paragraph{Impact of dataset size on Monte Carlo variance.}
+To investigate how dataset size influences Monte Carlo variance,
+we estimate the variance between analytical ground truth feature effects
+and estimated ground truth feature effects across different dataset sizes.
+We use equation~(\ref{eq:mc-variance-estimator-groundtruth}) as estimator
+for the Monte Carlo variance. This approach entirely omits the model, thereby
+eliminating other error sources such as model bias and variance (simulating a perfect model fit)
+
+We analyze 50 different dataset sizes ranging from $10^1$ to $10^6$ on a logarithmic
+scale. We restrict this analysis to the SimpleNormalCorrelated and Friedman1 datasets,
+as analytical computation of FeynmanI.29.16 ground truth feature effects is infeasible.
+All other experimental details follow the setup of the main experiment.
 
@@
 
-\section{Proof: Monte Carlo Variance for Groundtruth}\label{app:proof-mc-variance}
+\section{Proof: Monte Carlo Variance for Ground truth}\label{app:proof-mc-variance}
 \begin{proof}

### 2025-02-20 paper/paper.tex
paper: refine chapters 5 & 6 and appendix

@@
 
-\subsection{Main Experiment}
-
-We give the results of the main experiment error decomposition for the
+\subsection{Main Experiment Results}
+
+We present the results of the main experiment error decomposition for the GAM and
+XGBoost models on the
 SimpleNormalCorrelated dataset in \textbf{Fig.\@~\ref{fig:pdp-results-snc}} for
@@
 
-Generally, we observe that the feature effect error is in the most cases
-highest when the effect is estimated on validation data. For ALE, this
-observation is more pronounced than for PDP. This increase in error is mainly  % chktex 13
+Our analysis reveals that feature effect error is generally highest when the effect
+is estimated on validation data. For ALE, this
+observation is more pronounced than for PDP. The increase in error is mainly  % chktex 13
 due to an increase in variance (see e.g.,
@@
 \textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-of}}). For many models, the
-feature effect error is lowest when estimated in a cross-validated manner. This
+feature effect error is lowest when effects are estimated in a cross-validated manner. This
 is slightly more pronounced for overfitting models (OF) than for optimally
-tuned models (OT), and is mainly due to a decrease in variance (see e.g.,
+tuned models (OT), and mainly stems from a reduced variance (see e.g.,
 \textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-of}} and
-\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}}). Since the variance composes
-of the model variance and the Monte Carlo variance, both could be reduced by
-cross-validation. Monte Carlo variance may decrease as we have slightly more
-samples overall compared to the training set. Conversely, the model variance
-may decrease as it is averaged out through the cross-validation procedure,
-which may especially improve overfitting models that have a higher model
-variance. We further investigate these hypotheses in
+\textbf{Fig.\@~\ref{fig:pdp-results-snc-gam-ot}}). Since the variance is composed
+of model variance and Monte Carlo variance, cross-validation could potentially
+reduce each of these components. Monte Carlo variance may decrease due to
+the slightly larger effective sample size compared to using only the training set.
+Conversely, model variance
+may decrease as cross-validation averages out estimation variability across multiple
+models, which is particularly beneficial for overfitting models that typically exhibit
+higher model variance. We further investigate these hypotheses in
 \textsc{Section~\ref{sec:results-variance-decomposition}}.
 
-\begin{figure}[htbp]
+\begin{figure}[ht!]
     \centering
@@
     \end{subfigure}
-    \\[10pt]
-    \vfill
-    \begin{subfigure}[b]{0.49\textwidth}
-        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_pdp_SVM_OT.png}
-        \caption{SVM\_OT}
-        \label{fig:pdp-results-snc-svm-ot}  % chktex 24
-    \end{subfigure}
-    \caption{Main experiment results for PDP on SimpleNormalCorrelated dataset.\dots}
+    \caption{Main experiment results for PDP on SimpleNormalCorrelated dataset for GAM and
+        XGBoost models. The other models as well as the dummy features $x_3$ and $x_4$
+        are omitted for brevity.}
     \label{fig:pdp-results-snc}  % chktex 24
@@
 
-Interestingly, for ALE, we observe for some cases an increased bias when
-estimating the feature effects on validation data and via cross-validation
+\noindent Interestingly, for ALE, we observe for some cases an increased bias when
+estimating the feature effects on validation data or through cross-validation
 compared to an estimation on training data (see e.g.,
-\textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-ot}} and
-\textbf{Fig.\@~\ref{fig:ale-results-snc-svm-ot}}). Note that this is only the
-case when errors are generally low, and the bias is still small in absolute
-terms. Since this only occurs for ALE and only for the smaller dataset size
-(1000 training samples, 250 validation samples), we hypothesize that this is
-due to our approach of choosing the intervals based on theoretical quantiles.
-For the smaller validation sets, we may have intervals with few or no samples,
-leading to a bias in the estimation of the effect. This hypothesis is supported
-by the fact that the increase in bias only occurs in regions with low sample
+\textbf{Fig.\@~\ref{fig:ale-results-snc-gam-ot}} and
+\textbf{Fig.\@~\ref{fig:ale-results-snc-xgboost-ot}}). However, this phenomenon
+occurs only when errors are generally low, and the bias is still small in absolute
+terms. Since this appears exclusively for ALE and only with smaller dataset sizes
+(1000 training samples, 250 validation samples), we hypothesize it stems from our
+approach of defining intervals based on theoretical quantiles.
+For smaller validation sets, this may result in intervals with few or no samples,
+introducing a bias in the estimation of the effect. This hypothesis is supported
+by the fact that the increased bias only occurs in regions with low sample
 density, which we demonstrate in
 \textsc{Appendix~\ref{app:main-study-further-investigation}}. Thus, an open
-question remains whether choosing the intervals based on the empirical
-quantiles mitigates this issue.
-
-\begin{figure}[htbp]
+question remains whether the practical approach of choosing intervals based on
+empirical quantiles would mitigate this issue.
+
+\begin{figure}[ht!]
     \centering
@@
     \end{subfigure}
-    \\[10pt]
-    \vfill
-    \begin{subfigure}[b]{0.49\textwidth}
-        \includegraphics[width=\textwidth]{img/SNC/feature_effect_errors_ale_SVM_OT.png}
-        \caption{SVM\_OT}
-        \label{fig:ale-results-snc-svm-ot}  % chktex 24
-    \end{subfigure}
-    \caption{Main experiment results for ALE on SimpleNormalCorrelated dataset.\dots}
+    \caption{Main experiment results for ALE on SimpleNormalCorrelated dataset for GAM and
+    XGBoost models. The other models as well as the dummy features $x_3$ and $x_4$
+    are omitted for brevity.}
     \label{fig:ale-results-snc}  % chktex 24
@@
 estimating feature effects on training data. Further results on Friedman1 and
-Feynman I.29.16, as well as for the dummy features, largely confirm our
+Feynman I.29.16, as well as for the dummy features and the SVM largely confirm our
 described findings. For details, we refer to
-\textsc{Appendix~\ref{app:additional-results}}. This also means that we observe
+\textsc{Appendix~\ref{app:additional-results}}. This implies that we observe
 no substantial differences between scenarios with correlated features
 (SimpleNormalCorrelated) and independent features (Friedman1, Feynman I.29.16).
-Similarly, the more real-world relevant Feynman I.29.16 dataset shows results
+Similarly, the more real-world relevant Feynman I.29.16 dataset produces results
 in line with the other datasets. Note that we only observe one case where the
@@
 high bias. This occurs for the XGBoost\_OF model on the Friedman1 dataset,
-which we further look into in
+which we further examine in
 \textsc{Appendix~\ref{app:main-study-further-investigation}}.
@@
 
-We use the variance results from the main experiment and the Monte Carlo
-variance estimated in the ablation study to show the decomposition of the
-feature effect variance into model variance and Monte Carlo variance. The
-results for the Friedman1 dataset are given in
+Using variance results from the main experiment and Monte Carlo variance estimates
+from our ablation study, we decompose feature effect variance into its model
+variance and Monte Carlo variance components.
+Results for the Friedman1 dataset are presented in
 \textbf{Fig.\@~\ref{fig:pdp-variance-decomp}} for PDP and
 \textbf{Fig.\@~\ref{fig:ale-variance-decomp}} for ALE. Note that we now  % chktex 13
-consider the Friedman1 dataset as it provides a wider variety of different
-feature effects. The full results including the other datasets are provided in
+focus on Friedman1 as it offers a wider range of different
+feature effects. Full results including the other datasets are provided in
 \textsc{Appendix~\ref{app:additional-variance-decomposition}}.
 
-We observe that the Monte Carlo variance is generally highest when the feature
+Our analysis reveals that Monte Carlo variance is consistently highest when feature
 effects are estimated on the smaller validation set, which aligns with our
-expectations. Computing the feature effects in a cross-validated manner leads
-to the lowest Monte Carlo variance with a substantial reduction compared to
-estimation solely on training or validation data.
-
-In addition, we observe that the model variance is also reduced in many cases when
-estimating the feature effects in a cross-validated manner compared to an estimation
-on training or validation data, where model variance is (roughly) the same. The observed
-reduction through cross-validation is especially
+expectations. Cross-validated estimation yields the lowest Monte Carlo variance with
+a substantial reduction compared to
+estimation on either training or validation data alone.
+
+Additionally, model variance is also reduced in many cases through cross-validation
+compared to estimation on training or validation data, where model variance is
+approximately equal. The observed reduction through cross-validation is especially
 pronounced for the overfitting models (OF) and best visible for PDP
 (\textbf{Fig.\@~\ref{fig:pdp-variance-decomp}}). These observations are in line
-with our previous hypothesis that cross-validation reduces both model variance by averaging over
-different models as well as Monte Carlo variance by averaging over different
-samples. This is especially beneficial for overfitting models, which have a
-higher model variance.
+with our previous hypotheses that cross-validation reduces both model variance
+(by averaging across different models) and Monte Carlo variance (by leveraging
+different samples). This is particularly beneficial for overfitting models, which
+typically exhibit higher model variance.
 
@@
     \caption{Variance decomposition results for PDP on Friedman1 dataset. Variance is taken from the
-        main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
-        obtained by subtracting the Monte Carlo variance from the total variance. The second linear feature
+        main experiment, Monte Carlo variance is estimated in the ablation study. Model variance is
+        obtained by subtracting Monte Carlo variance from total variance. The second linear feature
         ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability. Results for XGBoost\_OT
@@
 
-While the variance due to MC integration is only a minor component of the total
-variance for PDP, it is the dominant component for ALE. Only for feature effect  % chktex 13
-estimation via cross-validation, the Monte Carlo variance is again reduced to a
-negligible level. This suggests that ALE is more sensitive to the dataset size
+\noindent We observe a striking difference between PDP and ALE:\@ while Monte Carlo
+variance represents only a minor component of total variance for PDP, it dominates
+the variance composition for ALE. Only with cross-validated estimation,  % chktex 13
+ALE's Monte Carlo variance is reduces to a negligible level. This suggests that
+ALE is more sensitive to the dataset size
 than PDP, which we further investigate in
-\textsc{Section~\ref{sec:results-dataset-size}}. We also suspect a certain
-amount of inaccuracy in the Monte Carlo variance estimation for ALE, which (a)
-leads to differences in the model variance between training and validation set
-and (b) causes the model variance to become negative in some cases (see
-\textbf{Fig.\@\ref{fig:ale-variance-decomp}}). Naturally, this is not possible
-and can most likely be attributed to numerical inaccuracies in the MC variance
+\textsc{Section~\ref{sec:results-dataset-size}}. We note some potential inaccuracies
+in Monte Carlo variance estimation for ALE, which manifest as: (a) differences
+in model variance between training and validation sets, and (b) negative model
+variance values in certain cases (see
+\textbf{Fig.\@\ref{fig:ale-variance-decomp}}). Since negative variance is theoretically
+impossible, we attribute these artifacts to numerical imprecisions in Monte Carlo variance
 estimation.
@@
     \caption{Variance decomposition results for ALE on Friedman1 dataset. Variance is taken from the
-        main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
-        obtained by subtracting the Monte Carlo variance from the total variance. The second linear feature
+        main experiment, Monte Carlo variance is estimated in the ablation study. Model variance is
+        obtained by subtracting Monte Carlo variance from total variance. The second linear feature
         ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability. Results for XGBoost\_OT
-        on 8000 training samples were not obtained due to excessive computation time. Negative variances are
-        due to numerical inaccuracies, marked by the hatch pattern.}
+        on 8000 training samples were not obtained due to excessive computation time. Negative variances, marked
+        by the hatch pattern, are due to numerical inaccuracies.}
     \label{fig:ale-variance-decomp}  % chktex 24
@@
 
-For both PDP and ALE, we find that the Monte Carlo variance, as expected,
-reduces with larger dataset size. This becomes clear both when comparing
+\noindent For both PDP and ALE, we find that the Monte Carlo variance, as expected,
+reduces with larger dataset size. This becomes clear both when contrasting
 training and validation data and when comparing the two dataset sizes (1000 and
-8000 samples). To analyze this not only for those few dataset sizes, we
-dedicate a separate ablation study to investigate this further in the next
-section.
+8000 samples). We systematically analyze this relationship across a broader range
+of dataset sizes in the next section.
 
@@
 
-We show the effect of the dataset size on the Monte Carlo variance for the
-features of the Friedman1 dataset for PDP and ALE in
-\textbf{Fig.\@~\ref{fig:mc-variance}}. The full results are provided in
+We show the effect of dataset size on Monte Carlo variance for four features
+of the Friedman1 dataset for PDP and ALE in
+\textbf{Fig.\@~\ref{fig:mc-variance}}. Complete results are provided in
 \textsc{Appendix~\ref{app:additional-mc-results}}. As expected, we observe that
-the Monte Carlo variance decreases with increasing dataset size (except for
-$x_3$ and $x_4$ for PDP, where the variance is generally very small). We
-find differences between the features: for more complex features like $x_1$ and
-$x_2$, which include interactions, the Monte Carlo variance is generally higher
-compared to simpler features like $x_3$ and $x_4$. Nonetheless, we observe an
-interesting difference between PDP and ALE. While they converge to a similar  % chktex 13
-Monte Carlo variance for large dataset sizes, the initial Monte Carlo variance
-for very small sample sizes is substantially higher for ALE than for PDP. This  % chktex 13
-demonstrates that ALE is more sensitive to the dataset size than PDP. This is  % chktex 13
+Monte Carlo variance generally decreases with increasing dataset size, with exceptions
+for PDP on features $x_3$ and $x_4$ where variance remains consistently low across all sample sizes.
+We observe feature-specific patterns: more complex features with interactions ($x_1$ and $x_2$)
+exhibit higher Monte Carlo variance compared to simpler features ($x_3$ and $x_4$).
+
+A striking difference emerges between PDP and ALE estimation methods. While both converge
+to similar Monte Carlo variance levels with large datasets, ALE exhibits substantially
+higher variance with small sample sizes. This
+demonstrates that ALE is more sensitive to dataset size compared to PDP. This is  % chktex 13
 most likely due to the fact that ALE estimates effects per interval and thus
-requires more samples to estimate the effect accurately compared to PDP, which
-uses the full dataset to estimate the effect. Note that this analysis
-also proves that the Monte Carlo variance introduced by estimating the ground truth
-feature effects in the main study is for most cases negligible compared to the
-other error sources.
+requires more samples to estimate the effect accurately, whereas PDP
+uses the full dataset to estimate the effect per grid point.
+
+This analysis also confirms that the Monte Carlo variance introduced by our ground truth
+feature effect estimation in the main study is negligible compared to other error sources
+for most scenarios.
 
@@
         For each data size $n_{mc}$, variances are averaged over all grid points.
-        Second linear feature ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability.}
+        The second linear feature ($x_5$) and dummy features ($x_6$, $x_7$) are omitted for better readability.}
     \label{fig:mc-variance}  % chktex 24
@@
 
-In addition, a detailed analysis of the Monte Carlo variance per grid point
-reveals that for PDP, the variance mostly depends on the sample density, while
-for ALE, the variance is also influenced by the shape of the underlying
-function. We provide detailed results in
-\textsc{Appendix~\ref{app:additional-mc-results}}.
-
-\clearpage
+\noindent Further analysis of the Monte Carlo variance per grid point
+reveals distinct patterns: for PDP, Monte Carlo variance is
+primarily influenced by sample density and shows larger values
+at distribution edges, while ALE variance mostly depends on 
+the underlying function's shape. We provide detailed grid point-level
+results in \textsc{Appendix~\ref{app:additional-mc-results}}.
+
 \section{Conclusions}\label{sec:conclusion}
@@
 
-Summarizing our main contributions, we (1) empirically investigated the
-impact of dataset choice for feature effect estimation. We (2) based our study on
-a theoretical framework that decomposes the feature effect error into different
-components and introduces estimators for each of them. We (3) conducted a
-review of test functions for simulation studies and build the choices for
-our study on these.
-
-Our comprehensive simulation study provided several key insights into
+Summarizing our main contributions, we (1) empirically investigated the impact of
+dataset choice for feature effect estimation, (2) embedded in a theoretical framework
+for decomposing feature effect error into distinct components with corresponding estimators,
+and (3) conducted a review of test functions for simulation studies to inform our experimental design.
+
+Our comprehensive simulation study provided several key insights regarding 
 the choice of dataset for feature effect estimation. First and foremost,
-we find no general evidence that speaks against the current predominant practice
-of using training data for computing feature effects. This is an important validation
-of existing approaches in the literature.
+we found no general evidence that speaks against the current predominant practice
+of using training data for estimating feature effects. This is an important validation
+of existing approaches.
 
 More specifically, our results demonstrated that feature effect errors are
-generally lower when more data is available for the estimation, which is usually
+generally lower when more data is available for estimation, which is usually
 the case for the training set. This is primarily due to reduced Monte Carlo variance,
 as shown in our ablation studies. The effect is particularly pronounced for ALE plots,
-which exhibit higher sensitivity to dataset size compared to PDPs, especially for smaller samples,
-due to the interval-based estimation. This suggests that for ALE plots, it is even more
+which exhibit higher sensitivity to dataset size compared to PDPs, especially for smaller sample
+sizes, due to interval-based estimation. This suggests that for ALE plots, it is even more
 important to use a sufficently large dataset for accurate estimation. Based on these
@@
 An interesting finding emerged regarding cross-validation for feature effect estimation.
-Our main study suggests that cross-validation is preferable in many scenarios,
-leading to lower overall errors. The ablation study reveals that this improvement stems from two distinct mechanisms:
+Our main study results suggest that cross-validation is preferable in many scenarios,
+leading to lower overall errors. The ablation study revealed that this improvement stems from two distinct mechanisms:
 (1) reduced Monte Carlo variance, and (2) reduced model variance through the averaging of estimates from different models.
@@
 
-We also observed that the correlation structure between features appears to have limited impact on our findings
-and are largely consistent across our set of different, diverse datasets.
+We also observed that the correlation structure between features appears to have limited impact on our findings,
+which are largely consistent across our set of different, diverse datasets.
 This suggests that our recommendations are robust across different data scenarios,
@@
 datasets or in regions of low data density. Moreover, this approach has limited real-world
-applicability, as we usually do not now the theoretical but only the empirical quantiles
-(which should be the same in expectation).
+applicability, as we usually do not now the theoretical but only the empirical quantiles.
 
@@
 particularly of ALE plots, indicate room for methodological improvements. Future research may
-focus on developing more robust estimation procedures for decomposing variance components,
+focus on developing more better estimation procedures for decomposing variance components,
 before applying these estimators to address other research questions.
@@
 applications, there are also use cases for which ``true to the model'' interpretations might be 
-preferred, which is not considered in our analysis.
+preferred.
 % also: data choice only with regard to feature effect error, not uncertainty estimation
-
-Future research could also investigate the impact of data shifts between training and validation sets.
-While our focused considered overfitting models, future work could explore the impact of data shifts
-between training and validation sets and their impact on feature effect estimation, analyzing the
-robustness of the estimation under distribution shifts.
 
@@
 conventional estimation on a single dataset, this approach could be enhanced further.
-Future research could explore strategies that utilize more data per fold, such as using the
-training set in each fold for feature effect estimation (and use the validation part only for
-perfomance estimation) or combining training and validation splits. Alternatively, bootstrapping
-strategies similar to the approaches pursued by Molnar et al.~\cite{molnar_relating_2023} could be
+Future research could explore strategies that utilize more data per fold such as using the
+training set in each fold for feature effect estimation (and validation only for e.g.\ perfomance estimation).
+Alternatively, bootstrapping strategies similar to the approaches pursued in~\cite{molnar_relating_2023} could be
 explored.
 
-Following the observations of Loecher~\cite{loecher_debiasing_2024} that hat MDI and SHAP show
+Following the observations of Loecher~\cite{loecher_debiasing_2024} that MDI and SHAP show
 a bias toward high-entropy variables for Random Forests, future work might also investigate if and how
@@
 have features with varying and also just partial effect strengths, different noise levels,
-higher-order interactions, and diverse marginal distributions. Future work could focus on
+higher-order interactions, and diverse marginal distributions. Future work should focus on
 developing more realistic datasets capturing these aspects, to further bridge the gap between
 research, such as simulation studies, and practical applications.
-
-These limitations and future directions highlight the rich opportunities for further research
-in this area to obtain practically relevant insights.
 
@@
          & \quad + \mathbb{E}_F{[PDP_{\hat f}]}^2 + \mathbb{E}_F\text{Var}_X[\widehat{PDP}_{\hat f}]                                    \\
-         & = {(PDP_f - \mathbb{E}_F[PDP_{\hat f}])}^2 + \text{Var}_F[PDP_{\hat f}] + \mathbb{E}_F[\text{Var}_X(\widehat{PDP}_{\hat f})]
+         & = {(PDP_f - \mathbb{E}_F[PDP_{\hat f}])}^2 + \text{Var}_F[PDP_{\hat f}] \\
+         &\quad + \mathbb{E}_F[\text{Var}_X(\widehat{PDP}_{\hat f})]
     \end{align*}
@@
 Specifically, we provide details on the hyperparameter selection and the
-finally used hyperparameters, as well as the performance of the models.
+finally used hyperparameters, as well as the performances of the models.
 
@@
 of 10000 to get a reliable performance estimate. While this scenario is
-unrealistic in practice, it allows as to consider hyperparameters as
+unrealistic in practice, it allows us to consider hyperparameters as
 ``pre-selected'' and avoid costly nested resampling strategies for the
@@
 both on the training data and a holdout test data with the latter consisting of
-10000 samples to get a reliable performance estimate. The performances
+10000 samples to get a reliable performance estimate. The performances (R2-scores)
 evaluated over the 30 repetitions are aggregated in
-\textbf{Fig.~\ref{fig:model-performance}} showing the R2-scores. As intended,
+\textbf{Fig.~\ref{fig:model-performance}}. As intended,
 the overfitting models perform better on the training data than the optimally
@@
 exhibiting higher variance in their generalization performance. Note that the
-linear regression model serves only as a baseline. The overfitted SVM (SVM\_OF)
+linear regression model serves only as a baseline. The overfitted SVMs (SVM\_OF)
 show substantially worse performances on test data compared to the other
-models. Therefore, we excluded it from further analysis.
+models. Therefore, we do not discuss them in more detail in our analyses.
 
@@
     \caption{R2-scores of the models evaluated on training samples (left) and 10000 holdout test samples (right).
-Each boxplot aggregates 30 repetitions with differing seeds for the data generation (cf.\ \textsc{Section~\ref{sec:experiment-main}}).}
+    Each boxplot aggregates 30 repetitions with differing seeds for data generation (cf.\ \textsc{Section~\ref{sec:experiment-main}}).}
     \label{fig:model-performance}  % chktex 24
@@
 considered datasets. In addition, we provide visualizations of the derived
-feature effects and compare them to the estimated feature effects as gut check.
+feature effects and compare them to the estimated feature effects as verification.
 
@@
         \includegraphics[width=\textwidth]{img/SNC-all/pdp_groundtruth_comparison.png}
-        \caption{PDP ground truth of SimpleNormalCorrelated}
+        \caption{PDP ground truth}
     \end{subfigure}
@@
         \includegraphics[width=\textwidth]{img/SNC-all/ale_groundtruth_comparison.png}
-        \caption{ALE ground truth of SimpleNormalCorrelated}
+        \caption{ALE ground truth}
     \end{subfigure}
@@
     effect on the ground truth function using 10000 Monte Carlo samples. Note that both
-    are centered that the average effect at the quantile-based grid points is 0.
-    Since $x_3$ and $x_4$, are dummy variables, they have a constant effect of 0 and are
+    are centered to mean 0 over the quantile-based grid-points.
+    Since $x_3$ and $x_4$ are dummy variables, they have a constant effect of 0 and are
     thus omitted.}
@@
                  &= x_1 + \frac{1}{2} \\
-                 &= x_1 + C_1
+                 &= x_1 + C
 \end{align*}
@@
 \begin{align*}
-    ALE_{f,1}(x_1) &= \int_{x_{min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[\frac{\partial}{\partial X_1} X_1 + 0.5 X_2^2 + X_1X_2 \Big| X_1=z_1\right] dz_1 - C_1 \\
-    &= \int_{x_{min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[1 + X_2 \Big| X_1=z_1\right] dz_1 - C_1 \\
+    ALE_{f,1}(x_1) &= \int_{x_{\min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[\frac{\partial}{\partial X_1} X_1 + 0.5 X_2^2 + X_1X_2 \Big| X_1=z_1\right] dz_1 - C \\
+    &= \int_{x_{\min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[1 + X_2 \Big| X_1=z_1\right] dz_1 - C \\
     &(\text{using } X_2 = \beta_{2|1}(X_1-\mu_1) + \varepsilon_1 + \mu_2\text{, where } \beta_{2|1}=\rho \frac{\sigma_2}{\sigma_1}) \\
-    &= \int_{x_{min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[1 + \beta_{2|1}(X_1-\mu_1) + + \varepsilon_1 + \mu_2 \Big| X_1=z_1\right] dz_1 - C_1 \\
-    &= \int_{x_{min,1}}^{x_1} 1 + \beta_{2|1}(z_1-\mu_1) + \mu_2 \,dz_1 - C_1 \\
-    &= [z_1 + \frac{1}{2}\beta_{2|1}(z_1-\mu_1)^2 + \mu_2 z_1]_{x_{min,1}}^{x_1} - C_1 \\
+    &= \int_{x_{\min,1}}^{x_1} \mathbb{E}_{X_2|X_1}\left[1 + \beta_{2|1}(X_1-\mu_1) + \varepsilon_1 + \mu_2 \Big| X_1=z_1\right] dz_1 - C \\
+    &= \int_{x_{\min,1}}^{x_1} 1 + \beta_{2|1}(z_1-\mu_1) + \mu_2 \,dz_1 - C \\
+    &= [z_1 + \frac{1}{2}\beta_{2|1}(z_1-\mu_1)^2 + \mu_2 z_1]_{x_{\min,1}}^{x_1} - C \\
     &= x_1 + \frac{\beta_{2|1}}{2}(x_1-\mu_1)^2 + \mu_2 x_1 \\
-    & \quad - x_{min,1} - \frac{\beta_{2|1}}{2}(x_{min,1}-\mu_1)^2 - \mu_2 x_{min,1} - C_1 \\
-    % &= (1 + \mu_2) \cdot (x_1 - x_{min,1}) + \frac{\beta_{2|1}}{2} \cdot ((x_1-\mu_1)^2 - (x_{min,1}-\mu_1)^2) - C \\
-    % &= (1 + \mu_2) \cdot (x_1 - x_{min,1}) + \frac{\beta_{2|1}}{2} \cdot (x_1^2 - 2\mu_1 x_1 - x_{min,1}^2 + 2\mu_1 x_{min,1}) - C \\
-    &= (1 + \mu_2) x_1 - (1 + \mu_2) x_{min,1} + \frac{\beta_{2|1}}{2} x_1^2 - \frac{\beta_{2|1}}{2} x_{min,1}^2 \\
-    & \quad - \mu_1 \beta_{2|1} x_1 - \mu_1 \beta_{2|1}  x_{min,1} - C_1 \\
-    &= \frac{\beta_{2|1}}{2} x_1^2 + (1 + \mu_2 - \mu_1 \beta_{2|1}) x_1 - C_1^\prime \\
-    &= 0.45 x_1^2 + x_1 - C_1^\prime
+    & \quad - x_{\min,1} - \frac{\beta_{2|1}}{2}(x_{\min,1}-\mu_1)^2 - \mu_2 x_{\min,1} - C \\
+    % &= (1 + \mu_2) \cdot (x_1 - x_{\min,1}) + \frac{\beta_{2|1}}{2} \cdot ((x_1-\mu_1)^2 - (x_{\min,1}-\mu_1)^2) - C \\
+    % &= (1 + \mu_2) \cdot (x_1 - x_{\min,1}) + \frac{\beta_{2|1}}{2} \cdot (x_1^2 - 2\mu_1 x_1 - x_{\min,1}^2 + 2\mu_1 x_{\min,1}) - C \\
+    &= (1 + \mu_2) x_1 - (1 + \mu_2) x_{\min,1} + \frac{\beta_{2|1}}{2} x_1^2 - \frac{\beta_{2|1}}{2} x_{\min,1}^2 \\
+    & \quad - \mu_1 \beta_{2|1} x_1 - \mu_1 \beta_{2|1}  x_{\min,1} - C \\
+    &= \frac{\beta_{2|1}}{2} x_1^2 + (1 + \mu_2 - \mu_1 \beta_{2|1}) x_1 - C^\prime \\
+    &= 0.45 x_1^2 + x_1 - C^\prime
 \end{align*}
@@
 \begin{align*}
-    ALE_{f,2}(x_2) &= (\frac{1}{2} + \mu_1 + \frac{\beta_{1|2}}{2}) x_2^2 - \beta_{1|2} \mu_2 x_2 - C_2^\prime \\
-    &= 0.95 x_2^2 - C_2^\prime
+    ALE_{f,2}(x_2) &= (\frac{1}{2} + \mu_1 + \frac{\beta_{1|2}}{2}) x_2^2 - \beta_{1|2} \mu_2 x_2 - C^\prime \\
+    &= 0.95 x_2^2 - C^\prime
 \end{align*}
@@
 \subsection{Friedman1 Dataset}
+
+\begin{figure*}[h!]
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/pdp_groundtruth_comparison.png}
+        \caption{PDP ground truth of Friedman1}
+    \end{subfigure}
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{img/Friedman1-all/ale_groundtruth_comparison.png}
+        \caption{ALE ground truth of Friedman1}
+    \end{subfigure}
+    \caption{Ground truth feature effects for Friedman1 dataset. The green line
+    represents the analytically derived ground truth, the orange line the estimated feature
+    effect on the ground truth function using 10000 Monte Carlo samples. Note that both
+    are centered to mean 0 over the quantile-based grid-points.
+    Since $x_6$ and $x_7$, are dummy variables, they have a constant effect of 0 and are
+    thus omitted.}
+    \label{fig:friedman1-groundtruth}  % chktex 24
+\end{figure*}
 
@@
 $\sigma_S^2 = Var[X]=\frac{(b-a)^2}{12}=\frac{1}{12}$.
-Si denotes the sine integral function, $C_S$ and $C_S^\prime$ are arbitrary constants.
+Si denotes the sine integral function, $C$ and $C^\prime$ are arbitrary constants.
 
@@
     &= 10 \cdot \frac{-\cos(\pi bx_1) + \cos(\pi ax_1)}{\pi x_1} + 20(\mu_3^2 + \sigma_3^2 - \mu_3) + 5 + 10\mu_4 + 5\mu_5 \\
-    &= 10 \cdot \frac{-\cos(\pi x_1) + 1}{\pi x_1} + C_1 \\
+    &= 10 \cdot \frac{-\cos(\pi x_1) + 1}{\pi x_1} + C \\
 \end{align*}
 
+\newpage
 \noindent Analogously, we get for $x_2$:
@@
 \begin{align*}
-    PDP_{f,2}(x_2) &= 10 \cdot \frac{-\cos(\pi x_2) + 1}{\pi x_2} + C_2
+    PDP_{f,2}(x_2) &= 10 \cdot \frac{-\cos(\pi x_2) + 1}{\pi x_2} + C
 \end{align*}
@@
     &\quad + 20x_3^2 - 20x_3 + 5 + 10\mu_4 + 5\mu_5 \\
-    &= 20x_3^2 - 20x_3 + C_3 \\
+    &= 20x_3^2 - 20x_3 + C \\
 \end{align*}
@@
 \begin{align*}
-    PDP_{f,4}(x_4) &= 10x_4 + C_4
+    PDP_{f,4}(x_4) &= 10x_4 + C
 \end{align*}
@@
 \begin{align*}
-    PDP_{f,2}(x_2) &= 5x_5 + C_5
+    PDP_{f,5}(x_5) &= 5x_5 + C
 \end{align*}
@@
 \begin{align*}
-    ALE_{f,1}(x_1) &= \dots
+    ALE_{f,1}(x_1) &= \int_{x_{1,\min}}^{x_1} \mathbb{E}_{X_{-1}|X_1}\Bigg[\frac{\partial}{\partial X_1}(10\sin(\pi X_1 X_2) + 20(X_3-\frac{1}{2})^2 + 10X_4 \Bigg. \\
+                   &\quad \Bigg. + 5X_5)\bigg|X_1=z_1\Bigg]dz_1 - C\\
+                   &= \int_{x_{1,\min}}^{x_1} \mathbb{E}_{X_{-1}|X_1}\bigg[10\pi X_2\cos(\pi X_1 X_2)\bigg|X_1=z_1\bigg]dz_1 - C \\
+                   &\stackrel{\scriptsize\text{(ind.)}}{=} \int_{x_{1,\min}}^{x_1} 10\pi \mathbb{E}_{X_2}\left[X_2\cos(\pi z_1 X_2)\right]dz_1 - C \\
+                   &= \int_{x_{1,\min}}^{x_1} 10\pi \int_{a}^{b} x_2\cos(\pi z_1 x_2) \cdot p(x_2) dx_2 dz_1 - C \\
+                   &= \int_{x_{1,\min}}^{x_1} \frac{10\pi}{b-a} \int_{a}^{b} x_2\cos(\pi z_1 x_2)dx_2 dz_1 - C \\
+                   &= \int_{x_{1,\min}}^{x_1} \frac{10\pi}{b-a} \left[\frac{x_2\sin(\pi z_1 x_2)}{\pi z_1} + \frac{\cos(\pi z_1 x_2)}{(\pi z_1)^2}\right]_a^b dz_1 - C \\
+                   &= \frac{10}{b-a}\int_{x_{1,\min}}^{x_1} \frac{b\sin(\pi z_1 b) - a\sin(\pi z_1 a)}{z_1} \\ 
+                   &\quad + \frac{\cos(\pi z_1 b) - \cos(\pi z_1 a)}{\pi z_1^2} dz_1 - C \\
+                   &= \frac{10}{b-a}\bigg(b \cdot \int_{x_{1,\min}}^{x_1} \frac{\sin(\pi z_1 b)}{z_1}dz_1 - a \cdot \int_{x_{1,\min}}^{x_1} \frac{\sin(\pi z_1 a)}{z_1}dz_1 \\
+                   &\quad + \int_{x_{1,\min}}^{x_1} \frac{\cos(\pi z_1 b)}{\pi z_1^2}dz_1 - \int_{x_{1,\min}}^{x_1} \frac{\cos(\pi z_1 a)}{\pi z_1^2}dz_1\bigg) - C \\
+                   &= \frac{10}{b-a}\bigg[b \cdot (Si(\pi x_1 b)-Si(\pi x_{1,\min} b)) - a \cdot (Si(\pi x_1 a)- Si(\pi x_{1,\min} a)) \\
+                   &\quad + b \cdot \left(-\frac{\cos(\pi x_1 b)}{\pi x_1 b} - Si(\pi x_1 b) + \frac{\cos(\pi x_{1,\min} b)}{\pi x_{1,\min} b} + Si(\pi x_{1,\min} b)\right) \\
+                   &\quad - a \cdot \left(-\frac{\cos(\pi x_1 a)}{\pi x_1 a} - Si(\pi x_1 a) + \frac{\cos(\pi x_{1,\min} a)}{\pi x_{1,\min} a} + Si(\pi x_{1,\min} a)\right)\bigg] \\
+                   &\quad - C \\
+                   &= \frac{10}{b-a}\bigg[-\frac{1}{\pi x_1}\left(\cos(\pi x_1 b)-\cos(\pi x_1 a)\right) \\
+                   &\quad + \frac{1}{\pi x_{1,\min}}\left(\cos(\pi x_{1,\min} b)-\cos(\pi x_{1,\min} a)\right)\bigg] - C \\
+                   &= \frac{10}{b-a}\frac{1}{\pi x_1}\left(\cos(\pi x_1 a) - \cos(\pi x_1 b)\right) - C^\prime \\
+                   &= \frac{10}{\pi x_1}\left(1 - \cos(\pi x_1)\right) - C^\prime
 \end{align*}
 
+\newpage
+\noindent Analogously, we get for $x_2$:
+
 \begin{align*}
-    ALE_{f,2}(x_2) &= \dots
+    ALE_{f,2}(x_2) &= \frac{10}{\pi x_2}\left(1 - \cos(\pi x_2)\right) - C^\prime
 \end{align*}
@@
     ALE_{f,3}(x_3) &= \int_{x_{3,\min}}^{x_3} \mathbb{E}_{-X_3|X_3}\Bigg[\frac{\partial}{\partial x_3}(10\sin(\pi X_1X_2) + 20(X_3-\frac{1}{2})^2 + 10X_4 \Bigg. \\
-    &\qquad\qquad\qquad\qquad\qquad \Bigg. + 5X_5) \big|X_3=z_3\Bigg]dz_3 - C_3 \\
-    &= \int_{x_{3,\min}}^{x_3} \mathbb{E}_{-X_3|X_3}\left[40z_3-20\right]dz_3 - C_3 \\
-    &= \int_{x_{3,\min}}^{x_3} 40z_3-20dz_3 - C_3 \\
-    &= \left[20z_3^2 - 20z_3\right]_{x_{3,\min}}^{x_3} - C_3 \\
-    &= 20x_3^2 - 20x_3 - 20x_{3,\min}^2 + 20x_{3,\min} - C_3 \\
+    &\qquad\qquad\qquad\qquad\quad \Bigg. + 5X_5) \big|X_3=z_3\Bigg]dz_3 - C \\
+    &= \int_{x_{3,\min}}^{x_3} \mathbb{E}_{-X_3|X_3}\left[40z_3-20\right]dz_3 - C \\
+    &= \int_{x_{3,\min}}^{x_3} 40z_3-20dz_3 - C \\
+    &= \left[20z_3^2 - 20z_3\right]_{x_{3,\min}}^{x_3} - C \\
+    &= 20x_3^2 - 20x_3 - 20x_{3,\min}^2 + 20x_{3,\min} - C \\
     &= 20x_3^2 - 20x_3 - C_3^\prime
@@
     ALE_{f,4}(x_4) &= \int_{x_{4,\min}}^{x_4} \mathbb{E}_{-X_4|X_4}\Bigg[\frac{\partial}{\partial x_4}(10\sin(\pi X_1X_2) + 20(X_3-\frac{1}{2})^2 + 10X_4 \Bigg. \\
-    &\qquad\qquad\qquad\qquad\qquad \Bigg. {} + 5X_5) \big|X_4=z_4\Bigg]dz_4 - C_4 \\
+    &\qquad\qquad\qquad\qquad\quad \Bigg. {} + 5X_5) \big|X_4=z_4\Bigg]dz_4 - C_4 \\
     &= \int_{x_{4,\min}}^{x_4} \mathbb{E}_{-x_4|x_4}\left[10\right]dz_4 - C_4 \\
@@
 
-\begin{figure*}[h!]
-    \begin{subfigure}[b]{\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{img/Friedman1-all/pdp_groundtruth_comparison.png}
-        \caption{PDP ground truth of Friedman1}
-    \end{subfigure}
-    \begin{subfigure}[b]{\textwidth}
-        \centering
-        \includegraphics[width=\textwidth]{img/Friedman1-all/ale_groundtruth_comparison.png}
-        \caption{ALE ground truth of Friedman1}
-    \end{subfigure}
-    \caption{Ground truth feature effects for Friedman1 dataset. The green line
-    represents the analytically derived ground truth, the orange line the estimated feature
-    effect on the ground truth function using 10000 Monte Carlo samples. Note that both
-    are centered that the average effect at the quantile-based grid points is 0.
-    Since $x_6$ and $x_7$, are dummy variables, they have a constant effect of 0 and are
-    thus omitted.}
-    \label{fig:friedman1-groundtruth}  % chktex 24
-\end{figure*}
-
+\newpage
 \subsection{FeynmanI.29.16 Dataset}
@@
     \end{subfigure}
-    \caption{Ground truth feature effects for FeynmanI.29.16 dataset. The green line
-    represents the analytically derived ground truth, the orange line the estimated feature
-    effect on the ground truth function using 10000 Monte Carlo samples. Note that both
-    are centered that the average effect at the quantile-based grid points is 0.
+    \caption{Ground truth feature effects for FeynmanI.29.16 dataset. The estimated feature
+    effect on the ground truth function using 10000 Monte Carlo samples is displayed in orange,
+    analytical feature effects are infeasible. The effects are centered to mean 0 over the quantile-based grid-points.
     Since $d_1$ and $d_2$, are dummy variables, they have a constant effect of 0 and are
@@
 
+In the following, we provide the complete results of the main experiment,
+including all datasets, models, and features. Note that the linear regression
+only serves as a performance baseline and the SVM\_OF is severly overfitted and
+not producing meaningful results. Therefore, we recommend to focus on the
+other results.
+
 \subsection{SimpleNormalCorrelated Dataset}
@@
 
+In this section, we provide a deeper investigation of selected interesting results from the main experiment.
+For this analysis, we plot the bias and variance per grid point over the feature range to gain
+further insights.
+
 \begin{figure}
@@
     \caption{$\text{Bias}^2$ and variance per grid point for PDP of XGBoost\_OF on SimpleNormalCorrelated dataset with $n_{train} = 1000$.
-        This example demonstrates the cross-validation reduces the variance mainly in regions where data is scarce, i.e.,
+        This example demonstrates that cross-validation reduces the variance mainly in regions where data is scarce, i.e.,
         at the edges of the feature space. In this regions, the model has only few data points to learn from, and approximations
@@
 
+In the following, we provide the complete results of the ablation study on variance decomposition,
+including all datasets, features, and considered models.
+
 \subsection{SimpleNormalCorrelated Dataset}
@@
     \end{subfigure}
-    \caption{Variance decomposition results for PDP on SimpleNormalCorrelated dataset. Variance is taken from the
+    \caption{Variance decomposition results for ALE on SimpleNormalCorrelated dataset. Variance is taken from the
         main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
@@
     \end{subfigure}
-    \caption{Variance decomposition results for PDP on Friedman1 dataset. Variance is taken from the
+    \caption{Variance decomposition results for ALE on Friedman1 dataset. Variance is taken from the
         main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
@@
 
-\begin{figure}[htbp]
+\begin{figure}[ht!]
     \centering
@@
     \end{subfigure}
-    \caption{Variance decomposition results for PDP on FeynmanI.29.16 dataset. Variance is taken from the
+    \caption{Variance decomposition results for PDP on FeynmanI.29.16. Variance is taken from the
         main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
@@
     \end{subfigure}
-    \caption{Variance decomposition results for PDP on FeynmanI.29.16 dataset. Variance is taken from the
+    \caption{Variance decomposition results for ALE on FeynmanI.29.16. Variance is taken from the
         main experiment, Monte Carlo variance is estimated in the ablation study. The model variance is
@@
 \section{Additional Results for Ablation Study: Datasize Impact on Monte Carlo Variance}\label{app:additional-mc-results}
+
+In this section, we provide additional results for the ablation study on the impact of dataset size
+on the Monte Carlo variance. This includes the complete aggregate results for both SimpleNormalCorrelated and
+Friedman1 for all features, as well as a more detailed view on the Monet Carlo variances per grid point.
 

### 2025-02-20 current_research_feature_effects/data_generating/data_generation.py
Clean up python library

@@ -1,3 +1,7 @@
+"""
+This module contains classes and functions for generating synthetic data based on a groundtruth function.
+"""
+
 from abc import abstractmethod, ABC
 from typing import Callable, Literal, List, Optional, Tuple
 import warnings

### 2025-02-20 current_research_feature_effects/data_generating/friedman1.py

@@ -1,3 +1,7 @@
+"""
+This module contains the groundtruth class for the Friedman1 function.
+"""
+
 from typing import Callable
 import pandas as pd
 import numpy as np
@@ -16,7 +20,7 @@ class Friedman1Groundtruth(Groundtruth):
     def predict(self, X) -> np.ndarray:
         """
         Returns target value (y) of the groundtruth (Friedman1) for each
-        sample in X. The output `y` is created according to the formula::
+        sample in X. The output `y` is created according to the formula:
 
         y(X) = 10 * sin(pi * X[:, 0] * X[:, 1]) + 20 * (X[:, 2] - 0.5) ** 2 \
         + 10 * X[:, 3] + 5 * X[:, 4].

### 2025-02-20 current_research_feature_effects/data_generating/physics_data.py

@@ -1,3 +1,8 @@
+"""
+This module contains classes for generating data based on physics laws.
+Currently, the Feynman equations I.9.18 and I.29.16 are implemented.
+"""
+
 from typing import Callable
 import numpy as np
 import pandas as pd

### 2025-02-20 current_research_feature_effects/data_generating/simple.py

@@ -1,3 +1,7 @@
+"""
+This module contains three simple groundtruth classes for generating data based on simple formulas.
+"""
+
 from typing import Callable
 import pandas as pd
 import numpy as np

### 2025-02-20 current_research_feature_effects/feature_effects.py

@@ -1,3 +1,8 @@
+"""
+This module defines classes and functions to compute feature effects
+and feature effect error components.
+"""
+
 from typing_extensions import List, Dict, Callable, Literal, Union
 from copy import deepcopy
 from dataclasses import dataclass
@@ -230,45 +235,6 @@ def _accumulated_local_effects(
     return res_df
 
 
-# def get_modified_grids(
-#     base_grids: List[np.ndarray], Xs: List[np.ndarray], feature_names: List[str]
-# ) -> Tuple[np.ndarray]:
-#     """
-#     Get modified grids based on the minimum and maximum values of the datasets.
-
-#     Parameters
-#     ----------
-#     base_grids : List[np.ndarray]
-#         List of base grids to be modified.
-#     Xs : List[np.ndarray]
-#         List of datasets to compute the minimum and maximum values from.
-#     feature_names : List[str]
-#         Names of features to compute the minimum and maximum values from.
-
-#     Returns
-#     -------
-#     Tuple[np.ndarray]
-#         Tuple of modified grids based on the minimum and maximum values of the datasets,
-#         length of the tuple is equal to the number of datasets.
-#     """
-#     mins = np.array([[np.min(X[:, i]) for i in range(X.shape[1])] for X in Xs]).T
-#     maxs = np.array([[np.max(X[:, i]) for i in range(X.shape[1])] for X in Xs]).T
-
-#     common_mins = mins.max(axis=1)
-#     common_maxs = maxs.min(axis=1)
-
-#     dataset_grids = [[] for _ in range(len(Xs))]
-#     for i in range(len(feature_names)):
-#         base_grid = base_grids[i]
-#         filtered_grid = base_grid[(base_grid > common_mins[i]) & (base_grid < common_maxs[i])]
-
-#         for j in range(len(Xs)):
-#             specific_grid = np.concatenate([[mins[i, j]], filtered_grid, [maxs[i, j]]])
-#             dataset_grids[j].append(specific_grid)
-
-#     return tuple(dataset_grids)
-
-
 def compute_pdps(
     model: BaseEstimator,
     X: np.ndarray,
@@ -396,7 +362,7 @@ def compute_theoretical_effects(
     remove_first_last: bool = False,
 ) -> FeatureEffect:
     """
-    Compute theoretical partial dependence plots for a given groundtruth
+    Compute theoretical feature effects (PDP or ALE) for a given groundtruth
     and apply it to the grid_values.
 
[…]

### 2025-02-20 current_research_feature_effects/mappings.py

@@ -1,3 +1,9 @@
+"""
+This module contains several mappings: dataset name to groundtruth, model name to estimator, and suggested
+hyperparameters for each model, as well as a wrapper for a Generalized Additive Model (GAM) compatible with
+the sklearn API.
+"""
+
 from typing_extensions import List, Tuple, Literal
 from itertools import combinations
 import numpy as np

### 2025-02-20 current_research_feature_effects/model_eval.py

@@ -1,3 +1,7 @@
+"""
+This module contains functions for evaluating model performance.
+"""
+
 from typing_extensions import Dict, List
 import numpy as np
 from sklearn.base import BaseEstimator

### 2025-02-20 current_research_feature_effects/model_training.py

@@ -1,3 +1,7 @@
+"""
+This module contains functions for hyperparameter optimization and model training.
+"""
+
 from typing_extensions import Literal, Dict
 from pathlib import Path
 from configparser import ConfigParser
@@ -20,6 +24,9 @@ def _objective(
     trial: optuna.trial.Trial,
     metric: str,
 ) -> float:
+    """
+    Objective function for hyperparameter optimization.
+    """
     hyperparams = suggested_hps_for_model(model, trial)
     model.set_params(**hyperparams)
 
@@ -48,6 +55,37 @@ def optimize(
     study_name: str,
     storage_name: Path,
 ) -> optuna.study.Study:
+    """
+    Optimize hyperparameters of a model using Optuna.
+
+    Parameters
+    ----------
+    model : BaseEstimator
+        Model to optimize.
+    X_train : np.ndarray
+        Training features.
+    y_train : np.ndarray
+        Training labels.
+    X_val : np.ndarray
+        Validation features.
+    y_val : np.ndarray
+        Validation labels.
+    n_trials : int
+        Number of trials.
+    metric : str
+        Metric to optimize.
+    direction : Literal[&quot;maximize&quot;, &quot;minimize&quot;]
+        Direction of optimization.
+    study_name : str
+        Name of the optuna study.
+    storage_name : Path
+        Optuna storage name.
+
+    Returns
+    -------
+    optuna.study.Study
+        Optuna optimization study.
+    """
     optuna.logging.set_verbosity(optuna.logging.ERROR)
 
     sampler = optuna.samplers.TPESampler(seed=42)

### 2025-02-20 current_research_feature_effects/plotting/__init__.py

@@ -1,0 +1,3 @@
+"""
+Module for formatting and plotting the results of the feature effects analysis.
+"""

### 2025-02-20 current_research_feature_effects/plotting/plots.py

@@ -1,5 +1,7 @@
-# from configparser import ConfigParser
-# import math
+"""
+This module contains functions to plot feature effects, model evaluation results, and feature effect errors.
+"""
+
 from typing_extensions import Dict, Literal, List, Tuple, Optional
 from pathlib import Path
 import pandas as pd
@@ -15,6 +17,7 @@ from current_research_feature_effects.plotting.utils import (
     set_style,
     get_boxplot_style,
     get_feature_effect_plot_style,
+    set_fontsize,
 )
 
 
@@ -92,9 +95,9 @@ def plot_feature_effect_vs_groundtruth(
     palette = sns.color_palette("Set2")
 
     if large_font:
-        _set_fontsize("xlarge")
+        set_fontsize("xlarge")
     else:
-        _set_fontsize("standard")
+        set_fontsize("standard")
 
     fig, axes = plt.subplots(1, len(feature_names), figsize=(5 * len(feature_names), 5), dpi=300, sharey=True)
     for i, f in enumerate(feature_names):
@@ -135,11 +138,30 @@ def boxplot_model_results(
     ylim: Optional[Tuple[float, float]] = None,
     large_font: bool = False,
 ) -> plt.Figure:
+    """
+    Create a boxplot of the model evaluation results.
+
+    Parameters
+    ----------
+    metric : Literal[&quot;mse&quot;, &quot;mae&quot;, &quot;r2&quot;]
+        The metric to plot.
+    df : pd.DataFrame
+        The dataframe containing the model evaluation results.
+    ylim : Optional[Tuple[float, float]], optional
+        The y-axis limits, by default None.
+    large_font : bool, optional
+        If True, use larger font sizes, by default False.
+
+    Returns
+    -------
+    plt.Figure
+        The figure object containing the generated plot.
+    """
     set_style()
     if large_font:
-        _set_fontsize("large")
+        set_fontsize("large")
     else:
-        _set_fontsize("standard")
+        set_fontsize("standard")
     fig, ax = plt.subplots(1, 2, figsize=(12, 6), dpi=300, sharey=True)
     fig.suptitle("Model evaluation", fontsize=16, fontweight="bold")
     ax[0].set_title(f"{metric} on train set")
@@ -170,53 +192,6 @@ def boxplot_model_results(
     return fig
 
 
-def _set_fontsize(size: Literal["standard", "large", "xlarge"]) -> None:
-    """
-    Set the font size of the plots.
-
-    Parameters
-    ----------
-    size : {'standard', 'large', 'xlarge'}
[…]

### 2025-02-20 current_research_feature_effects/plotting/table_formatting.py

@@ -1,3 +1,8 @@
+"""
+This module contains functions to format dataframes for better readibility and
+to simplify plotting.
+"""
+
 from pathlib import Path
 from configparser import ConfigParser
 from typing import Literal

### 2025-02-20 current_research_feature_effects/plotting/utils.py

@@ -1,8 +1,15 @@
+"""
+This module contains utility functions for plotting.
+"""
+
+from typing import Literal
 import matplotlib.pyplot as plt
-import pandas as pd
 
 
 def set_style():
+    """
+    Set the style of the plots.
+    """
     plt.style.use("fivethirtyeight")
 
     plt.rcParams.update(
@@ -24,6 +31,14 @@ def set_style():
 
 
 def get_boxplot_style():
+    """
+    Set the style of the boxplot.
+
+    Returns
+    -------
+    Dict
+        Dictionary containing the style of the boxplot.
+    """
     style = dict(
         boxprops=dict(edgecolor="black"),  # Box properties
         whiskerprops=dict(color="black"),  # Whisker properties
@@ -37,60 +52,61 @@ def get_boxplot_style():
 
 
 def get_feature_effect_plot_style():
+    """
+    Set the style of the feature effect plot.
+
+    Returns
+    -------
+    Dict
+        Dictionary containing the style of the feature effect plot.
+    """
     style = dict(linewidth=2, marker="+", markeredgewidth=1, markersize=5)
 
     return style
 
 
-def create_joined_melted_df(
-    df_model_res: pd.DataFrame,
-    df_pdp_res: pd.DataFrame,
-    noise_name: str = "snr_x",
-    value_vars: list[str] = ["x_1", "x_2", "x_3", "x_4", "x_5"],
-) -> pd.DataFrame:
+def set_fontsize(size: Literal["standard", "large", "xlarge"]):
     """
-    Merge two DataFrames on 'model_id' and reshape the resulting DataFrame to a long format.
-
-    This function performs an inner join on two DataFrames based on the 'model_id' column.
-    After merging, it reshapes the DataFrame to long format, transforming feature columns
-    ('x_1' to 'x_5') into a single 'feature' column and their values into an 'effect_error'
-    column.
+    Set the font size of the plots.
 
     Parameters
     ----------
-    df_model_res : pd.DataFrame
-        DataFrame containing model results, which must include the 'model_id' column
-        along with other model performance metrics such as mse_test, r2_train, etc.
-    df_pdp_res : pd.DataFrame
-        DataFrame containing PDP results, also must include the 'model_id' column and
-        the values for each feature ('x_1' to 'x_5').
-    noise_name : str, optional
-        Name of the noise column in the merged DataFrame, by default "snr_x"
-    value_vars : list[str], optional
[…]

### 2025-02-20 current_research_feature_effects/utils.py

@@ -1,3 +1,7 @@
+"""
+This module contains utility functions for the simulation study.
+"""
+
 from configparser import ConfigParser
 import logging
 import warnings
@@ -263,11 +267,26 @@ def save_fe_results(fe_metrics: Dict, params: SimulationParameter, type: Literal
 
 
 def _warning_to_logger(message, category, filename, lineno, file=None, line=None):
-    # Convert warnings to logging messages
+    """
+    Convert warnings to log messages.
+    """
     logging.warning(f"{category.__name__} at {filename} - {lineno}: {message}")
 
 
 def setup_logger(log_dir: Path = Path("logs")):
+    """
+    Setup logging for main process.
+
+    Parameters
+    ----------
+    log_dir : Path, optional
+        Directory for log files, by default Path("logs")
+
+    Returns
+    -------
+    Tuple[multiprocessing.Queue, multiprocessing.QueueListener]
+        Tuple of logging queue and listener.
+    """
     # Create a queue for passing logging messages between processes
     queue = multiprocessing.Queue()
 
@@ -297,7 +316,9 @@ def setup_logger(log_dir: Path = Path("logs")):
 
 
 def configure_worker_logger(queue):
-    # Configure logging for worker processes
+    """
+    Configure logging for worker processes.
+    """
     logger = logging.getLogger()
     logger.handlers = []  # Remove any existing handlers
     logger.addHandler(QueueHandler(queue))

### 2025-02-20 scripts/ablation_study_mc.py
Add descriptions to simulation scripts

@@ -1,3 +1,7 @@
+"""
+Script to perform the ablation study on dataset size impact on Monte Carlo variance.
+"""
+
 import argparse
 import configparser
 from typing import Dict

### 2025-02-20 scripts/ablation_study_variance_decomposition.py

@@ -1,3 +1,7 @@
+"""
+Script to perform the ablation study on estimating Monte Carlo variances to enable variance decomposition.
+"""
+
 import argparse
 from configparser import ConfigParser
 from pathlib import Path

### 2025-02-20 scripts/main.py

@@ -1,3 +1,7 @@
+"""
+Script to run the main simulation study on feature effect error decomposition.
+"""
+
 import argparse
 from configparser import ConfigParser
 from pathlib import Path
@@ -187,21 +191,25 @@ def simulate(
 
 
 if __name__ == "__main__":
+    # parse arguments and read config
     parser = argparse.ArgumentParser()
     parser.add_argument("--config", "-c", type=str, required=True, help="Path to config.ini file")
     args = parser.parse_args()
     sim_config = ConfigParser()
     sim_config.read(Path(args.config))
 
+    # setup logging
     log_queue, log_listener = setup_logger(Path(sim_config.get("storage", "log_dir")))
 
+    # create parameter space
     param_space = create_parameter_space(sim_config)
     logging.info(f"Created parameter space with {len(param_space)} simulation parameters.")
 
+    # create directories and processes
     create_and_set_sim_dir(sim_config, config_path=Path(args.config))
-
     num_processes = min(len(param_space), cpu_count())
 
+    # run simulations
     with Pool(processes=num_processes, initializer=configure_worker_logger, initargs=(log_queue,)) as pool:
         pool.map(
             simulate,

### 2025-02-21 paper/paper.tex

@@
 %\author{Author information scrubbed for double-blind reviewing}
-\author{Timo Heiß\inst{1}}
+\author{Timo Heiß} % \inst{1}
 % You may leave out the orcidID information, if you want to.
@@
 
-    \keywords{Interpretable Machine Learning  \and Feature effects \and Partial dependence plot \and Accumulated local effects}
+    \keywords{Interpretable Machine Learning  \and Feature Effects \and Partial Dependence Plot \and Accumulated Local Effects}
 \end{abstract}
@@
 using validation MSE as objective to minimize (equivalent to maximizing
-R2-score).
+R2-score). Hyperparameter search spaces were chosen following the
+recommendations in~\cite{probst_tunability_2019}.
 
